---
title: "Adding a New C++ Module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a New C++ Module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

FIMS separates each model component into three layers: templated C++ model
logic, a registered native R interface, and an R wrapper. New modules should
follow an existing component with similar parameter and linking requirements.
Selectivity is the smallest useful example; recruitment and fleet demonstrate
more complex registration and model assembly.

# 1. Implement the model logic

Add the templated implementation under
`inst/include/population_dynamics/<category>/functors/` and include it from the
category umbrella header. Keep R-specific types out of this layer. Parameters
used with TMB should use `fims::Vector<Type>` and the implementation should work
with both `double` and automatic-differentiation types.

Add focused GoogleTests under `tests/gtest/` for known values, boundary cases,
and both supported numeric types where appropriate.

# 2. Add a native registry

Objects created from R are owned by a registry in
`inst/include/interface/call/`. Add a registry entry for the new object type,
following the existing category registry. The registry must:

- allocate a stable integer ID;
- retain ownership for as long as the R model needs the object;
- expose lookup and clear operations; and
- reject unknown or stale IDs with a clear error.

Extend `tests/gtest/test_native_NativeRegistries.cpp` to cover creation,
lookup, unique IDs, and clearing.

# 3. Declare and implement native calls

Declare the R entry points in `inst/include/interface/call/<category>.hpp` and
implement them in `src/call_<category>.cpp`. Entry points use `SEXP`, validate
all R inputs, convert them to C++ values, update the registry object, and return
an R value. Wrap exported functions in `extern "C"` and use the R API's
`PROTECT`/`UNPROTECT` discipline for allocated objects.

Register every entry point in the `CallEntries` table in
`inst/include/interface/TMB/init_tmb.hpp`, including its exact argument count.
FIMS disables dynamic symbol lookup, so an unregistered routine is unavailable
to R by design.

# 4. Add the R wrapper

Add a small wrapper in `R/<category>_call.R` that calls the registered routine:

```{r, eval = FALSE}
.Call("fims_call_create_example", value, PACKAGE = "FIMS")
```

Keep validation that improves the user-facing error in R and repeat essential
type, length, and range validation in C++ because native routines may be called
directly. Update `initialize_modules.R` or model assembly when the object needs
to be linked to another component.

# 5. Test the complete path

Add testthat coverage for the R wrapper, invalid inputs, object linking, and
repeated model construction. Then run the standard contributor checks:

```{r, eval = FALSE}
devtools::document()
devtools::test()
devtools::check()
```

Build and run the C++ suite with CMake (or `setup_and_run_gtest()`), and test a
clean package installation. A complete module change normally touches the core
header, native registry, native declaration and source, routine-registration
table, R wrapper, GoogleTests, and testthat tests.
