---
title: "Multi-population modeling in the Fisheries Integrated Modeling System (FIMS)"
output: github_document
vignette: >
  %\VignetteIndexEntry{Multi-population modeling in the Fisheries Integrated Modeling System (FIMS)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates how to configure a FIMS catch-at-age model with two
population objects and then verify that the two populations are represented by
separate model components. The example is written so that the two populations can
be interpreted as two sexes, such as females and males. Most modules are shared
between the two populations to keep the example compact, but the initial numbers
at age are made population-specific so the fitted parameters can be inspected
separately.

This distinction is important: a vignette that only creates two populations shows
that the syntax works, but it does not show how a user would confirm that the two
populations can carry different assumptions or estimates. The checks near the end
of this vignette show how to inspect the fixed-effect parameter table and compare
population-specific quantities.

The workflow is:

1. load operating-model data,
2. create observed-data objects for landings, survey indices, and age compositions,
3. configure fleet and survey modules,
4. define biological modules shared by both populations,
5. create two population modules that can be interpreted as two sexes,
6. make at least one parameter different between the populations,
7. add both populations to a `CatchAtAge` model,
8. build and fit the TMB objective function, and
9. inspect the estimated parameter table to confirm the populations are distinct.

> **Important:** this example uses two populations as a lightweight stand-in for a
> sex-specific model. The two populations share recruitment, growth, maturity,
> fishery, and survey modules, but they have separate `Population` objects and
> separate initial numbers-at-age parameters. A fully sex-specific model may also
> require sex-specific growth, maturity, natural mortality, selectivity, or data
> likelihoods.

## Load FIMS and example data

```{r load-data}
library(FIMS)
library(dplyr)
library(purrr)
library(tibble)

# Set the iteration ID to 1 for accessing specific input/output lists.
iter_id <- 1

# Toggle estimation. If FALSE, the model reports values at the input parameters.
estimation_mode <- TRUE

# Optional random-effects configuration. Set to NULL for the default example.
random_effects <- NULL
map <- list()

c_case <- 1
github_dir <- paste0(
  "https://github.com/NOAA-FIMS/",
  "Age_Structured_Stock_Assessment_Model_Comparison/raw/main/",
  "FIMS_integration_test_data/FIMS_C",
  c_case,
  "/output/OM/"
)

r_data_file <- "OM1.RData"
load(url(paste0(github_dir, r_data_file)))

# Clear any previously registered FIMS modules before building the model.
clear()
```

## Create observed data objects

The fleet module uses observed landings and age-composition data. The survey
module uses an abundance index and age-composition data.

```{r observed-data}
# Observed fishery landings.
landings <- em_input[["L.obs"]][["fleet1"]]
fishing_fleet_landings <- methods::new(Landings, om_input[["nyr"]])

purrr::walk(
  seq_len(om_input[["nyr"]]),
  \(i) fishing_fleet_landings$landings_data$set(i - 1, landings[i])
)

# Observed fishery age compositions, scaled by sample size for likelihood weighting.
fishing_fleet_age_comp <- methods::new(
  AgeComp,
  om_input[["nyr"]],
  om_input[["nages"]]
)

fishery_age_comp_counts <-
  c(t(em_input[["L.age.obs"]][["fleet1"]])) * em_input[["n.L"]][["fleet1"]]

purrr::walk(
  seq_along(fishery_age_comp_counts),
  \(i) fishing_fleet_age_comp$age_comp_data$set(i - 1, fishery_age_comp_counts[i])
)

# Observed survey index.
survey_index <- em_input[["survey.obs"]][["survey1"]]
survey_fleet_index <- methods::new(Index, om_input[["nyr"]])

purrr::walk(
  seq_len(om_input[["nyr"]]),
  \(i) survey_fleet_index$index_data$set(i - 1, survey_index[i])
)

# Observed survey age compositions, scaled by sample size.
survey_fleet_age_comp <- methods::new(
  AgeComp,
  om_input[["nyr"]],
  om_input[["nages"]]
)

survey_age_comp_counts <-
  c(t(em_input[["survey.age.obs"]][["survey1"]])) * em_input[["n.survey"]][["survey1"]]

purrr::walk(
  seq_along(survey_age_comp_counts),
  \(i) survey_fleet_age_comp$age_comp_data$set(i - 1, survey_age_comp_counts[i])
)
```

## Configure the fishery fleet

```{r fishery-fleet}
fishing_fleet_selectivity <- methods::new(LogisticSelectivity)
fishing_fleet_selectivity$inflection_point[1]$value <-
  om_input[["sel_fleet"]][["fleet1"]][["A50.sel1"]]
fishing_fleet_selectivity$inflection_point[1]$estimation_type$set("fixed_effects")

fishing_fleet_selectivity$slope[1]$value <-
  om_input[["sel_fleet"]][["fleet1"]][["slope.sel1"]]
fishing_fleet_selectivity$slope[1]$estimation_type$set("fixed_effects")

fishing_fleet <- methods::new(Fleet)
fishing_fleet$n_years$set(om_input[["nyr"]])
fishing_fleet$n_ages$set(om_input[["nages"]])

fishing_fleet$log_Fmort$resize(om_input[["nyr"]])
for (y in seq_len(om_input[["nyr"]])) {
  fishing_fleet$log_Fmort[y]$value <- log(om_output[["f"]][y])
}
fishing_fleet$log_Fmort$set_all_estimable(TRUE)

# Catchability is fixed at 1 for the fishery landings example.
fishing_fleet$log_q[1]$value <- log(1.0)
fishing_fleet$log_q[1]$estimation_type$set("constant")

fishing_fleet$SetSelectivityID(fishing_fleet_selectivity$get_id())
fishing_fleet$SetObservedLandingsDataID(fishing_fleet_landings$get_id())
fishing_fleet$SetObservedAgeCompDataID(fishing_fleet_age_comp$get_id())

# Landings likelihood: lognormal observation model.
fishing_fleet_landings_distribution <- methods::new(DlnormDistribution)
fishing_fleet_landings_distribution$log_sd$resize(om_input[["nyr"]])

for (y in seq_len(om_input[["nyr"]])) {
  # Convert coefficient of variation to lognormal standard deviation,
  # then store it on the log scale required by DlnormDistribution.
  cv <- em_input[["cv.L"]][["fleet1"]]
  fishing_fleet_landings_distribution$log_sd[y]$value <- log(sqrt(log(cv^2 + 1)))
}
fishing_fleet_landings_distribution$log_sd$set_all_estimable(FALSE)
fishing_fleet_landings_distribution$set_observed_data(
  fishing_fleet$GetObservedLandingsDataID()
)
fishing_fleet_landings_distribution$set_distribution_links(
  "data",
  fishing_fleet$log_landings_expected$get_id()
)

# Age-composition likelihood: multinomial observation model.
fishing_fleet_agecomp_distribution <- methods::new(DmultinomDistribution)
fishing_fleet_agecomp_distribution$set_observed_data(
  fishing_fleet$GetObservedAgeCompDataID()
)
fishing_fleet_agecomp_distribution$set_distribution_links(
  "data",
  fishing_fleet$agecomp_proportion$get_id()
)
```

## Configure the survey fleet

The survey is represented with a fleet object that has negligible fishing
mortality and an estimable survey catchability parameter.

```{r survey-fleet}
survey_fleet_selectivity <- methods::new(LogisticSelectivity)
survey_fleet_selectivity$inflection_point[1]$value <-
  om_input[["sel_survey"]][["survey1"]][["A50.sel1"]]
survey_fleet_selectivity$inflection_point[1]$estimation_type$set("fixed_effects")

survey_fleet_selectivity$slope[1]$value <-
  om_input[["sel_survey"]][["survey1"]][["slope.sel1"]]
survey_fleet_selectivity$slope[1]$estimation_type$set("fixed_effects")

survey_fleet <- methods::new(Fleet)
survey_fleet$n_years$set(om_input[["nyr"]])
survey_fleet$n_ages$set(om_input[["nages"]])

survey_fleet$log_Fmort$resize(om_input[["nyr"]])
for (y in seq_len(om_input[["nyr"]])) {
  survey_fleet$log_Fmort[y]$value <- -200
}
survey_fleet$log_Fmort$set_all_estimable(FALSE)

survey_fleet$log_q[1]$value <- log(om_output[["survey_q"]][["survey1"]])
survey_fleet$log_q[1]$estimation_type$set("fixed_effects")

survey_fleet$SetSelectivityID(survey_fleet_selectivity$get_id())
survey_fleet$SetObservedIndexDataID(survey_fleet_index$get_id())
survey_fleet$SetObservedAgeCompDataID(survey_fleet_age_comp$get_id())

# Survey index likelihood: lognormal observation model.
survey_fleet_index_distribution <- methods::new(DlnormDistribution)
survey_fleet_index_distribution$log_sd$resize(om_input[["nyr"]])

for (y in seq_len(om_input[["nyr"]])) {
  cv <- em_input[["cv.survey"]][["survey1"]]
  survey_fleet_index_distribution$log_sd[y]$value <- log(sqrt(log(cv^2 + 1)))
}
survey_fleet_index_distribution$log_sd$set_all_estimable(FALSE)
survey_fleet_index_distribution$set_observed_data(
  survey_fleet$GetObservedIndexDataID()
)
survey_fleet_index_distribution$set_distribution_links(
  "data",
  survey_fleet$log_index_expected$get_id()
)

# Survey age-composition likelihood.
survey_fleet_agecomp_distribution <- methods::new(DmultinomDistribution)
survey_fleet_agecomp_distribution$set_observed_data(
  survey_fleet$GetObservedAgeCompDataID()
)
survey_fleet_agecomp_distribution$set_distribution_links(
  "data",
  survey_fleet$agecomp_proportion$get_id()
)

# Age-to-length conversion matrix used by the survey fleet.
for (i in seq_along(em_input[["age_to_length_conversion"]])) {
  survey_fleet$age_to_length_conversion[i]$value <-
    c(t(em_input[["age_to_length_conversion"]]))[i]
}
survey_fleet$age_to_length_conversion$set_all_estimable(FALSE)
survey_fleet$age_to_length_conversion$set_all_random(FALSE)
```

## Optional selectivity random effects

```{r selectivity-random-effects}
if (!is.null(random_effects) && "selectivity" %in% names(random_effects)) {
  if (random_effects[["selectivity"]] == "log_devs") {
    fishing_fleet_selectivity$log_devs$set_all_random(TRUE)
    survey_fleet_selectivity$log_devs$set_all_random(TRUE)
  }

  if (random_effects[["selectivity"]] == "log_sel") {
    fishing_fleet_selectivity$log_sel$set_all_random(TRUE)
    survey_fleet_selectivity$log_sel$set_all_random(TRUE)
  }

  if (random_effects[["selectivity"]] == "pars") {
    fishing_fleet_selectivity$inflection_point$estimation_type$set("random_effects")
    fishing_fleet_selectivity$slope$estimation_type$set("random_effects")
    survey_fleet_selectivity$inflection_point$estimation_type$set("random_effects")
    survey_fleet_selectivity$slope$estimation_type$set("random_effects")
  }
}
```

## Define shared recruitment

Both populations use the same Beverton-Holt recruitment module in this example.
Sharing the module means both populations reference the same recruitment
parameters and process.

```{r recruitment}
recruitment <- methods::new(BevertonHoltRecruitment)

if (is.null(random_effects) || random_effects[["recruitment"]] == "log_devs") {
  recruitment_process <- methods::new(LogDevsRecruitmentProcess)
} else {
  recruitment_process <- methods::new(LogRRecruitmentProcess)
}

recruitment$SetRecruitmentProcessID(recruitment_process$get_id())
recruitment$n_years$set(om_input[["nyr"]])

# Equilibrium recruitment.
recruitment$log_rzero[1]$value <- log(om_input[["R0"]])
recruitment$log_rzero[1]$estimation_type$set("fixed_effects")

# Beverton-Holt steepness on the FIMS logit scale.
recruitment$logit_steep[1]$value <-
  -log(1.0 - om_input[["h"]]) + log(om_input[["h"]] - 0.2)
recruitment$logit_steep[1]$estimation_type$set("constant")

if (is.null(random_effects) || random_effects[["recruitment"]] == "log_devs") {
  recruitment$log_devs$resize(om_input[["nyr"]] - 1)
  for (y in seq_len(om_input[["nyr"]] - 1)) {
    recruitment$log_devs[y]$value <- om_input[["logR.resid"]][y + 1]
  }
}

if (!is.null(random_effects) && "recruitment" %in% names(random_effects)) {
  if (random_effects[["recruitment"]] == "log_devs") {
    recruitment$log_devs$set_all_random(TRUE)
  }

  if (random_effects[["recruitment"]] == "log_r") {
    recruitment$log_r$resize(om_input[["nyr"]] - 1)
    recruits_true <- matrix(
      c(t(om_output[["N.age"]])),
      om_input[["nyr"]],
      om_input[["nages"]],
      byrow = TRUE
    )[, 1]

    for (y in seq_len(om_input[["nyr"]] - 1)) {
      recruitment$log_r[y]$value <- if (estimation_mode) 0 else recruits_true[y + 1]
    }
    recruitment$log_r$set_all_random(TRUE)
  }
}

if (is.null(random_effects)) {
  recruitment$log_devs$set_all_estimable(TRUE)
}

recruitment_distribution <- methods::new(DnormDistribution)
recruitment_distribution$log_sd$resize(1)
recruitment_distribution$log_sd[1]$value <- log(om_input[["logR_sd"]])

recruitment_distribution$observed_values$resize(om_input[["nyr"]] - 1)
recruitment_distribution$expected_values$resize(om_input[["nyr"]] - 1)

for (i in seq_len(om_input[["nyr"]] - 1)) {
  recruitment_distribution$observed_values[i]$value <- 0
  recruitment_distribution$expected_values[i]$value <- 0
}

if (!is.null(random_effects) && "recruitment" %in% names(random_effects)) {
  if (random_effects[["recruitment"]] == "log_devs") {
    recruitment_distribution$log_sd[1]$estimation_type$set("fixed_effects")
    recruitment_distribution$set_distribution_links(
      "random_effects",
      recruitment$log_devs$get_id()
    )
  }

  if (random_effects[["recruitment"]] == "log_r") {
    recruitment_distribution$log_sd[1]$value <- log(1)
    recruitment_distribution$log_sd[1]$estimation_type$set("fixed_effects")
    recruitment_distribution$set_distribution_links(
      "random_effects",
      c(
        recruitment$log_r$get_id(),
        recruitment$log_expected_recruitment$get_id()
      )
    )
  }
} else {
  recruitment_distribution$set_distribution_links(
    "random_effects",
    recruitment$log_devs$get_id()
  )
}
```

## Define shared growth and maturity

The growth and maturity modules are shared by both populations in this example.
That means the two populations use the same weights-at-age and maturity schedule.
For a more detailed sex-specific application, this is one place where users may
choose to create separate female and male growth or maturity modules.

```{r growth-maturity}
# Empirical weight-at-age growth object. This object is shared by both
# populations in the example.
ewaa_growth <- methods::new(EWAAGrowth)
ewaa_growth$n_years$set(om_input[["nyr"]])
ewaa_growth$ages$resize(om_input[["nages"]])

purrr::walk(
  seq_along(om_input[["ages"]]),
  \(i) ewaa_growth$ages$set(i - 1, om_input[["ages"]][i])
)

# FIMS expects the annual weights-at-age as a flat vector. Here, the same
# weight-at-age vector is repeated across years.
ewaa_growth$weights$resize((om_input[["nyr"]] + 1) * om_input[["nages"]])

purrr::walk(
  seq_len(ewaa_growth$weights$size()),
  \(i) ewaa_growth$weights$set(
    i - 1,
    om_input[["W.mt"]][((i - 1) %% om_input[["nages"]]) + 1]
  )
)

# Logistic maturity is also shared. In a sex-specific model, females and males
# could receive separate maturity modules if their maturity schedules differ.
maturity <- methods::new(LogisticMaturity)
maturity$inflection_point[1]$value <- om_input[["A50.mat"]]
maturity$inflection_point[1]$estimation_type$set("constant")
maturity$slope[1]$value <- om_input[["slope.mat"]]
maturity$slope[1]$estimation_type$set("constant")
```

## Define two population modules

The helper below keeps the population setup consistent and avoids duplicating the
same block of code for each population. Each population receives its own
`Population` object but references the same biological and fleet modules.

To make the example more useful than a pure setup demonstration, the two
populations are labeled as `female` and `male`, and the male initial numbers at
age are initialized with a multiplier. This gives the vignette a concrete
sex-specific interpretation and provides a parameter block that can be inspected
after fitting.

```{r populations}
make_population <- function(init_naa_multiplier = 1.0) {
  population <- methods::new(Population)

  # Natural mortality is held fixed in this vignette. The same value is used for
  # all years and ages in both populations. For a sex-specific model, this could
  # be changed to use different female and male M-at-age schedules.
  population$log_M$resize(om_input[["nyr"]] * om_input[["nages"]])
  for (i in seq_len(om_input[["nyr"]] * om_input[["nages"]])) {
    population$log_M[i]$value <- log(om_input[["M.age"]][1])
  }
  population$log_M$set_all_estimable(FALSE)

  # Initial numbers at age are population-specific parameters. The multiplier is
  # used only to make the two populations start from different values so that the
  # user can verify that the two parameter blocks are distinct.
  population$log_init_naa$resize(om_input[["nages"]])
  for (i in seq_len(om_input[["nages"]])) {
    population$log_init_naa[i]$value <- log(
      om_output[["N.age"]][1, i] * init_naa_multiplier
    )
  }
  population$log_init_naa$set_all_estimable(TRUE)

  # Basic population dimensions and age vector.
  population$n_ages$set(om_input[["nages"]])
  population$ages$resize(om_input[["nages"]])
  purrr::walk(
    seq_along(om_input[["ages"]]),
    \(i) population$ages$set(i - 1, om_input[["ages"]][i])
  )

  # Each population is connected to the fishery and survey fleets.
  population$n_fleets$set(sum(om_input[["fleet_num"]], om_input[["survey_num"]]))
  population$n_years$set(om_input[["nyr"]])

  # Attach the shared biological modules.
  population$SetRecruitmentID(recruitment$get_id())
  population$SetGrowthID(ewaa_growth$get_id())
  population$SetMaturityID(maturity$get_id())

  # Attach the shared fishery and survey modules.
  population$AddFleet(fishing_fleet$get_id())
  population$AddFleet(survey_fleet$get_id())

  population
}

# In this example, population 1 is interpreted as female and population 2 as
# male. The male initial abundance is initialized lower so that the vignette can
# demonstrate how to identify a population-specific parameter difference.
female_population <- make_population(init_naa_multiplier = 1.0)
male_population <- make_population(init_naa_multiplier = 0.75)

# Keep a small lookup table that maps FIMS population IDs to the user-facing
# labels used in the result summaries.
population_lookup <- tibble::tibble(
  population_id = c(female_population$get_id(), male_population$get_id()),
  sex = c("female", "male"),
  init_naa_multiplier = c(1.0, 0.75)
)

population_lookup
```

## Create and fit the catch-at-age model

```{r fit-model}
caa <- methods::new(CatchAtAge)

# Register both populations with the catch-at-age model. Because these are two
# distinct Population objects, their population-level parameters can differ even
# when they share other modules.
caa$AddPopulation(female_population$get_id())
caa$AddPopulation(male_population$get_id())

CreateTMBModel()

# The fixed-effect and random-effect vectors are the objects passed to TMB. The
# population-specific initial numbers at age should appear in the fixed-effect
# vector because they were marked estimable above.
parameters <- list(
  p = get_fixed(),
  re = get_random()
)

obj <- TMB::MakeADFun(
  data = list(),
  parameters = parameters,
  DLL = "FIMS",
  silent = TRUE,
  map = map,
  random = "re"
)

if (estimation_mode) {
  # Fit the model by minimizing the joint negative log-likelihood.
  opt <- stats::nlminb(
    obj[["par"]],
    obj[["fn"]],
    obj[["gr"]],
    control = list(eval.max = 10000, iter.max = 10000, trace = 0)
  )

  # Update the FIMS fixed-effect vector with the optimizer solution before
  # extracting model output.
  FIMS::set_fixed(opt$par)
  fims_finalized <- caa$get_output()

  # sdreport gives parameter estimates and standard errors for fixed effects,
  # random effects, and reported quantities.
  sdr <- TMB::sdreport(obj)
  sdr_report <- summary(sdr, "report")
  sdr_fixed <- summary(sdr, "fixed")
  sdr_random <- summary(sdr, "random")
} else {
  opt <- NULL
  sdr <- list()
  sdr_report <- list()
  sdr_fixed <- matrix(
    obj[["env"]]$parList()[["p"]],
    ncol = 1,
    dimnames = list(NULL, "Estimate")
  )
  sdr_random <- matrix(
    obj[["env"]]$parList()[["re"]],
    ncol = 1,
    dimnames = list(NULL, "Estimate")
  )
}

row.names(sdr_fixed) <- names(FIMS:::get_parameter_names(sdr_fixed[, 1]))
row.names(sdr_random) <- names(FIMS:::get_random_names(sdr_random[, 1]))

report <- obj[["report"]](obj[["env"]][["last.par.best"]])
```

## Inspect results

The first check is numerical: confirm that the optimizer converged and that the
final gradient is small. The second check is biological/model-structural: confirm
that the parameter table contains population-specific parameters and that the
female and male initial numbers at age can be inspected separately.

```{r inspect-results}
# Gradient at the final parameter values. Values near zero indicate that the
# optimizer has likely reached a local optimum.
obj$gr()

# Optimizer diagnostics and convergence code. For nlminb(), convergence == 0
# usually indicates successful convergence.
opt

# Reported quantities from the model.
head(sdr_report)
```

### Inspect population-specific parameters

The exact parameter names can change as FIMS evolves, so the safest first step is
to convert the `sdreport` output into a data frame and search for the parameter
block of interest. Here, we search for `log_init_naa` because those parameters
were intentionally made population-specific above.

```{r inspect-population-parameters}
fixed_effects <- tibble::as_tibble(
  sdr_fixed,
  rownames = "parameter"
)

# Show the available fixed-effect parameter names. This is useful when adapting
# the vignette to a new FIMS version or to a different population-specific
# parameter, such as sex-specific natural mortality or growth.
fixed_effects$parameter

# Pull out the initial numbers-at-age parameters. These should contain separate
# entries for the two Population objects because each population has its own
# log_init_naa vector.
init_naa_parameters <- fixed_effects |>
  dplyr::filter(grepl("log_init_naa", parameter))

init_naa_parameters
```

### Compare the two population blocks

For a sex-specific interpretation, it is helpful to summarize the estimated
initial numbers at age by population. The code below assumes the `log_init_naa`
parameters are ordered by population and then by age, which is the order used when
the two populations are created and registered above. If the printed parameter
names include explicit population IDs, users can join on those IDs instead.

```{r compare-sex-specific-parameters}
init_naa_by_sex <- init_naa_parameters |>
  dplyr::mutate(
    sex = rep(population_lookup$sex, each = om_input[["nages"]]),
    age = rep(om_input[["ages"]], times = nrow(population_lookup)),
    initial_numbers_at_age = exp(Estimate)
  ) |>
  dplyr::select(sex, age, parameter, Estimate, initial_numbers_at_age)

init_naa_by_sex

# A compact comparison by sex. Because the male population was initialized with a
# 0.75 multiplier, this table should make it easy to confirm that the two
# population-specific parameter blocks are not identical.
init_naa_by_sex |>
  dplyr::group_by(sex) |>
  dplyr::summarise(
    total_initial_abundance = sum(initial_numbers_at_age),
    .groups = "drop"
  )
```

The same pattern can be used for a more complete sex-specific model. For example,
users could make female and male growth modules, assign each population a
different growth module ID, and then inspect the corresponding growth parameters
in `sdr_fixed`.

## Clean up

```{r cleanup}
clear()
```

## Notes for extending this example

This example demonstrates the mechanics of adding more than one population to a
single catch-at-age model and checking that at least one population-specific
parameter differs between the two population blocks. Extensions may include:

- sex-specific or population-specific recruitment modules,
- sex-specific growth or maturity parameters,
- sex-specific natural mortality schedules,
- separate fleets or surveys for each population or sex,
- movement between populations or areas, and
- population-specific likelihood components.

When adapting this example, check whether a module should be shared across
populations or instantiated separately. Sharing a module means the populations use
the same registered object and therefore the same parameterization. Creating
separate modules makes the sex-specific or population-specific assumptions more
explicit and makes the corresponding parameters easier to inspect in the fitted
parameter tables.
