---
title: "How FIMS Passes Parameters to the Backend and Links Random Effects to Distributions"
output: github_document
vignette: >
  %\VignetteIndexEntry{How FIMS Passes Parameters to the Backend and Links Random Effects to Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set local options so vignette output is clean and consistent
# (avoids color codes and subtle formatting in rendered docs)
withr::local_options(list(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.subtle = FALSE
))
```


## Overview

This vignette traces the full lifecycle of a random effects parameter in FIMS,
using annual log-scale recruitment (`log_r`) from the Beverton--Holt recruitment
module as a concrete example. The goal is to show how a value set in R ultimately
participates in TMB's automatic differentiation and is linked to its likelihood
distribution, from the R interface through the Rcpp layer into the C++/TMB backend.

The journey has five stages:

1. **R setup** -- the user configures parameter objects and their estimation type.
2. **Rcpp interface** -- `BevertonHoltRecruitmentInterface` and `DnormDistributionsInterface`
   store initial values and register distribution links.
3. **`CreateTMBModel()`** -- each interface object calls `add_to_fims_tmb()`, which
   pushes parameters into the `Information` singleton and populates the `variable_map`.
4. **`SetupRandomEffects()`** -- pointers are resolved so each distribution knows
   which parameter vector its random effect lives in.
5. **TMB objective function** -- `re` values are written into memory, and the model
   evaluates the joint negative log-likelihood.

---

## Two formulations of recruitment random effects

Before tracing the code path, it is worth being precise about the statistical
distinction between the two recruitment random effect formulations FIMS supports,
because it determines which parameters are random effects, what the likelihood
penalizes, and which process module the population model calls.

### Log-deviations (`LogDevs`)

In the deviations formulation the Beverton--Holt mean is computed first and
recruitment deviations are added on the log scale:

$$
\log R_y = \log \hat{R}_y + \delta_y, \qquad \delta_y \sim N(0,\, \sigma_r)
$$

where $\hat{R}_y = \text{BH}(S_{y-1})$ is the stock-recruit mean and
$\delta_y$ is `log_devs[y]`. The random effects are the deviations themselves,
and the distribution links `log_devs` (observed) to zero (expected). The
process module `LogDevs::evaluate_process` returns
`log_expected_recruitment[y] + log_devs[y]`.

### Log-recruitment (`LogR`)

In the log-recruitment formulation, the annual log-recruitment values are the
random effects, and the Beverton--Holt mean appears as the *expected value* of
the distribution rather than an additive correction:

$$
\log R_y \sim N\!\left(\log \hat{R}_y,\, \sigma_r\right)
$$

Here `log_r[y]` is the random effect and `log_expected_recruitment[y]` (the
BH mean on the log scale) is the distribution's expected value. The process
module `LogR::evaluate_process` returns `log_r[y]` directly, so the population
model uses the full random-effect draw as the realized log-recruitment.

The table below summarizes the mapping:

| | `LogDevs` | `LogR` |
|---|---|---|
| Random effect | `log_devs` | `log_r` |
| `set_distribution_links` observed | `log_devs` id | `log_r` id |
| `set_distribution_links` expected | (zero vector, scalar) | `log_expected_recruitment` id |
| `evaluate_process` returns | `log_expected_recruitment[y] + log_devs[y]` | `log_r[y]` |
| BH mean enters as | additive term in process | expected value of likelihood |

Both approaches produce the same numbers at age: `CalculateRecruitment` writes
`log_expected_recruitment[y]` from the BH mean and then calls
`process->evaluate_process(y)` to get the realized log-recruitment that
determines `numbers_at_age`. The difference is purely in which quantity is
the random effect and how it is penalized.

---

## Stage 1: R Setup

The user begins by creating a `BevertonHoltRecruitment` object and a companion
process object. The choice of process module (`LogRRecruitmentProcess` vs.
`LogDevsRecruitmentProcess`) determines the formulation above. The example
below uses `LogRRecruitmentProcess`; the `LogDevs` case is noted where it
differs. The process object is linked to the recruitment object by ID so
`CalculateRecruitment` can dispatch `evaluate_process` through the correct
module at runtime.

```{r r-setup, eval = FALSE}
recruitment <- methods::new(BevertonHoltRecruitment)

# Choose the process module: LogRRecruitmentProcess (log_r formulation)
# or LogDevsRecruitmentProcess (log_devs formulation)
recruitment_process <- methods::new(LogRRecruitmentProcess)

# Link the process to the recruitment object by ID.
# CalculateRecruitment() calls process->evaluate_process() through this link.
recruitment$SetRecruitmentProcessID(recruitment_process$get_id())

# Fixed scalar parameters shared by both formulations
recruitment$log_rzero[1]$value <- log(om_input[["R0"]])
recruitment$log_rzero[1]$estimation_type$set("fixed_effects")

recruitment$logit_steep[1]$value <- -log(1.0 - om_input[["h"]]) +
                                      log(om_input[["h"]] - 0.2)
recruitment$logit_steep[1]$estimation_type$set("constant")

recruitment$n_years$set(om_input[["nyr"]])
```

#### `LogR` formulation

`log_r` is resized to `n_years - 1` and every element is declared a random
effect. Initial values of zero place the draws on the natural log scale with
no starting offset.

```{r log-r-setup, eval = FALSE}
recruitment$log_r$resize(om_input[["nyr"]] - 1)
for (y in 1:(om_input[["nyr"]] - 1)) {
  recruitment$log_r[y]$value <- 0
}
recruitment$log_r$set_all_random(TRUE)
```

The `DnormDistribution` is then configured with two keys: `log_r` as the
observed (random effect) vector and `log_expected_recruitment` as the expected
value vector. When `SetupRandomEffects()` runs it resolves these IDs to raw
pointers, so during model evaluation the distribution sees the BH mean in
`re_expected_values` and the current random effect draw in `re`.

```{r dnorm-setup-logr, eval = FALSE}
recruitment_distribution <- methods::new(DnormDistribution)

recruitment_distribution$log_sd$resize(1)
recruitment_distribution$log_sd[1]$value <- log(1)
recruitment_distribution$log_sd[1]$estimation_type$set("fixed_effects")

# Both observed and expected slots are sized to n_years - 1.
# Their actual values at runtime come through the re / re_expected_values
# pointers set by SetupRandomEffects(), not from these initial values.
recruitment_distribution$observed_values$resize(om_input[["nyr"]] - 1)
recruitment_distribution$expected_values$resize(om_input[["nyr"]] - 1)
for (i in 1:(om_input[["nyr"]] - 1)) {
  recruitment_distribution$observed_values[i]$value <- 0
  recruitment_distribution$expected_values[i]$value <- 0
}

# Two-key link: key[0] = log_r (the RE), key[1] = log_expected_recruitment
# (the BH mean, used as the distribution's expected value)
recruitment_distribution$set_distribution_links(
  "random_effects",
  c(recruitment$log_r$get_id(),
    recruitment$log_expected_recruitment$get_id())
)
```

#### `LogDevs` formulation (contrast)

In the deviations case the setup is analogous, but only one key is passed to
`set_distribution_links`: the `log_devs` vector. The expected value of the
penalty is zero (deviations are mean-zero), so there is no second key and
`SetupRandomEffects()` sets `re_expected_values = &d->expected_values` (the
zero vector allocated inside the distribution itself).

```{r dnorm-setup-logdevs, eval = FALSE}
# LogDevs: only one key -- log_devs is penalized toward zero
recruitment_distribution$set_distribution_links(
  "random_effects",
  c(recruitment$log_devs$get_id())   # no second key
)
```

---

## Stage 2: The Rcpp Interface Layer

Each R-level object corresponds to a C++ class that inherits from
`FIMSRcppInterfaceBase`. These classes hold `ParameterVector` members, which
wrap individual parameters along with metadata like `initial_value_m`,
`final_value_m`, and `estimation_type_m`.

When `set_distribution_links()` is called from R, `DnormDistributionsInterface`
stores the `input_type` string and the vector of integer keys in `key_m`:

```cpp
virtual bool set_distribution_links(std::string input_type,
                                    Rcpp::IntegerVector ids) {
  this->input_type_m.set(input_type);
  this->key_m->resize(ids.size());
  for (R_xlen_t i = 0; i < ids.size(); i++) {
    this->key_m->at(i) = ids[i];
  }
  return true;
}
```

No pointers are resolved yet. The interface simply records the user's intent.
All actual linkage happens after `CreateTMBModel()` is called.

---

## Stage 3: `CreateTMBModel()` and `add_to_fims_tmb()`

`CreateTMBModel()` iterates over every registered interface object and calls
`add_to_fims_tmb()` on each one. For `BevertonHoltRecruitmentInterface`, this
ultimately calls `add_to_fims_tmb_internal<Type>()` for each TMB type (double,
and the AD types needed for first, second, and third derivatives).

Inside `add_to_fims_tmb_internal`, the interface does three things for each
parameter vector:

**1. Copy initial values into a TMB-typed object.**

```cpp
recruitment->log_r.resize(this->log_r.size());
for (size_t i = 0; i < log_r.size(); i++) {
  recruitment->log_r[i] = this->log_r[i].initial_value_m;
  ...
}
```

**2. Register estimable parameters.**

If `estimation_type_m` is `"fixed_effects"`, the parameter's address is passed
to `info->RegisterParameter()`, which appends a pointer to
`information->fixed_effects_parameters`. If it is `"random_effects"`, it goes
to `information->random_effects_parameters` instead. Constants are neither
registered nor differentiated.

```cpp
if (this->log_r[i].estimation_type_m.get() == "random_effects") {
  ss.str("");
  ss << "Recruitment." << this->id << ".log_r." << this->log_r[i].id_m;
  info->RegisterRandomEffectName(ss.str());
  info->RegisterRandomEffect(recruitment->log_r[i]);
}
```

**3. Register the vector in `variable_map`.**

After all elements are processed, the entire parameter vector is entered into
`Information::variable_map` under the `ParameterVector`'s unique ID. This map
is the mechanism by which distributions later find the memory location of their
linked parameters.

```cpp
info->variable_map[this->log_r.id_m] = &(recruitment)->log_r;
```

The same pattern applies to `log_expected_recruitment`:

```cpp
info->variable_map[this->log_expected_recruitment.id_m] =
    &(recruitment)->log_expected_recruitment;
```

`DnormDistributionsInterface::add_to_fims_tmb_internal` follows the same
pattern for `log_sd`. It also copies `key_m` into the C++ distribution object's
`key` vector, which is what `SetupRandomEffects()` uses in the next stage.

---

## Stage 4: `SetupRandomEffects()` Resolves Pointers

After all interface objects have pushed their content into `Information`,
`info->CreateModel()` is called. One of the internal steps is
`SetupRandomEffects()`, which iterates over all registered density components.
For any distribution whose `input_type` is `"random_effects"`, it uses the
stored `key` values to look up actual memory addresses in `variable_map` and
assign them to the distribution's `re` and `re_expected_values` pointers.

```cpp
void SetupRandomEffects() {
  for (density_components_iterator it = this->density_components.begin();
       it != this->density_components.end(); ++it) {
    std::shared_ptr<fims_distributions::DensityComponentBase<Type>> d =
        (*it).second;

    if (d->input_type == "random_effects") {
      variable_map_iterator vmit;

      // key[0] -> the random effect vector (log_r)
      vmit = this->variable_map.find(d->key[0]);
      d->re = (*vmit).second;

      // key[1] -> the expected value vector (log_expected_recruitment)
      if (d->key.size() == 2) {
        vmit = this->variable_map.find(d->key[1]);
        d->re_expected_values = (*vmit).second;
      } else {
        d->re_expected_values = &d->expected_values;
      }
    }
  }
}
```

After this step, the `DnormDistribution` object in the backend directly holds
raw pointers into the same memory as the recruitment model's `log_r` and
`log_expected_recruitment` vectors. When the optimizer updates those values,
the distribution sees the changes immediately.

---

## Stage 5: The TMB Objective Function

After `CreateTMBModel()`, the user extracts the parameter vectors and passes
them to `TMB::MakeADFun`:

```{r make-adfun, eval = FALSE}
parameters <- list(
  p  = get_fixed(),
  re = get_random()
)

obj <- TMB::MakeADFun(
  data       = list(),
  parameters = parameters,
  DLL        = "FIMS",
  silent     = TRUE,
  map        = map,
  random     = "re"
)
```

TMB's Laplace approximation marginalizes over `re`. Internally, each call to the
objective function executes `FIMS.cpp`:

```cpp
template<class Type>
Type objective_function<Type>::operator()() {
  PARAMETER_VECTOR(p);
  PARAMETER_VECTOR(re);

  std::shared_ptr<fims_info::Information<Type>> information =
    fims_info::Information<Type>::GetInstance();

  // Write fixed effects into model memory
  for (size_t i = 0; i < information->fixed_effects_parameters.size(); i++) {
    *information->fixed_effects_parameters[i] = p[i];
  }

  // Write random effects into model memory
  for (size_t i = 0; i < information->random_effects_parameters.size(); i++) {
    *information->random_effects_parameters[i] = re[i];
  }

  Type nll = 0;
  nll = model->Evaluate();
  return nll;
}
```

The write loop copies each element of TMB's `re` vector directly into the
addresses held by `information->random_effects_parameters`. Because those
addresses are the same memory locations as `recruitment->log_r[i]`, and because
the `DnormDistribution`'s `re` pointer points into that same vector, the
likelihood computation in `model->Evaluate()` immediately sees the updated
values. No data copying occurs; the entire system operates through shared memory
addresses registered at model-build time.

`model->Evaluate()` then loops over all density components. For the
`DnormDistribution` linked to `log_r`, it computes the normal log-density of
each `log_r[i]` value relative to the corresponding `log_expected_recruitment[i]`
value, using `exp(log_sd)` as the standard deviation. The result is accumulated
into the joint negative log-likelihood returned to TMB.

---

## How `CalculateRecruitment` uses the process module

The process module is the link between the statistical random effect and the
population dynamics. Each time step `CalculateRecruitment` is called with the
current year index. It first writes the Beverton--Holt mean into
`log_expected_recruitment`:

```cpp
population->recruitment->log_expected_recruitment[year - 1] =
    fims_math::log(population->recruitment->evaluate_mean(
        dq_["spawning_biomass"][year - 1], phi0));
```

It then calls `process->evaluate_process(year - 1)` to get the realized
log-recruitment that populates `numbers_at_age`:

```cpp
dq_["numbers_at_age"][i_age_year] = fims_math::exp(
    population->recruitment->process->evaluate_process(year - 1));
```

For `LogR`, `evaluate_process` returns `log_r[y]` -- the random effect draw --
so the Laplace approximation marginalizes over the full distribution
$\log R_y \sim N(\log\hat{R}_y,\, \sigma_r)$.

For `LogDevs`, `evaluate_process` returns
`log_expected_recruitment[y] + log_devs[y]`, so the BH mean is a deterministic
correction and the deviation is the quantity that carries uncertainty.

In both cases the population model sees a single log-recruitment value per year;
only the source and interpretation of that value differ between formulations.

---

## Summary

The table below summarizes how each piece of information flows through the system
for both formulations.

| | `LogR` | `LogDevs` |
|---|---|---|
| Random effect | `log_r` | `log_devs` |
| RE registered in `variable_map` | `id_log_r → &log_r` | `id_log_devs → &log_devs` |
| `set_distribution_links` keys | `[id_log_r, id_log_exp_rec]` | `[id_log_devs]` |
| `NormalLPDF::re` points to | `log_r` | `log_devs` |
| `NormalLPDF::re_expected_values` points to | `log_expected_recruitment` | internal zero vector |
| `evaluate_process` returns | `log_r[y]` | `log_expected_recruitment[y] + log_devs[y]` |
| BH mean enters likelihood as | expected value | additive process correction |

The design achieves a clean separation of concerns: the R user specifies *what*
parameters exist and *how* they relate to distributions; the Rcpp interface
records that intent; `CreateTMBModel()` + `SetupRandomEffects()` resolve
everything into raw pointer connections; and `CalculateRecruitment` dispatches
through the process module to get a single realized log-recruitment value per
year that is agnostic to which formulation is in use.
