---
title: "SPR and MSY From a Biological Snapshot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPR and MSY From a Biological Snapshot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(FIMS)
```

`calculate_reference_points()` computes equilibrium reference points from a
fixed biological snapshot. This makes it possible to explore assumptions
without creating or fitting a population model. To obtain a snapshot from a
live model, see [Reference Points From a CatchAtAge Model](fims-reference-points-model.html).

The examples use invented biological values to demonstrate the calculations.
They are not reference-point recommendations for a stock.

## Define the biology and fishing pattern

Each biological vector contains one entry per annual age, youngest first. One
recruit enters the first age each year, and the last age is a plus group.
Natural mortality is an instantaneous annual rate. Weights use a consistent
biomass unit throughout the calculation.

```{r inputs}
ages <- 1:6
inputs <- list(
  natural_mortality = rep(0.2, length(ages)),
  weight = c(0.2, 0.5, 1, 1.5, 2, 2.5),
  maturity = c(0, 0.1, 0.5, 0.9, 1, 1),
  proportion_female = rep(0.5, length(ages)),
  fleets = list(list(
    share = 1,
    selectivity = c(0.1, 0.3, 0.7, 1, 1, 1),
    weight = c(0.2, 0.5, 1, 1.5, 2, 2.5)
  ))
)
```

Fleet shares must sum to one. At age $a$, total fishing mortality is

$$F_a = F \sum_f \mathrm{share}_f\,\mathrm{selectivity}_{f,a}.$$

The reported $F$ is the coefficient in this expression. Selectivity is used
exactly as supplied, so $F$ need not equal the maximum mortality at age. Here,
one fleet has maximum selectivity one, making the interpretation direct.
Catch weights may differ from population weights in a standalone snapshot.

## Find fishing mortality at an SPR target

Spawning potential ratio is spawning biomass per recruit under fishing divided
by its unfished value:

$$\mathrm{SPR}(F) = \mathrm{SBPR}(F) / \mathrm{SBPR}(0).$$

Both biomass quantities are measured at the start of the year. SPR is an
equilibrium per-recruit ratio; it is different from the annual spawning biomass
ratio in a fitted population trajectory.

```{r spr}
spr_result <- calculate_reference_points(
  inputs,
  spr_targets = c(0.5, 0.4, 0.35),
  msy = FALSE
)
knitr::kable(spr_result$spr, digits = 6)
```

A target of `0.4` means 40% of unfished spawning biomass per recruit. Check the
`status` column before using its fishing mortality. The residual is achieved
SPR minus the requested target. SPR calculations do not require a
stock-recruitment relationship.

```{r check-spr, include=FALSE}
stopifnot(
  all(spr_result$spr$status == "converged"),
  all(abs(spr_result$spr$residual) <= spr_result$settings$tolerance)
)
```

## Add Beverton-Holt recruitment and calculate MSY

MSY combines yield per recruit with equilibrium recruitment. Supply unfished
recruitment (`rzero`), steepness, and the unfished spawning biomass per recruit
that defines the stock-recruitment baseline (`phi0`). For this example, that
baseline uses the same biology as the snapshot.

```{r msy}
inputs$recruitment <- list(
  type = "beverton_holt",
  rzero = 1000,
  steepness = 0.75,
  phi0 = spr_result$unfished_per_recruit$spawning_biomass
)
result <- calculate_reference_points(inputs, spr_targets = c(0.4, 0.35))
unlist(result$msy[c(
  "fishing_mortality", "yield", "biomass", "spawning_biomass",
  "recruitment", "status"
)])
```

`yield` is the MSY candidate: annual catch weight at the reported fishing
mortality. `biomass`, `spawning_biomass`, and `recruitment` describe that
equilibrium, while `fleet_yield` partitions catch among fleets. Recruitment
deviations are excluded. Nonpositive recruitment equilibria represent collapse
and have zero biomass and yield.

Currently, MSY supports Beverton-Holt recruitment with steepness strictly
between 0.2 and 1. Calculations use fixed biology and fleet shares and do not
provide uncertainty estimates.

## Inspect bounds and convergence

A search bound is a numerical setting, not a biological reference point. This
intentionally small bound illustrates two different outcomes:

```{r restricted-bounds}
restricted <- calculate_reference_points(inputs, max_f = 0.001)
restricted$spr[, c("target", "fishing_mortality", "status")]
restricted$msy[c("fishing_mortality", "status")]
```

`not_bracketed` means the SPR target was not reached within `[0, max_f]`.
`upper_bound` means the best MSY candidate was the largest allowed F; it is not
an interior optimum. Increase the bound and check the result again.

```{r wider-search}
wider <- calculate_reference_points(inputs, max_f = 10, grid_intervals = 200)
comparison <- data.frame(
  max_f = c(result$settings$max_f, wider$settings$max_f),
  F_MSY = c(result$msy$fishing_mortality, wider$msy$fishing_mortality),
  MSY = c(result$msy$yield, wider$msy$yield),
  status = c(result$msy$status, wider$msy$status)
)
knitr::kable(comparison, digits = 6)
```

An `iteration_limit` status means refinement stopped before meeting tolerance.
For SPR, `tolerance` controls the absolute SPR residual; for MSY, it controls
the width of the refined F interval. `no_positive_yield` means the MSY search
found no positive yield at its search resolution.

The MSY solver first samples a grid, then refines the best sampled region.
Convergence does not guarantee a global maximum for every possible fishing
pattern. Compare bounds and grid resolution when assessing unusual patterns.

```{r check-bounds, include=FALSE}
stopifnot(
  restricted$spr$status == "not_bracketed",
  restricted$msy$status == "upper_bound",
  result$msy$status == "converged",
  wider$msy$status == "converged",
  abs(result$msy$yield - wider$msy$yield) < 1e-5
)
```

## Compare fleet allocations

Fleet allocation is an assumption in the reference-point calculation. Add a
second fleet that selects younger fish, then compare fixed allocations while
keeping the same recruitment baseline.

```{r fleet-allocations}
allocation_results <- lapply(c(0.25, 0.5, 0.75), function(first_share) {
  scenario <- inputs
  scenario$fleets[[1]]$share <- first_share
  scenario$fleets[[2]] <- list(
    share = 1 - first_share,
    selectivity = c(0.7, 1, 0.8, 0.4, 0.2, 0.1),
    weight = inputs$weight
  )
  reference <- calculate_reference_points(scenario)
  stopifnot(reference$spr$status == "converged", reference$msy$status == "converged")
  data.frame(
    first_fleet_share = first_share,
    F_40 = reference$spr$fishing_mortality,
    F_MSY = reference$msy$fishing_mortality,
    MSY = reference$msy$yield
  )
})
knitr::kable(do.call(rbind, allocation_results), digits = 4)
```

The returned `inputs` and `settings` make each calculation reproducible. Save
those with the results when comparing scenarios. Changing a snapshot does not
change a live population model.
