 ---
title: "Introducing the Fisheries Integrated Modeling System (FIMS)"
output: github_document
vignette: >
  %\VignetteIndexEntry{Introducing the Fisheries Integrated Modeling System (FIMS)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set local options so vignette output is clean and consistent
# (avoids color codes and subtle formatting in rendered docs)
withr::local_options(list(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.subtle = FALSE
))
```

## FIMS - Surplus Production
This demo sets up a surplus production model and walks through the steps of running the Bayesian implementaion through tmbstan. The model set-up relies on "power user" functions, which is a more advanced way of accessing FIMS code than the helper functions that have been developed to set up a catch-at-age model.

```{r memory, warning=FALSE, message=FALSE}
library(FIMS)
library(ggplot2)
library(stockplotr)

# Use the {stockplotr} theme for all figures
# ggplot2::theme_set(stockplotr::theme_noaa())


# clear memory
clear()
```

## Data

Like the FIMS catch-at-age model, data are set up in long format. 

### `data_limited_tuna`

A sample data frame for a surplus production model is stored in the package as `data_limited_tuna`. This data set is based on data that was used in [Meyer and Millar 1999](https://d1wqtxts1xzle7.cloudfront.net/103328457/92d9f6b68861cc5d1c40beda0ad76d18db69-libre.pdf?1686646854=&response-content-disposition=inline%3B+filename%3DBUGS_in_Bayesian_stock_assessments.pdf&Expires=1772950518&Signature=FVjSv8rfM2XCCaCLReaKdw7ODsgA0kuclz0j82MnrjeXlJlDZUc4ysTDsW13K9cakbhBExer9xblpR3dcBA6ktXBH31Zr2snLoAih2UJ0T~VyQw~Wv8UtFG2bu5J5l8QNJE7XJ8JnBsFF3dK1O-OnzCALzjQ4593Xs8rxiMxqd7-8520e1dncKZAZgdBWGgt8uFEkFWfM62jfiL6Fzl7CB3jCw5-oUSpYcUvIDeQTf1l11K7PGj9VRtThb55M~8NSQNTe867fHMgpnJua7JvqtfmjOwzcnrZGCiMNKfrTXVPcZEaITckF79nRQOB09980CrKlR~SqTsLbXaaKUq3~A__&Key-Pair-Id=APKAJLOHF5GGSLRBV4ZA) to demonstrate a surplus production model in JAGS. The same analysis was later replicated by [Best and Punt]() to demonstrate surplus production model best practives using [Stan](). Prior and inital values used to set up the model are replicated from both analyses.

```{r load-data}
data("data_limited_tuna")
carrying_capacity_prior <- c(5.042905, sqrt(3.7603664))
growth_rate_prior <- c(-1.38, sqrt(3.845))

inits <- list(depletion=c(0.99,0.98,0.96,0.94,0.92,0.90,0.88,0.86,0.84,0.82, 0.80,0.78,0.76,0.74,0.72,0.70,0.68,0.66,0.64,0.62,0.60,0.58,0.56, 0.56), growth_rate = 0.8, carrying_capacity = 200)

nyears <- 23
```

## Index and Catch modules

The "power user" approach builds up model components using the `methods::new()` function. This function creates an Rcpp module, afterwhich values and value attributes can be set directly. Users first set up the Data modules, followed by the Fleet module for both survey indeices and fishery landings. The data modules are then linked to their respective fleet modules.

```{r setup-data-and-fleet}

# Pull data from data_limited_tuna
survey_index <- data_sp_tuna |> dplyr::filter(type == "index")
landings <- data_sp_tuna |> dplyr::filter(type == "landings")


# Create data index module and assign values to index_data
survey_fleet_index <- methods::new(Index, nyears)
purrr::walk(
1:nyears,
\(x) survey_fleet_index$index_data$set(x - 1, survey_index$value[x])
)
 
# create data landings module and assign values to landings_data
fishing_fleet_landings <- methods::new(Landings, nyears)
purrr::walk(
1:nyears,
\(x) fishing_fleet_landings$landings_data$set(x - 1, landings$value[x])
)

# Initialize the survey fleet module and link to data index module
survey_fleet <- methods::new(Fleet)
survey_fleet$n_years$set(nyears)
survey_fleet$SetObservedIndexDataID(survey_fleet_index$get_id())

# Initialize the fishing fleet module and link to data landing module
fishing_fleet <- methods::new(Fleet)
# Set number of years
fishing_fleet$n_years$set(nyears)
fishing_fleet$SetObservedLandingsDataID(fishing_fleet_landings$get_id())
```

## Distribution Modules

Likelihood distributions are separate modules and linked to the respective index and landings modules. In this surplus production model, the q-marginalization method from [Best and Punt]() is used to improve model identifiability. Under this approach, the index data are not fit directly to expected index. Rather, a trandformation of index, the ratio: $Z = log(\frac{I}{PK})$ is used instead, where $I$ is Index, $P$ is depletion, and $K$ is the population carrying capactivy. Under this formulation, the sum of $Z$ is an approzimation of the mean of $log(q)$ and $Z \sim Normal(mean(log(q)), \sigma_{obs})$. 

```{r distributions}

# Set up distribution module for survey index and link to the survey 
# fleet module and data index module
survey_fleet_index_distribution <- new(DnormDistribution)
survey_fleet_index_distribution$set_observed_data(
    survey_fleet$GetObservedIndexDataID())

# Fix sd to posterior median from Stan as FIMS currently does not have the 
# ability to fit a prior inverse gamma prior distribution on sd
survey_fleet_index_distribution$log_sd$resize(nyears)
for (y in 1:nyears) {
    survey_fleet_index_distribution$log_sd[y]$value <- 
        1.151663e-02 |> sqrt() |> log() 
}
survey_fleet_index_distribution$log_sd$set_all_estimable(FALSE)

# Link the dsitribution to observed ($Z$, a derived quantity in the model), and 
# expected (mean of $log(q)$, a derived quantity in the model)
# Using "random_effects" as a hack to implement a distribution on a derived parameter
survey_fleet_index_distribution$set_distribution_links("random_effects", 
    c(survey_fleet$log_index_to_depletion_carrying_capacity_ratio$get_id(), survey_fleet$mean_log_q$get_id()))
```


## Depletion

Surplus Production models rely on a depletion module that caotures the depletion of the population over time. FIMS currently provides one option: the Pella-Tomlison model. This model includes a shape parameter that provides flexibility in the functional form. In this example, the shape parameter is fixed at 2, which reduces the Pella-Tominson down to a Shafer model. Again, the depletion module is first created using `methods::new()`, and then parameter values and their attributes are set.

```{r depletion}
# create depletion module
production <- new(PTDepletion)

# Set initial values and attributes
production$growth_rate[1]$value <- inits$growth_rate
production$growth_rate[1]$min <- 0
production$growth_rate[1]$estimation_type$set("fixed_effects")
production$carrying_capacity[1]$value <- inits$carrying_capacity
production$carrying_capacity[1]$min <- 0
production$carrying_capacity[1]$estimation_type$set("fixed_effects")

# Fix to get Schaefer model (parameters are constant by default)
production$log_shape[1]$value <- log(2)

production$log_init_depletion[1]$value <- 0 # inital depletion ~ 1

# Set-up the depletion random effect parameter 
production$depletion$resize(nyears+1)
for (i in 1:(nyears+1)) {
    production$depletion[i]$value <- inits$depletion[i]
    production$depletion[i]$min <- 0
}
production$depletion$set_all_random(TRUE)
production$n_years$set(nyears)
```

## Depletion Distribution

Next we need to set-up the distribution for the random effects model. In this example, depletion will follow a Lognormal distribution

```{r depletion-distribution}
production_distribution <- new(DlnormDistribution)

# Fix sd to posterior median from Stan as FIMS currently does not have the 
# ability to fit a prior inverse gamma prior distribution on sd
production_distribution$log_sd[1]$value <- log(sqrt(3.1e-3))

# depletion ~ LNormal(log_expected_depletion, sd)
production_distribution$set_distribution_links(
    "random_effects",
    c(production$depletion$get_id(), production$log_expected_depletion$get_id())
)
```

## Priors

Priors can be added by initializing a new Distribution module and linking the observed and expected values to the parameter of interest and it's hypermean value. In this example, we put Lognormal priors on growth rate and carrying capacity. 

```{r priors}
growth_rate_Prior <- new(DlnormDistribution)
growth_rate_Prior$expected_values[1]$value <- growth_rate_prior[1]
growth_rate_Prior$log_sd[1]$value <- log(growth_rate_prior[2])
growth_rate_Prior$set_distribution_links("prior", 
    production$growth_rate$get_id())

carrying_capacity_Prior <- new(DlnormDistribution)
carrying_capacity_Prior$expected_values[1]$value <- carrying_capcity_prior[1]
carrying_capacity_Prior$log_sd[1]$value <- log(carrying_capacity_prior[2])
carrying_capacity_Prior$set_distribution_links("prior",         
    production$carrying_capacity$get_id())
```
  
## Population and Model

The last two modules to set up are the population and the surplus production model. The depletion and fleet modules are linked to the population module and the population module is then linked to the surplus production mdoule.

```{r population-and-surplus-production}

# Set up new population
population <- new(Population)
population$n_years$set(nyears)
population$n_ages$set(1) # only one age in surplus production
population$ages$resize(1)
population$ages$set(0, 0) # only one age in surplus production

# Link modules to population
population$SetDepletionID(production$get_id())
population$AddFleet(fishing_fleet$get_id())
population$AddFleet(survey_fleet$get_id())

# Set up surplus production model and link the population module
surplus_production <- methods::new(SurplusProduction)
surplus_production$AddPopulation(population$get_id())
```


## TMB

After the Rcpp modules are all intiialized and set-up, the user needs to call the `CreateTMBModel()` function. This function call loops over all the Rcpp modules and passes information into the respective C++ classes within the main FIMS code. After this step, functions are used to pull out parameter and random effects initial values to create the parameter list for TMB. Currently the FIMS TMB model does not require a data list because data are all passed into the model through the Rcpp data modules. The parameter list, therefore, is the only input input into `TMB::MakeADFun()`.


```{r tmb}
# Pass information into C++
CreateTMBModel()

# Create parameter list from Rcpp modules
parameters <- list(
    p = get_fixed(),
    re = get_random()
)

# Create TMB object
obj <- TMB::MakeADFun(
    data = list(), parameters, DLL = "FIMS",
    random = "re", 
    silent = TRUE, map = list()
)
```

## Optimization or MCMC

After the TMB object is created, `obj` can be passed either to `nlminb()` or `tmbstan()`. This model set-up to be run as a Bayesian model so the demo will highlight how to pass this into `tmbstan()`. The `growth_rate` and `carrying_capacity` were set-up on the natural scale to avoid the need for Jacobian transformations, which are not available yet in FIMS. It is therefore neccessary to pass in a vector of parameter minimums (all set to 0) to prevent parameter sampling outside realistic ranges. The FIMS functions `get_parameter_min_vector()` and `get_random_effects_min_vector()` pull these values from the `min` parameter attribute. 
  
```{r fit-mcmc, eval = FALSE}

opt <- stats::nlminb(obj[["par"]], obj[["fn"]], obj[["gr"]],
    lower = c(FIMS:::get_parameter_min_vector(), 
        FIMS:::get_random_effects_min_vector()), 
    upper = c(FIMS:::get_parameter_max_vector(), 
        FIMS:::get_random_effects_max_vector()), 
    control = list(eval.max = 10000, iter.max = 10000)
)

fit <- tmbstan::tmbstan(obj, init =  "last.par.best", iter = 8000,
    lower = c(FIMS:::get_parameter_min_vector(), 
        FIMS:::get_random_effects_min_vector()), 
    upper = c(FIMS:::get_parameter_max_vector(), 
        FIMS:::get_random_effects_max_vector()), 
    control = list(adapt_delta = 0.99))
```
