---
title: "FIMS Demo"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fisheries Integrated Modeling System
The NOAA Fisheries Integrated Modeling System (FIMS) is a new modeling framework for fisheries modeling. FIMS is a software system designed and architected to support next-generation fisheries stock assessment, ecosystem, and socioeconomic modeling. It's important to note that FIMS itself is not a model, but rather a framework for creating models. The framework is made up of many modules that come together to create a "the best model" that suites the needs of the end-user. What follows is a demo of creating an assessment model using FIMS.

## Creating Models in FIMS
To begin, we import the FIMS library and define some important variables that describe the dimensions of our data.
```{r fims1}
library(FIMS)

fims <- Rcpp::Module("fims", PACKAGE = "FIMS")

nyears<-30  # the number of years which we have data for.
nseanons<-1 # the number of seasons in each year.
ages<-c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)   # age vector.
nages<-12   # the number of age groups.

```


## Creating a Population
Each module in the FIMS-R interface is made of S4 objects. These S4 objects serve as a interface between R and the underlining C++ code that defines FIMS. Let's now define a population.

### Recruitment
We'll start with a recruitment module.

```{r recruitment}
# Recruitment
recruitment <- new(fims$BevertonHoltRecruitment)
recruitment$log_sigma_recruit$value <- log(0.4)
recruitment$log_rzero$value <- log(1e+06)
recruitment$log_rzero$is_random_effect <- FALSE
recruitment$log_rzero$estimated <- TRUE
recruitment$logit_steep$value <- -log(1.0 - 0.75) + log(0.75 - 0.2)
recruitment$logit_steep$is_random_effect <- FALSE
recruitment$logit_steep$estimated <- FALSE
recruitment$estimate_deviations <- FALSE
recruitment$deviations <- rep(0.01,30)

```
### Growth

### Maturity

## Creating a Fleet
### Selectivity
### Adding Data
## Creating a Survey
### Selectivity
### Adding Data
## Putting It All Together 
## Creating a TMB Model



