---
title: "Multiple recruitment phases"
output: github_document
vignette: >
  %\VignetteIndexEntry{Multiple recruitment phases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
```

Recruitment phases split one annual recruitment total across fixed dates.
Each phase has a biological age at entry and retains its own biological age
as it moves through annual age classes. Recruitment is entry into the modeled
population, not necessarily birth. Dates and fractions are configuration, not
estimated parameters or additional recruitment-process likelihoods.

This example allocates 40% of recruitment to January 1 and 60% to July 1,
both entering at biological age 1. At July 1, the early recruits have aged by
181/365 years in 2027, while the late recruits have just reached age 1.
Calendar-year age labels still advance together on January 1.

## Recruitment in the input data

For an explicit setup of the C++ modules, annual fishery and April survey data,
and their likelihoods, see [Recruitment phases with direct Rcpp
interfaces](fims-recruitment-direct.html).

```{r phase-data, message=FALSE}
library(FIMS)
library(dplyr)
observations <- data_big |>
  filter(is.na(timing) | timing <= 3 | (type == "weight_at_age" & timing == 4))
observations$timing <- ifelse(is.na(observations$timing), NA_character_,
  sprintf("%04d", observations$timing + 2026L))
july <- observations |>
  filter(fleet == "survey1", type %in% c("index", "age_comp", "length_comp"))
july$timing <- paste0(july$timing, "-07-01")
recruitment <- tibble(
  type = "recruitment_fraction",
  timing = paste0(rep(2027:2029, each = 2),
    rep(c("-01-01", "-07-01"), 3)),
  phase = rep(c("early", "late"), 3),
  observed = rep(c(0.4, 0.6), 3),
  entry_age = 1
)
frame <- FIMSFrame(bind_rows(observations, july, recruitment))
recruitment |> select(phase, timing, observed, entry_age)
```

The extra survey samples reuse example observations to illustrate the interface;
they are not new measured data. Use actual observations and uncertainties in an
assessment. Index and composition dates need not coincide with recruitment dates.

The recruitment rows belong in the same input data frame as catch, surveys, and
biological inputs. `timing` remains the input column for both observation and
recruitment dates. `observed` holds the allocated fraction for these rows; they
are fixed configuration, not additional observations or likelihood terms.
Leave `fleet`, `age`, `length`, and `uncertainty` missing. `unit` may be
`"proportion"` or missing, and `population` defaults to `"population1"`.
No separate schedule argument is needed. Normalized `date` and calendar-year
coordinates are derived internally; full dates entered through `timing` take
precedence over that derived metadata when editing normalized input tables.

Schedules require the same phase identifiers in every modeled year, unique dates,
and nonnegative fractions summing to one within 1e-8 each year. Fractions and
dates may differ between years; entry age must be constant within each phase.
The default entry age is the youngest modeled age. Dates use full ISO strings
or whole-day R Dates. The extra terminal reporting year is not an input year.

## Fit and inspect

```{r phase-fit, message=FALSE}
parameters <- setup_default_parameters(frame) |>
  filter(module_name != "Growth") |>
  bind_rows(setup_default_Growth(frame, module_type = "VonBertalanffySchnute"))
parameters$estimation_type <- "constant"
parameters$distribution[parameters$module_name == "Recruitment"] <- NA_character_
parameters$distribution_type[parameters$module_name == "Recruitment"] <- NA_character_
parameters$estimation_type[parameters$module_name == "Fleet" &
  parameters$fleet == "survey1" & parameters$label == "log_q"] <- "fixed_effects"
fit <- initialize_fims(parameters, frame) |>
  fit_fims(number_of_loops = 1)
get_max_gradient(fit)

get_estimates(fit) |>
  filter(fleet == "survey1", label == "index_expected") |>
  select(date, estimated, weight_timing)
```

The fit estimates only survey catchability to keep the example small. Shared
parametric growth and maturity are evaluated at each cohort's biological age.
Empirical weights and fixed age-to-length matrices keep their annual lookup
policy. Mortality and selectivity remain annual age-class quantities.

```{r phase-reports}
report <- get_report(fit)
report$recruitment_events
cohorts <- as.data.frame(report$recruitment_cohorts)
cohorts |> filter(year_i == 2, age_i == 2)
```

`recruitment_events` gives the amount added at each phase date. `expected_recruitment`
remains the annual recruitment budget, which differs from youngest abundance on
January 1 when some recruits arrive later. `recruitment_cohorts` gives abundance
and biological age at `available_from` within each year: zero for surviving
cohorts and the entry fraction for new cohorts. It is not a table of all cohorts
at one common observation date. `phase_i` identifies the phase in the resolved
schedule retained on the fitted input:

```{r retained-schedule}
clear()
attr(get_input(fit), "recruitment_schedule") |>
  distinct(phase_i, phase, entry_age)
```

Annual abundance, biomass, and spawning biomass reports remain January 1
post-recruitment snapshots. The phase cohort reports replace the single-age
`observation_biology` diagnostic in phase mode. Annual growth-product reports
still describe the provider's reference ages, not phase-weighted population means.

## Survival, catch, and boundaries

Before a phase enters, it contributes neither abundance nor catch. After entry,
its abundance decays by `exp(-Z * elapsed_year_fraction)`. Catch numbers use the
Baranov calculation over the remaining part of that year. Catch weight and
fishery length compositions use eight-point Gauss-Legendre quadrature over each
cohort's exposure interval, normalized to its exact catch numbers. These
biological integrals are approximations; their accuracy should be checked for
unusually sharp growth or length-bin transitions. Survey dates do not determine
the integration nodes and cannot change annual catch.

Same-day surveys observe newly recruited fish. Annual survival is completed
before aging and January 1 recruitment. December 31 recruits experience one day
of mortality before the next annual boundary, using 365 or 366 days as appropriate.

## Initial, reference, and terminal conditions

- The youngest `log_init_naa` parameter defines the first year's total recruitment
  budget. It is allocated across phases rather than also inserted in full on January 1.
- Older initial age-class totals retain their existing meaning. Their phase shares
  use unfished survival under the first modeled year's schedule and natural mortality.
- Unfished spawning biomass per recruit uses that same first-year reference schedule,
  including its calendar-year fractions, entry ages, biological products, and mortality.
  Later annual schedules drive subsequent unfished trajectories. A changing schedule
  need not maintain the initial reference equilibrium.
- The oldest plus group uses a representative biological age for each phase.
  Cohorts retain their entry-time offset but stop advancing their representative
  age after reaching the plus group; exact individual ages inside it are not tracked.
- The extra terminal January 1 report uses mean recruitment without a process
  deviation. It repeats the final modeled year's phase pattern and inserts only
  its January 1 allocation; later unmodeled events are not inserted.

With one January 1 phase, fraction one, and the default entry age, FIMS uses the
original annual calculation exactly. The alternative initialization and integrated
catch-biology conventions above apply only when phase mode is enabled.

## Retrospectives and projections

`run_modified_data_fims()` and `run_fims_retrospective()` retain the recruitment
rows automatically. Peels retain the annual catch horizon, so the full schedule
remains applicable. Likelihood profiles preserve the fitted schedule as well.

Existing callers may still pass an explicit `recruitment_schedule` accepted by
`setup_recruitment_schedule()`. It accepts `timing` for full dates and retains
support for the earlier `date` column. If a table also contains recruitment
rows, an explicit schedule must agree with those rows; conflicts are rejected.

For projections, add recruitment rows for every new modeled year as well as
the annual fishing and biological inputs. Missing schedule years are rejected;
FIMS does not silently supply January 1 recruitment in those years.
