---
title: "FIMS Path: Growth (VonBertalanffy)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FIMS Path: Growth (VonBertalanffy)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
.rchunk {
  border: solid #2e75b6;
}
.rcppchunk {
  border: solid #c55a11;
}
.fimschunk {
  border: solid #548235;
}
.infochunk {
  border: solid #262626;
}
```

<style type="text/css">
code span.co{
  color: #8c8c8c;
}
</style>

This vignette walks through the new Von Bertalanffy growth pathway in FIMS.
It is intended to help explain the changes that make growth an active and
estimable part of the model, run the new pathway using the standard FIMS
workflow, and inspect the outputs needed to confirm that the branch is working
as expected.

The first example uses `data_big` so the new growth pathway can be explored
within a familiar FIMS model setup. The main worked example then uses generated
data with a wider age range and known growth values. This makes it easier to
estimate the growth curve, compare fitted results with the known truth, and
explore how changes to growth inputs affect the model.

The Von Bertalanffy model uses the familiar Stock Synthesis 3 two reference-age
parameterization. The growth curve is defined by length at two reference ages
and the growth coefficient `K`. In the examples, the following three parameters
are estimated:
- `length_at_ref_age_1`
- `length_at_ref_age_2`
- `growth_coefficient_K`

FIMS uses the fitted growth curve and supplied variability in length-at-age to
prepare mean length-at-age, standard deviation of length-at-age, and mean
weight-at-age. These results then feed into the shared population size system
and the growth-derived age-to-length conversion used by the length-based fleet
calculations.

This vignette uses the interpolation-based variability pathway, which supplies
standard deviation at the two reference ages through
`length_at_age_sd_at_ref_ages`.

Along the way, the vignette highlights outputs and settings that reviewers can
change to explore the implementation and check that the growth-derived pathway
is behaving as expected.

> Note: The main worked example uses the interpolation-based variability path
> by supplying `length_at_age_sd_at_ref_ages`. The delta-method path is
> wired into FIMS, but it is not used here because it depends on
> the traditional Von Bertalanffy parameterization, which is not yet available.

## Brief `data_big` Setup

The first example uses `data_big` to show where the Von Bertalanffy growth fits
into the standard FIMS workflow. The same sequence used in other FIMS vignettes
to prepare the data, configurations, create the parameter table, initialize the
model, fit it, and inspect the results.

`data_big` includes historical fixed `age_to_length_conversion` rows. They do
not need to be removed for this example. Once the Growth module is changed to
`VonBertalanffy`, FIMS infers the growth-derived ALK path from the Growth
configuration and available data instead of asking users to choose it manually.
For that path, explicit `length_bin` rows define the fleet observation bins when
present. If `length_bin` rows are absent, fleet-specific `length_comp` bins can
serve as compatibility observation geometry. Fixed `age_to_length_conversion`
rows are used only for the fixed ALK path and do not define fleet observation
bins for the growth-derived ALK path.

```{r}
#| warning: false
#| message: false
# Load the packages used in the example.
library(FIMS)
library(dplyr)
library(tidyr)

# Clear objects left in C++ memory by earlier FIMS runs.
FIMS::clear()

# Load FIMS example data and prepare the FIMSFrame.
 data("data_big")
fims_frame <- FIMS::FIMSFrame(data_big)

# Start from the default model configurations and replace the default
# empirical weight-at-age growth module with Von Bertalanffy growth.
configurations <- FIMS::create_default_configurations(data = fims_frame) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

# Create the parameter table for the updated configuration.
# The three core VonB curve parameters are estimable by default.
parameters <- FIMS::create_default_parameters(
  configurations = configurations,
  data = fims_frame
) |>
  tidyr::unnest(cols = data)

# Inspect the default growth inputs and their estimation settings.
parameters |>
  dplyr::filter(module_name == "Growth") |>
  dplyr::select(label, age, value, estimation_type)
```

The default Von Bertalanffy parameter table makes
`length_at_ref_age_1`, `length_at_ref_age_2`, and
`growth_coefficient_K` estimable. The reference ages, length-weight
parameters, and interpolation-based standard-deviation inputs remain fixed.

Selecting `VonBertalanffy` produces an estimable growth setup through the
standard configuration and parameter workflow without requiring users to make
extra manual edits to the growth parameter table.

```{r}
# Initialize the complete FIMS model.
input <- parameters |>
  FIMS::initialize_fims(data = fims_frame)

# Request the full realized growth-derived ALK tensor for inspection.
input$model$ReportGrowthDerivedALKTensor(TRUE)

# Fit the model
fit <- FIMS::fit_fims(
  input = input,
  optimize = TRUE,
  number_of_loops = 5,
  get_sd = FALSE
)

# Keep the Growth-module estimates from the fitted model.
growth_estimates <- FIMS::get_estimates(fit) |>
  dplyr::filter(module_name == "Growth")

# Inspect the three estimable Von Bertalanffy curve parameters.
growth_estimates |>
  dplyr::filter(
    label %in% c(
      "length_at_ref_age_1",
      "length_at_ref_age_2",
      "growth_coefficient_K"
    )
  ) |>
  dplyr::select(label, input, estimated, estimation_type)

# Check optimizer convergence and the maximum absolute gradient.

FIMS::get_opt(fit)$convergence
FIMS::get_max_gradient(fit)
```

For a successful branch check, the three estimated parameters should remain
finite and positive, the estimated second reference length should be greater
than the first, and the maximum gradient should be small.


```{r}
# Pull the report here because these diagnostics are only available there.
report <- FIMS::get_report(fit)

# A value of 1 indicates that the fleet used the growth-derived ALK.
report[["growth_derived_alk_used"]][[1]]

# A length of 0 indicates that no fixed ALK matrix is stored.
length(report[["age_to_length_conversion"]][[1]])

# These prepared growth products should contain finite values.

# This is the sdLAA, one value per year x age combination.
report[["growth_sd_LAA"]][[1]]

# This is the mean WAA, one value per year x age combination.
report[["growth_mean_WAA"]][[1]]

# This is the fleet-mapped mean WAA that the fleet path uses.
report[["growth_derived_mean_WAA"]][[1]]
```

## Main Example: Estimable VonB With Generated Data

This example uses generated data with ages 1-30 and "known" growth parameters.
The wider age range makes the shape of the fitted curve easier to inspect than
in the previous `data_big` example. Because the values used to generate the
data are retained, you can compare the estimated parameters and prepared
growth products with a "known truth".

### Build a Compact Generated Data Set

The helper below creates one population, one fleet, and two model years. Two
years are used so the generated data form a complete catch-at-age model while
remaining small enough for interactive branch testing.

The generated data include:
- age-composition proportions
- length-composition proportions
- weight-at-age values
- a landings observation
- explicit fleet length-bin rows
- the true mean length-at-age and standard deviation of length-at-age

It also generates a small set of noisy length-at-age observations for the
comparison plot. Those plotting observations are retained separately and are
not passed to FIMS as model inputs.

```{r}
#| warning: false
#| message: false
library(FIMS)
library(dplyr)
library(tidyr)
library(ggplot2)

# Define the SS3-style two-reference-age Von Bertalanffy equation used to
# generate the "known" growth truth for this example.
vonb_two_reference_age <- function(
    age,
    length_at_ref_age_1,
    length_at_ref_age_2,
    growth_coefficient_K,
    reference_age_for_length_1,
    reference_age_for_length_2) {
  # Normalizes the curve so it is anchored to both reference-age lengths.
  denominator <- 1 - exp(
    -growth_coefficient_K *
      (reference_age_for_length_2 - reference_age_for_length_1)
  )

  # Predict mean length at age using the same curve shape FIMS estimates.
  length_at_ref_age_1 +
    (length_at_ref_age_2 - length_at_ref_age_1) *
      (
        1 - exp(
          -growth_coefficient_K *
            (age - reference_age_for_length_1)
        )
      ) / denominator
}

# Calculate the asymptotic length implied by the two-reference-age form.
vonb_implied_linf <- function(
    length_at_ref_age_1,
    length_at_ref_age_2,
    growth_coefficient_K,
    reference_age_for_length_1,
    reference_age_for_length_2) {
  length_at_ref_age_1 +
    (length_at_ref_age_2 - length_at_ref_age_1) /
      (
        1 - exp(
          -growth_coefficient_K *
            (reference_age_for_length_2 - reference_age_for_length_1)
        )
      )
}

# Build a compact generated data set with one population and one fleet.
# FIMS fits the model later; this function only creates the example data.
make_vonb_demo_data <- function(
    ages = 1:30,
    lengths = seq(0, 1400, 50),
    years = 1:2,
    fleet = "fleet_demo",
    sample_size = 300,
    length_at_ref_age_1 = 180,
    length_at_ref_age_2 = 850,
    growth_coefficient_K = 0.18,
    reference_age_for_length_1 = 1,
    reference_age_for_length_2 = 12,
    length_weight_a = 2.5e-11,
    length_weight_b = 3,
    sd_at_ref_ages = c(30, 73)) {
  # Calculate the "true" growth products used to generate the example.
  true_curve <- tibble::tibble(
    age = ages,
    true_mean_length = vonb_two_reference_age(
      age = ages,
      length_at_ref_age_1 = length_at_ref_age_1,
      length_at_ref_age_2 = length_at_ref_age_2,
      growth_coefficient_K = growth_coefficient_K,
      reference_age_for_length_1 = reference_age_for_length_1,
      reference_age_for_length_2 = reference_age_for_length_2
    ),
    true_sd_length = stats::approx(
      x = c(
        reference_age_for_length_1,
        reference_age_for_length_2
      ),
      y = sd_at_ref_ages,
      xout = ages,
      rule = 2
    )$y
  ) |>
    dplyr::mutate(
      true_weight_at_age =
        length_weight_a * true_mean_length^length_weight_b
    )

  # Derive the implied Linf from the known two-reference-age growth parameters.
  # Linf is not estimated directly here; it is used only as a truth reference.
  true_linf <- vonb_implied_linf(
    length_at_ref_age_1 = length_at_ref_age_1,
    length_at_ref_age_2 = length_at_ref_age_2,
    growth_coefficient_K = growth_coefficient_K,
    reference_age_for_length_1 = reference_age_for_length_1,
    reference_age_for_length_2 = reference_age_for_length_2
  )

  # Use a declining age distribution for the generated population.
  age_weights <- exp(-0.18 * (ages - min(ages)))
  age_probabilities <- age_weights / sum(age_weights)

  # Create age-composition observations for each modeled year and age.
  # The same known age distribution is repeated across years so the generated
  # data give the model age-structure information without adding extra
  # year-specific variation.
  age_comp_rows <- tidyr::expand_grid(
    timing = years,
    age = ages
  ) |>
    dplyr::mutate(
      type = "age_comp",
      fleet = fleet,
      length = NA_real_,
      value = age_probabilities[match(age, ages)],
      unit = "proportion",
      uncertainty = sample_size
    ) |>
    dplyr::select(
      type,
      fleet,
      age,
      length,
      timing,
      value,
      unit,
      uncertainty
    )

  # Calculate the probability of each modeled length conditional on age.
  age_length_probabilities <- tidyr::expand_grid(
    age = ages,
    length = lengths
  ) |>
    dplyr::left_join(true_curve, by = "age") |>
    dplyr::group_by(age) |>
    dplyr::mutate(
      conditional_probability = stats::dnorm(
        x = length,
        mean = true_mean_length,
        sd = true_sd_length
      ),
      conditional_probability =
        conditional_probability / sum(conditional_probability)
    ) |>
    dplyr::ungroup()

  # Marginalize across ages to create the fleet length composition.
  length_comp_rows <- age_length_probabilities |>
    dplyr::mutate(
      age_probability = age_probabilities[match(age, ages)]
    ) |>
    dplyr::group_by(length) |>
    dplyr::summarise(
      value = sum(age_probability * conditional_probability),
      .groups = "drop"
    ) |>
    tidyr::crossing(timing = years) |>
    dplyr::mutate(
      type = "length_comp",
      fleet = fleet,
      age = NA_real_,
      unit = "proportion",
      uncertainty = sample_size
    ) |>
    dplyr::select(
      type,
      fleet,
      age,
      length,
      timing,
      value,
      unit,
      uncertainty
    )

  # Use the known growth curve to create weight-at-age inputs.
  # These values give the generated example a truth-based weight-at-age source
  # that can be compared with the growth-derived WAA reported after fitting.
  weight_at_age_rows <- true_curve |>
    dplyr::transmute(
      type = "weight_at_age",
      fleet = fleet,
      age = age,
      length = NA_real_,
      timing = min(years),
      value = true_weight_at_age,
      unit = "mt",
      uncertainty = NA_real_
    )

  # Add simple landings observations for each modeled year.
  # These keep the generated model connected to a fleet-level catch component
  landings_rows <- tibble::tibble(
    type = "landings",
    fleet = fleet,
    age = NA_real_,
    length = NA_real_,
    timing = years,
    value = rep(2500, length(years)),
    unit = "mt",
    uncertainty = 0.2
  )

  # Make the fleet observation geometry explicit so it can be modified.
  length_bin_rows <- tibble::tibble(
    type = "length_bin",
    fleet = fleet,
    age = NA_real_,
    length = lengths,
    timing = NA_real_,
    value = NA_real_,
    unit = "proportion",
    uncertainty = NA_real_
  )

  # Combine all generated row types into one long FIMS input table.
  # The resulting table contains the fleet observations, weight-at-age inputs,
  # and length-bin geometry used to build the example FIMSFrame.
  raw_data <- dplyr::bind_rows(
    landings_rows,
    age_comp_rows,
    length_comp_rows,
    weight_at_age_rows,
    length_bin_rows
  )

  # Create separate noisy observations for plotting only.
  observed_laa <- true_curve |>
    dplyr::mutate(
      observed_length = stats::rnorm(
        n = dplyr::n(),
        mean = true_mean_length,
        sd = true_sd_length / sqrt(15)
      )
    ) |>
    dplyr::select(
      age,
      true_mean_length,
      true_sd_length,
      observed_length
    )

  # Return both the FIMS-ready data object and the known-truth pieces used for
  # later checks and plots. Keeping these together makes each scenario easy to
  # rerun with different growth or bin settings.
  list(
    frame = FIMS::FIMSFrame(raw_data),
    raw_data = raw_data,
    ages = ages,
    lengths = lengths,
    years = years,
    true_curve = true_curve,
    observed_laa = observed_laa,
    true_linf = true_linf,
    true_parameters = tibble::tibble(
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      true_value = c(
        length_at_ref_age_1,
        length_at_ref_age_2,
        growth_coefficient_K,
        reference_age_for_length_1,
        reference_age_for_length_2
      )
    )
  )
}

# Set the seed so the generated observed length-at-age values are reproducible.
set.seed(1908)

# Build the generated VonB example data
demo <- make_vonb_demo_data()

# Inspect the data types and modeled ranges before configuring the model.
methods::show(demo$frame)
demo$true_parameters
demo$true_linf
```

The explicit `length_bin` rows make the fleet observation geometry easy to
locate and modify during branch testing. You can change the spacing or
range in `lengths` and rerun the example to examine how fleet geometry is
carried into the growth-derived pathway.

### Configure the Von Bertalanffy Model

The generated data use the same standard configuration and parameter workflow
as the `data_big` example. Changing the Growth module to
`VonBertalanffy` causes `create_default_parameters()` to create the required
reference ages, interpolation-based variability inputs, and estimable curve
parameters.

By default, the reference ages are set from the minimum and maximum ages in the
`FIMSFrame`. This example overrides the second reference age to 12 so that the
modeled ages beyond age 12 show the fitted curve approaching its implied
asymptote.

```{r}
# Clear C++ objects created by the previous example.
FIMS::clear()

# Identify the three estimable VonB curve parameters used for comparisons,
# plots, and checks throughout the generated-data example.
core_growth_labels <- c(
  "length_at_ref_age_1",
  "length_at_ref_age_2",
  "growth_coefficient_K"
)

# Replace the default empirical weight-at-age module with VonB growth.
configurations <- FIMS::create_default_configurations(data = demo$frame) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

# Create the complete model parameter table.
parameters <- FIMS::create_default_parameters(
  configurations = configurations,
  data = demo$frame
) |>
  tidyr::unnest(cols = data)
```

The three core curve parameters are already `"fixed_effects"`. The example
changes their initial values so the optimizer does not begin at the values used
to generate the data.

```{r}
# Pull out the default VonB interpolation SD rows.
growth_sd_rows <- parameters |>
  dplyr::filter(
    module_name == "Growth",
    label == "length_at_age_sd_at_ref_ages"
  ) |>
  dplyr::mutate(
    # Move the interpolation ages to the demo reference ages.
    age = c(1, 12),
    # Match the generated-truth SD values for this example.
    value = c(30, 73)
  )

# Replace the defaults with the demo settings.
parameters <- parameters |>
  dplyr::filter(
    !(module_name == "Growth" &
        label == "length_at_age_sd_at_ref_ages")
  ) |>
  dplyr::bind_rows(growth_sd_rows) |>
  dplyr::rows_update(
    tibble::tibble(
      module_name = "Growth",
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      value = c(150, 800, 0.22, 1, 12)
    ),
    by = c("module_name", "label")
  )

# Review the Growth configuration before initializing the model.
parameters |>
  dplyr::filter(module_name == "Growth") |>
  dplyr::arrange(label, age) |>
  dplyr::select(label, age, value, estimation_type)
```

You can change the three starting values and rerun the fit to examine
whether the model reaches the same solution. The two SD-anchor values can also
be changed to explore how assumed variability in length-at-age affects the
growth-derived age-to-length probabilities.

### Fit the Model

Initialize and fit the model using the prepared parameter table. The report
flag requests the complete realized growth-derived ALK tensor, which is used
later to inspect the handoff from growth to fleet length bins.

```{r}
# Build the FIMS input from the prepared demo frame and parameters.
input <- parameters |>
  FIMS::initialize_fims(data = demo$frame)

# Ask FIMS to include the growth-derived ALK tensor in the report
# so we can inspect it.
input$model$ReportGrowthDerivedALKTensor(TRUE)

# Fit the VonB demo
fit <- FIMS::fit_fims(
  input = input,
  optimize = TRUE,
  number_of_loops = 5,
  number_of_newton_steps = 0,
  get_sd = FALSE
)

# Keep only the core VonB growth estimates.
growth_estimates <- FIMS::get_estimates(fit) |>
  dplyr::filter(
    module_name == "Growth",
    label %in% core_growth_labels
  )
```

Compare the starting values and estimates with the known parameters used to
generate the data:

```{r}
# Compare the fitted growth values against the generated truth
growth_comparison <- growth_estimates |>
  dplyr::select(label, input, estimated) |>
  dplyr::left_join(
    demo$true_parameters |>
      dplyr::filter(label %in% core_growth_labels),
    by = "label"
  )

# Show the side-by-side comparison for the core VonB parameters.
growth_comparison
```

Check both the optimizer convergence code and the maximum gradient:

```{r}
tibble::tibble(
  convergence = FIMS::get_opt(fit)$convergence,
  max_gradient = FIMS::get_max_gradient(fit)
)
```

A convergence code of `0` indicates normal optimizer convergence. The maximum
gradient should also be small. These diagnostics should be considered
together with whether the estimated growth parameters are biologically valid
and reasonably close to the known values.

### Plot the Observed and Predicted Growth Curve

The fitted FIMS report is defined on the modeled ages, but the plotting grid is
extended beyond the modeled ages to show the implied Von Bertalanffy asymptote.
The extended estimated curve is calculated from the fitted parameters and is
for visualization only.

```{r growth-curve-plot}
#| warning: false
#| message: false
#| fig.width: 9
#| fig.height: 6
#| fig.cap: >
#|   Observed lengths-at-age, the true Von Bertalanffy curve used to generate
#|   the data, and the fitted Von Bertalanffy curve from FIMS.
#| fig.alt: >
#|   Point and line plot comparing generated observed lengths-at-age with the
#|   true and fitted Von Bertalanffy growth curves, with the plotting range
#|   extended to show the implied asymptote.

# Pull true parameter values into one row.
true_growth_values <- demo$true_parameters |>
  tidyr::pivot_wider(
    names_from = label,
    values_from = true_value
  )

# Pull estimated curve parameters into one row.
estimated_growth_values <- growth_estimates |>
  dplyr::select(label, estimated) |>
  tidyr::pivot_wider(
    names_from = label,
    values_from = estimated
  )

# Pull the reference ages used by the fitted model.
growth_reference_ages <- parameters |>
  dplyr::filter(
    module_name == "Growth",
    label %in% c(
      "reference_age_for_length_1",
      "reference_age_for_length_2"
    )
  ) |>
  dplyr::select(label, value) |>
  tidyr::pivot_wider(
    names_from = label,
    values_from = value
  )

# Extend the plotting grid beyond the modeled ages so the asymptote is visible.
plot_ages <- seq(
  from = min(demo$ages),
  to = 50,
  by = 0.25
)

true_plot_curve <- tibble::tibble(
  age = plot_ages,
  mean_length = vonb_two_reference_age(
    age = plot_ages,
    length_at_ref_age_1 = true_growth_values$length_at_ref_age_1,
    length_at_ref_age_2 = true_growth_values$length_at_ref_age_2,
    growth_coefficient_K = true_growth_values$growth_coefficient_K,
    reference_age_for_length_1 =
      true_growth_values$reference_age_for_length_1,
    reference_age_for_length_2 =
      true_growth_values$reference_age_for_length_2
  ),
  curve = "True curve",
  series = "True curve"
)

estimated_plot_curve <- tibble::tibble(
  age = plot_ages,
  mean_length = vonb_two_reference_age(
    age = plot_ages,
    length_at_ref_age_1 = estimated_growth_values$length_at_ref_age_1,
    length_at_ref_age_2 = estimated_growth_values$length_at_ref_age_2,
    growth_coefficient_K = estimated_growth_values$growth_coefficient_K,
    reference_age_for_length_1 =
      growth_reference_ages$reference_age_for_length_1,
    reference_age_for_length_2 =
      growth_reference_ages$reference_age_for_length_2
  ),
  curve = "Estimated curve",
  series = "Estimated curve"
)

curve_lines <- dplyr::bind_rows(
  true_plot_curve,
  estimated_plot_curve
)

# Keep a model-age prediction table for residuals.
predicted_curve <- tibble::tibble(
  age = demo$ages,
  predicted_mean_length = vonb_two_reference_age(
    age = demo$ages,
    length_at_ref_age_1 = estimated_growth_values$length_at_ref_age_1,
    length_at_ref_age_2 = estimated_growth_values$length_at_ref_age_2,
    growth_coefficient_K = estimated_growth_values$growth_coefficient_K,
    reference_age_for_length_1 =
      growth_reference_ages$reference_age_for_length_1,
    reference_age_for_length_2 =
      growth_reference_ages$reference_age_for_length_2
  )
)

curve_plot_data <- demo$observed_laa |>
  dplyr::left_join(predicted_curve, by = "age") |>
  dplyr::mutate(series = "Generated observations")

true_linf_line <- tibble::tibble(
  yintercept = demo$true_linf,
  series = "True Linf"
)

ggplot2::ggplot() +
  ggplot2::geom_point(
    data = curve_plot_data,
    ggplot2::aes(x = age, y = observed_length, colour = series),
    size = 2
  ) +
  ggplot2::geom_line(
    data = curve_lines,
    ggplot2::aes(x = age, y = mean_length, colour = series),
    linewidth = 1
  ) +
  ggplot2::geom_hline(
    data = true_linf_line,
    ggplot2::aes(yintercept = yintercept, colour = series),
    linetype = "dotted",
    linewidth = 0.8
  ) +
  ggplot2::scale_colour_manual(
    name = NULL,
    breaks = c(
      "Generated observations",
      "Estimated curve",
      "True curve",
      "True Linf"
    ),
    values = c(
      "Generated observations" = "black",
      "Estimated curve" = "#0072B2",
      "True curve" = "#D55E00",
      "True Linf" = "gray50"
    )
  ) +
  ggplot2::labs(
    title = "Observed and predicted VonB growth curve",
    x = "Age",
    y = "Length"
  ) +
  stockplotr::theme_noaa()
```

The estimated curve should follow the generated truth closely when the fit
converges and the length-composition information is informative. Large
departures are useful for debugging because they can point to growth starting
values, length-composition information, or downstream length-bin behavior.

### Plot Residuals

Here, the residual is the generated observed length minus the fitted mean
length-at-age from `get_estimates()`. This is not a truth-vs-fit check; it is
just a visual check that the fitted curve tracks the demo observations.

```{r growth-residual-plot}
#| warning: false
#| message: false
#| fig.width: 8
#| fig.height: 8
#| fig.cap: >
#|   Residual plot showing observed minus predicted length-at-age for the
#|   generated Von Bertalanffy example.
#| fig.alt: >
#|   Residual plot with age on the x-axis and observed minus predicted length
#|   on the y-axis for the generated Von Bertalanffy example.

# Build residuals from the fitted VonB curve
residual_plot_data <- demo$observed_laa |>
  dplyr::left_join(predicted_curve, by = "age") |>
  dplyr::mutate(
    residual = observed_length - predicted_mean_length
  )

ggplot2::ggplot(residual_plot_data, ggplot2::aes(x = age, y = residual)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dotted") +
  ggplot2::geom_point(size = 2) +
  ggplot2::labs(
    title = "Length-at-age residuals",
    x = "Age",
    y = "Observed minus predicted length"
  ) +
  stockplotr::theme_noaa()
```

These residuals are not a formal model diagnostic because the observed lengths
were created only for plotting. They are still useful for a branch-level check
because they make it easy to see whether the fitted curve is tracking the
known growth pattern.

### One Light Downstream Check of the Shared Size Path

The final plot is not meant to duplicate the test suite. It shows one visible
downstream object from the migrated pathway: a realized growth-derived
age-to-length row for a representative age.

```{r growth-size-handoff-plot}
#| warning: false
#| message: false
#| fig.width: 8
#| fig.height: 8
#| fig.cap: >
#|   One realized growth-derived age-to-length row, showing how the fitted
#|   growth curve is mapped into fleet observation-bin probabilities.
#| fig.alt: >
#|   Bar plot of one realized growth-derived age-to-length row across fleet
#|   observation-bin centers for a representative age.

# Pull the report here because the realized ALK tensor and path flags are
# only available there.
report <- FIMS::get_report(fit)

# View the ages in the demo object to understand the age range.
demo$ages

# Pick a representative age/year slice so we can show one realized ALK row.
# `ceiling(length(demo$ages) / 2)` picks the middle modeled age, rounding up
# when there are an even number of ages.
representative_age <- demo$ages[ceiling(length(demo$ages) / 2)]
representative_year <- demo$years[1]

# Pull the full realized growth-derived ALK tensor from the report.
realized_alk <- report[["growth_derived_age_to_length_conversion"]][[1]]

# Reshape the flattened tensor into length x age x year form for slicing.
realized_alk_array <- array(
  realized_alk,
  dim = c(
    length(demo$lengths),
    length(demo$ages),
    length(demo$years)
  )
)

# Extract one age-year row so the probability mass across fleet bins is easy to read.
downstream_check <- tibble::tibble(
  length = demo$lengths,
  probability = realized_alk_array[
    ,
    match(representative_age, demo$ages),
    match(representative_year, demo$years)
  ]
)

# Plot the realized ALK row as probabilities over the fleet observation bins.
ggplot2::ggplot(
  downstream_check,
  ggplot2::aes(x = length, y = probability)
) +
  ggplot2::geom_col() +
  ggplot2::labs(
    title = paste(
      "Realized growth-derived ALK row at age",
      representative_age,
      "in year",
      representative_year
    ),
    x = "Fleet observation bin",
    y = "Probability"
  ) +
  stockplotr::theme_noaa()
```

The probabilities in the plotted row should sum to one:

```{r}
sum(downstream_check$probability)
```

For a quick non-graphical check, the report flag should also be `1` when the
derived path is being used:

```{r}
report[["growth_derived_alk_used"]][[1]][1]
```

The fixed historical age-to-length matrix should not be present on the active
growth-derived path:

```{r}
length(report[["age_to_length_conversion"]][[1]])
```

## Architecture Path

This branch changed more than "turning on growth estimation". The Growth
module now prepares growth products, including mean length-at-age, standard
deviation of length-at-age, and mean weight-at-age. But the larger change is
that these products now move through the new size subsystem before they are
used by fleets.

The size subsystem separates population biology from fleet observation
biology. Growth defines the biological size structure of the population. Those
growth products are then used to build `ProbSize`, the probability that a fish
of a given year and age falls into a particular population size bin. The
growth-derived age-to-length pathway then maps those population size
probabilities onto the fleet observation bins used by the length-composition
data.
The biological size grid remains separate from fleet observation bins, and
user-specified size grids are preserved when provided; FIMS only builds a
default grid when needed.

Conceptually, the path is:

```text
estimated VonB growth parameters
        |
        v
mean length-at-age and SD length-at-age
        |
        v
population size probabilities
        |
        v
fleet observation-bin probabilities
        |
        v
length-composition and landings calculations
```

## Next Steps

This vignette shows the baseline Von Bertalanffy growth-derived pathway. A
companion vignette, `Things To Try With Von Bertalanffy Growth.rmd`, walks through
small changes that can be made to the generated example to see how growth
starting values, fleet length bins, length-at-age variability, reference ages,
report settings, and downstream model outputs respond.
