---
title: "Simulation and recovery with recruitment phases"
output: github_document
vignette: >
  %\VignetteIndexEntry{Simulation and recovery with recruitment phases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
```

This exercise tests whether the direct-interface model can recover known fishing
mortality and survey catchability from observations generated by that model.
It follows [Recruitment phases with direct Rcpp interfaces](fims-recruitment-direct.html).
The executable helpers are installed with FIMS; they build each model through
Rcpp and TMB, without the FIMS model-setup or fitting wrappers.

The helper is deliberately limited to the three-year, five-age design below.
Before changing the FIMS registry, it checks that observations are finite and
positive, composition count matrices have three rows and five columns with
positive row totals, and each survey date belongs to its stated model year.
Years must be ordered 2027–2029. Starting values must be positive and ordered
`F_2027`, `F_2028`, `F_2029`, `q`; named vectors must use those exact names.

Use a fresh R session. Each fit releases its TMB objective before clearing the
global FIMS registry and returns only ordinary R values. These helpers should
not run alongside another live FIMS objective.

## Design

The generating model has three years (2027–2029), five age classes, January and
July recruitment fractions of 0.4 and 0.6, and April 1 surveys. It uses the same
fixed growth, maturity, selectivity, initial population, and natural mortality as
the direct-interface example. The estimated quantities are:

| Parameter | Generating value |
|:--|--:|
| Fishing mortality, 2027 | 0.12 |
| Fishing mortality, 2028 | 0.20 |
| Fishing mortality, 2029 | 0.28 |
| Survey catchability | 0.03 |

Recruitment dates, fractions, and entry ages remain fixed. Recruitment deviations
are zero, so this is an observation-error experiment, not a recruitment-process
simulation. We do not estimate the biological parameters or initial abundance.

Each replicate contains annual catch and age compositions plus April index and
age compositions. Catch and index are drawn from lognormal distributions with
log-scale SDs 0.20 and 0.25. Age counts are multinomial samples of 200 fish per
annual fishery sample and 150 fish per survey. Samples are drawn independently
conditional on the generating population.

The lognormal `meanlog` is the log of the model prediction. The prediction is
therefore the median, not the arithmetic mean; no mean-bias correction is added.
All random observations are drawn in R after obtaining predictions from FIMS.

## Run the exercise

The source file contains the explicit module connections, simulation functions,
optimizer, numerical Hessian, and summaries. It does not depend on sourcing the
interactive example or modifying its text.

```{r recovery-run}
library(FIMS)
source(system.file("examples", "recruitment-phase-recovery.R", package = "FIMS"))
result <- run_recruitment_recovery(n_replicates = 30, seed = 20260924)
```

Thirty replicates keep this vignette quick and illustrate the workflow. For a
more precise study, increase `n_replicates` to several hundred or more and inspect
all convergence diagnostics. The seed must be a nonnegative integer and `n_replicates` an integer of at least
two, both within R's integer range. The function returns its seed and raw replicate
results, replicate count, and `sessionInfo()` (including package versions).
It also restores the caller's random-number state. Save the complete result for
reproduction and inspection:

```{r save-recovery, eval=FALSE}
saveRDS(result, "recovery.rds")
```

## Exact-data recovery from different starting points

First we evaluate the model at the generating parameters. Catch and index are
set exactly to their predicted medians. Compositions use expected counts
(sample size times predicted probabilities), without rounding. FIMS accepts
these fractional counts, while base R's `dmultinom()` rounds counts, so the latter
is not an equivalent likelihood check for this deterministic case.

The model is fitted from three starting vectors: half the truth, 1.8 times the
truth, and a mixture of low and high values. This checks the optimizer and the
parameter/data connections without sampling noise.

```{r exact-recovery}
knitr::kable(result$exact_recovery, digits = 6)
```

The runner requires convergence code zero, maximum absolute gradient below
`0.001`, and relative parameter error below `0.001` for every exact-data fit.
These numerical tolerances are validation checks, not requirements that noisy
estimates equal the generating values.

## Independent observation-likelihood check

At three parameter vectors away from the optimum, the runner independently
calculates the lognormal and multinomial observation likelihoods in R. The
multinomial calculation uses `lgamma()` so fractional expected counts are
preserved. Likelihood differences relative to the fitted exact-data model must
agree with TMB within `1e-6`.

```{r independent-likelihood}
knitr::kable(result$likelihood_check, digits = 8)
```

This checks the observation likelihood connections and scales; it still uses
FIMS predictions and does not independently validate population dynamics.

## What changes if all recruitment enters in January?

The same observations are also fitted with fractions 1 and 0 for January and
July. The zero-weight July phase deliberately keeps the **phase evaluator**
active, so the comparison uses the same biological integration and
initialization rules. It does not switch to the legacy single-phase annual
calculation. Phase survival shares at initialization still respond to the
schedule; this is a comparison of complete timing assumptions, not just one
isolated survey prediction.

```{r timing-comparison}
knitr::kable(result$timing_comparison, digits = 5)
knitr::kable(result$timing_parameters, digits = 5)
```

The runner stops if either comparison fit fails the usability checks described
below. The table exposes that status before likelihood differences are interpreted.

Both models have four estimated parameters and use identical data and likelihood
constants. Their negative log likelihoods can therefore be compared directly.
This is a sensitivity comparison between fixed schedules, not a likelihood-ratio
test with a claimed chi-squared reference distribution.

## Noisy recovery and uncertainty

For each replicate, both schedules are fitted to the **same** simulated data.
Starting vectors rotate among the three choices above. The fit uses `nlminb()`
with up to three additional restarts. A usable fit has convergence code zero,
a finite maximum gradient below `0.001`, and a positive-definite numerical
Hessian in log-parameter space. The likelihood, estimates, standard errors, and
interval endpoints must also be finite, with positive estimates, standard errors,
and interval endpoints.

Approximate 95% Wald intervals are constructed on the log scale using the inverse
Hessian, then exponentiated. They are conditional on the fixed biological and
recruitment settings. They omit uncertainty in those settings and are not a
substitute for a larger assessment uncertainty analysis.

```{r noisy-summary}
knitr::kable(result$summary, digits = 3)
result$failures
```

`relative_bias` is mean estimated-to-true ratio minus one; `relative_rmse` is the
root mean squared relative error. `coverage` is the fraction of usable intervals
containing the truth. `coverage_mcse` is the binomial Monte Carlo standard error
of that fraction. With just 30 replicates, even a method with 95% coverage has
substantial sampling variation: near 95%, its coverage Monte Carlo SE is about
four percentage points.

Summaries use usable fits only and report `attempted`, `usable`, `failed`
(exceptions), and `unusable` (returned fits that fail diagnostics). The last three
counts sum to the number attempted for every parameter and schedule. When no
fits are usable, bias, RMSE, coverage, and coverage Monte Carlo SE are `NA`. Exceptions are retained in `failures`; nonzero convergence
codes, large gradients, or unsuitable Hessians remain in `draws` with
`usable = FALSE`. Hessian calculation or inversion errors retain the fitted
estimates and optimizer diagnostics, with unavailable intervals and a reason in
`uncertainty_message`; optimizer messages are retained in `message`. Do not interpret conditional coverage without checking these
failures. No test forces noisy bias or coverage into a preselected range.

```{r recovery-boxplot, fig.width=8, fig.height=6, fig.alt="Relative estimation errors for fishing mortality and catchability under phased and January-only recruitment; horizontal zero lines mark the generating values."}
local({
  old <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
  on.exit(par(old))
  for (parameter in names(result$truth)) {
    d <- result$draws[result$draws$usable & result$draws$parameter == parameter, ]
    if (!nrow(d)) {
      plot.new()
      title(main = paste(parameter, "(no usable fits)"))
      next
    }
    boxplot((estimate / truth - 1) ~ model, data = d,
            main = parameter, ylab = "Relative error", xlab = "Schedule",
            col = c("grey85", "lightblue"))
    abline(h = 0, lty = 2)
  }
})
```

## Interpretation and limits

Exact-data recovery shows that this model can recover the four estimated
parameters from its own predictions. Noisy replicates show sampling variation
and allow examination of bias and interval coverage under the chosen design.
The January-only fit illustrates how an incorrect recruitment schedule can be
absorbed by fishing mortality and catchability.

This exercise uses FIMS for both generation and fitting, so it cannot independently
verify every population-dynamics equation. Analytic survival/catch tests and
independent likelihood checks remain necessary. Nor does it establish that
recruitment timing itself is identifiable: dates and fractions were held fixed.
Further studies can vary sample sizes, observation SDs, phase separation, and
survey season, or add recruitment-process variation explicitly.
