---
title: "Introducing the FIMS Logging System"
output: github_document
vignette: >
  %\VignetteIndexEntry{Introducing the FIMS Logging System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r fims, warning=FALSE, message=FALSE, include = FALSE}
# loads the FIMS native library
library(FIMS)

# clear memory
clear()
```

## Importance of Logging 

Logging in FIMS is important because it allows developers to understand what is happening within a model. It is especially useful when problems arise. The FIMS logging system has been designed to handle common issues, such as undefined modules, dimension issues, and software errors that may occur when a model is not properly defined. This logging system is accessible from both R and C++ and messages are provided using a JSON format.

## How FIMS Logging works

At run time, FIMS records log entries in C++ and exposes them to R as JSON strings. From R, you can inspect all entries (`get_log()`), or only warnings and errors (`get_log_warnings()`, `get_log_errors()`).

For implementation details (log schema, severity macros, write-on-exit behavior,
and signal handling), see the following resources:

* The source file: [inst/include/common/def.hpp](https://github.com/NOAA-FIMS/FIMS/blob/main/inst/include/common/def.hpp)
* The [documentation](https://noaa-fims.github.io/FIMS/doxygen/classfims_1_1FIMSLog.html) for the same header

Model assembly can be triggered directly with `native_create_model()`. In normal
use, `initialize_fims()` performs this step as part of constructing the model.
Here we initialize the bundled example data to obtain native log entries.

```{r example-one-line}
data_4_model <- FIMSFrame(data_big)
model <- setup_default_parameters(data_4_model) |>
  initialize_fims(data = data_4_model)
log_string <- get_log()
log_data_frame <- jsonlite::fromJSON(log_string)
head(log_data_frame)

clear()
```

These entries originate in the native model initialization code.
The `level`, `file`, `routine`, and `line` fields identify the severity and C++
location that produced each entry.

See the [introduction to FIMS vignette](fims-demo.html) for a log of a complete model run.

## Using the Logging System in R

The FIMS Logging System is also available from R, with a caveat! Logging from R gives less information than logging from C++. Entries from R use `R_env` for the file, `R_script_entry` for the routine, and -1 for the line instead of a C++ source location. Below are examples of adding log entries from R.

```{r r-logging-info, eval=TRUE}
log_info("info entry from R script")
log_warning("warning entry from R script")
jsonlite::fromJSON(get_log())

clear()
```

You can also record errors. Use detailed messages to identify their cause; entries from R identify the routine as `R_script_entry`.

```{r r-logging-error, eval=FALSE}
error <- log_error("error entry from R script")
jsonlite::fromJSON(get_log_errors())
```

The logging functions are R wrappers around the native logger. Use `?FIMS::log_error` to read their documentation.

```{r}
ls("package:FIMS") |>
  grep(pattern = "_log|log_", value = TRUE) |>
  cli::cli_bullets()

# Inspect the R wrapper
methods::show(log_error)
```

Use `set_log_throw_on_error(TRUE)` to raise an R error immediately after recording an error message. This stops the current expression; it does not terminate the R session. `write_log(TRUE)` enables writing the log when the session exits.

```{r throw, eval = FALSE}
set_log_throw_on_error(TRUE)
log_error("throwing now!")
```
