---
title: "Annual catch with dated survey observations"
output: github_document
vignette: >
  %\VignetteIndexEntry{Annual catch with dated survey observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

FIMS keeps annual population transitions and annual catch while predicting
surveys at their observation dates. Dates use `YYYY-MM-DD`; a year alone means
January 1, and `YYYY-MM` means the first day of that month. Mortality rates remain
annual. Elapsed time is calendar days divided by 365 or 366.

This example uses three years of the built-in data, adds July surveys, and uses
parametric growth at fractional age. To keep the fit small, it estimates only
survey catchability while holding the other parameters fixed.

```{r dated-data, message=FALSE}
library(FIMS)
library(dplyr)

observations <- data_big |>
  filter(is.na(timing) | timing <= 3 | (type == "weight_at_age" & timing == 4))
observations$timing <- ifelse(is.na(observations$timing), NA_character_,
  sprintf("%04d", observations$timing + 2026L))

july <- observations |>
  filter(fleet == "survey1", type %in% c("index", "age_comp", "length_comp"))
july$timing <- paste0(july$timing, "-07-01")
frame <- FIMSFrame(bind_rows(observations, july))

get_observations(frame) |>
  filter(fleet == "survey1", type == "index") |>
  select(date, year_i, sample_i, year_fraction)
```

These extra samples reuse the example observations to demonstrate the interface;
they do not represent additional measured data. In an assessment, supply each
survey's actual observed values and uncertainty. Index, age-composition, and
length-composition dates may differ.

```{r fit, message=FALSE}
parameters <- setup_default_parameters(frame) |>
  filter(module_name != "Growth") |>
  bind_rows(setup_default_Growth(frame, module_type = "VonBertalanffySchnute"))
parameters$estimation_type <- "constant"
parameters$distribution[parameters$module_name == "Recruitment"] <- NA_character_
parameters$distribution_type[parameters$module_name == "Recruitment"] <- NA_character_
parameters$estimation_type[parameters$module_name == "Fleet" &
  parameters$fleet == "survey1" & parameters$label == "log_q"] <- "fixed_effects"

fit <- initialize_fims(parameters, frame) |>
  fit_fims(number_of_loops = 1)
get_max_gradient(fit)
```

`estimated` holds the reported prediction on its natural scale. For likelihood
rows, `expected` retains the likelihood input: for example, a log mean for a
lognormal index. `uncertainty` is the standard error of the reported estimate.
`augment()` pairs observations with predictions on the same scale and excludes
missing (`-999`) observations.

```{r predictions}
get_estimates(fit) |>
  filter(fleet == "survey1", label == "index_expected") |>
  select(observation_id, date, estimated, uncertainty, weight_timing)

generics::augment(fit) |>
  filter(fleet == "survey1", label == "index_expected") |>
  select(date, .truth, .pred)
```

Parametric growth uses `age + year_fraction`, including the representative age
of the oldest plus group. Age-class labels stay fixed within the year. Empirical
annual weights and fixed age-to-length matrices use annual lookup instead.
Maturity at observation dates is diagnostic; spawning and recruitment retain
their annual calculations.

```{r shared-products}
head(get_report(fit)$observation_biology)
get_report(fit)$observation_work
```

The work counters give distinct non-January-1 dates, biological size rows built,
and observation requests. Shared survey dates reuse biological products, while
each fleet retains its own length-bin mapping. Size distributions are built
only for dates with length-composition requests.

A retrospective peel removes terminal survey/composition years but preserves
annual catch, biological inputs, bin configuration, and the annual model horizon.
Missing survey years receive placeholder predictions, which diagnostics exclude.

```{r peel, message=FALSE}
peeled <- run_modified_data_fims(1, frame, parameters)
generics::augment(peeled) |>
  filter(fleet == "survey1", label == "index_expected") |>
  select(date, .truth, .pred)
clear()
```

For projections, extend annual catch or fishing-mortality inputs and the required
biological tables as in [Simple projections using FIMS](fims-demo-projections.html).
Future survey predictions may have dates and missing observed values. Adding
those prediction dates does not change projected population states, catch, or
spawning biomass when parameters and annual inputs stay fixed.
