---
title: "Minimal Example for Running a FIMS Model"
output: github_document
vignette: >
  %\VignetteIndexEntry{Minimal Example for Running a FIMS Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This vignette provides the minimal steps needed to run a Fisheries Integrated Modeling System (FIMS) model.

## Setup

Load the FIMS package and clear memory:
```{r memory, warning=FALSE, message=FALSE}
library(FIMS)

# clear memory
clear()
```

## Data and parameters

First, load the available data and create a FIMS data frame to create a model with. Then create a set of default parameters.
```{r data}
# Load sample data
data("data1")
# Prepare data for FIMS model
data_4_model <- FIMSFrame(data1)

# Create parameters
parameters <- data_4_model |>
  create_default_configurations() |>
  create_default_parameters(data = data_4_model)
```

## Initialize and Fit

Initialize the FIMS modules and fit the model.
```{r fit, max.height='100px', attr.output='.numberLines', eval=TRUE}
# Run the  model with optimization
fit <- parameters |>
  initialize_fims(data = data_4_model) |>
  fit_fims(optimize = TRUE)

# Clear memory post-run
clear()
```
