---
title: "Running diagnostics on a FIMS model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running diagnostics on a FIMS model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 7,
  out.width = "100%"
)
```

```{r setup}
library(FIMS)
packageVersion("FIMS")
clear()
```

## Set up FIMS model

Following the [minimal FIMS demo vignette](https://noaa-fims.github.io/FIMS/articles/fims-demo-minimal.html), we will set up a FIMS model and run it. This model will contain the all years of data and will be our reference model (see section below for more information about reference models). We need to run it initially to get the MLE parameter estimates. For this example we are using the dataset included in the FIMS package and creating a set of default parameters.

```{r run_fims_model}
# Load sample data
data("data_big")
# Prepare data for FIMS model
data_4_model <- FIMSFrame(data_big)

# Create parameters
parameters <- setup_default_parameters(data = data_4_model)

# Run the  model with optimization
base_model <- parameters |>
  initialize_fims(data = data_4_model) |>
  fit_fims(optimize = TRUE)

# Clear memory post-run
clear()
```

## Likelihood profile

Once we have a model that is fit, we can run a likelihood profile to see the impact of each data type on a given parameter. In this example, we will test the impact on R~0~ but any value can be tested by changing the `parameter_name` argument. *Note: The parameter name must match what it is called in the `label` column of the `parameters` tibble.* An optional input is the `module_name` the parameter is associated with. However, as long as there is only one parameter with that name, `module_name` can be left NULL. `run_fims_likelihood()` function takes a fitted FIMS model, finds the MLE estimate for the parameter being profiled over (in this case R~0~), and profiles over a vector of values based on the `min`, `max`, and `length` user inputs. In the example below, it will profile over 5 values that range from the `MLE value - 1` to the `MLE value + 1`. If you are running many models at once, it can be computationally more efficient to use multiple cores. To do this, change the `n_cores` argument to as many as you want/can use. *NOTE: If left empty, the default for `n_cores` is one less than the number of cores on your machine so this could unexpectedly use up a lot of your computing power.*

```{r like_profile}
#| message: false
#| warning: false

like_fit <- run_fims_likelihood(
  model = base_model,
  parameters = parameters,
  parameter_name = "log_rzero",
  data = data_big,
  n_cores = 3,
  min = -1,
  max = 1,
  length = 3
)
```

### Visualize results

Use `plot_likelihood()` to visualize the profile. The plot shows the change in total likelihood over the R~0~ values profiled in the solid black line and the change in likelihood for each data type in the color lines below it. This plot indicates that the R~0~ value estimated by the base model (13.9) is the value that leads to the lowest likelihood and all of the data types support this (i.e., the data do not conflict).  

```{r}
#| fig.alt: >
#|   Curved lines showing change in likelihood in total and for each data type.
fig_likelihood_profile <- plot_likelihood(like_fit)
print(fig_likelihood_profile)
```

This function creates a `ggplot` object that can be customized and added onto. The default theme is `stockplotr::theme_noaa`, however to change the theme or colors used, you can simply add them after the main plot function.

```{r}
#| fig.alt: >
#|   The same curved-line plot as above but with a different ggplot theme.
fig_likelihood_custom_theme <- plot_likelihood(like_fit) +
  ggplot2::theme_bw()
print(fig_likelihood_custom_theme)
```

### Clean up

Once you are finished running a FIMS model, it is always good practice to clear the C++ memory from one FIMS model run to the next.
```{r}
clear()
```

## Retrospective analysis

A retrospective analysis is a common model diagnostic used to assess the stability of estimates of parameters and derived quantities, such as fishing mortality and spawning biomass, and if there is a consistent pattern in the estimates as years of data are removed. The process involves removing *n* years of data from the end of the time series and refitting the model. This is an iterative process where the user typically runs 5--10 models, with each run (often called "peels") progressively removing another year of data. Estimates from these peels are then compared to the model run with the full dataset (reference model). The Mohn's rho statistic can be used to compare the relative difference between the reference model and each peel. Mohn's rho statistic is calculated as: 

$$\rho = \frac{1}{n} \sum_{i=1}^{n} \frac{x_{p,y} - x_{base,y}}{x_{base,y}}$$

where $n$ is the number of retrospective peels, $x_{p, y}$ is the estimated value (e.g., spawning biomass, fishing mortality, etc.) from peel $p$ at its terminal year $y$, and $x_{base,y}$ is the value from the reference model, $base$, at that same year $y$.

Some things to note about `run_fims_retrospective()` is that you can provide data that is already a `FIMSFrame` object or you can give it one that is not. In the example below, we are using the `FIMSFrame` data object (`data_4_model`) created above. Additionally, it expects a vector of positive values of `years_to_remove` from the base model, starting from 0. If you are running many peels, it may be more efficient to use multiple cores, which can be specified with the `n_cores` argument. 

> NOTE: If left empty, the default for `n_cores` is one less than the number of cores on your machine so this could unexpectedly use up a lot of your computing power. 

```{r, retro}
retro_fit <- run_fims_retrospective(
  years_to_remove = 0:2,
  data_4_model,
  parameters,
  n_cores = 1
)
```

### Mohn's Rho

Once you have run a retrospective, you can then calculate the Mohn's rho statistic for spawning biomass by: 

```{r, mohns_rho}
rho_sb <- calculate_mohns_rho(retro_fit, quantity = "spawning_biomass")
```

### Visualize the results

To visualize the results of the retrospective analysis, we can use `plot_retrospective()` to display the quantity of interest. Below, we want to assess the impact on spawning biomass. The function returns a `ggplot` object, which you can customize by adding add layers as you normally would to a `ggplot`. For example, below, we specify the x and y axes labels and the legend title.

```{r, plot-spawning-biomass, message = FALSE, warning = FALSE}
#| fig.alt: >
#|   Plot of spawning biomass showing a small negative bias with removed years.
fig_retrospective_ssb <- stockplotr::plot_spawning_biomass(
  retro_fit$estimates |>
    dplyr::rename(
      year = year_i,
      estimate = estimated
    ) |>
    dplyr::mutate(
      uncertainty_label = "se"
    ),
  group = "retrospective_peel"
)
print(fig_retrospective_ssb)
```

We can compare the spawning biomass estimates from the reference model and each peel for the last six years. 
```{r}
retro_fit[["estimates"]] |>
  dplyr::filter(label == "spawning_biomass") |>
  dplyr::select(label, year_i, estimated, retrospective_peel) |>
  tidyr::pivot_wider(
    names_from = retrospective_peel,
    values_from = estimated
  ) |>
  dplyr::rename(
    "Year" = year_i,
    "Reference model" = `0`,
    "Peel 1" = `1`,
    "Peel 2" = `2`
  ) |>
  tail()
```

### Clean up

Once you are finished running a FIMS model, it is always good practice to clear the C++ memory from one FIMS model run to the next.
```{r}
clear()
```
