---
title: "Input data and the FIMSFrame object"
output: github_document
vignette: >
  %\VignetteIndexEntry{Input data and the FIMSFrame object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set local options so vignette output is clean and consistent
# (avoids color codes and subtle formatting in rendered docs)
withr::local_options(list(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.subtle = FALSE
))
```

## Other sources of information
This vignette is an expansion of the section on "Data" in the [FIMS overview vignette](https://noaa-fims.github.io/FIMS/articles/fims-demo.html). It provides more details about the data structure needed for a FIMS model and how to prepare your data for use in a FIMS model using the `FIMSFrame()` function.

The FIMS cheatsheet (https://github.com/NOAA-FIMS/FIMS/blob/dev/pkgdown/assets/fims-cheatsheet.pdf) includes a quick summary of the input data format.

The help page for the `FIMS::data_big` data set (https://noaa-fims.github.io/FIMS/reference/data_big.html or via `?FIMS::data_big`) provides detailed information about the input data format.

Note: this vignette does not include anything about model configuration, which is covered in the overview vignette linked above.

## Data format

The input data for a FIMS model is a single long data frame with the following columns: `type`, `name`, `age`, `length`, `timing`, `value`, and `unit`. Both the single table and long format are a contrast to the input format for some legacy stock assessment models where the data are provided in multiple tables, each with a different format. The SS3 data file, for instance, has separate tables for catch, indices, age compositions, length compositions, with wide tables for the composition data (each age or length bin has a different column). The use of a single, long table for data input used by FIMS makes it easier to summarize and filter the data across data types (e.g. by fleet, time step, age, etc.), as well as being better suited to the widely-used tidyverse collection of packages in R.

Two drawbacks of the long format are (1) some information is duplicated, and (2) the long format is harder to read and understand by looking at the raw data frame. To meet this second challenge, the FIMS team is working on functions to summarize, visualize, and check for errors in the input data.

The currently available values for the `type` column are "landings", "index", "age_comp", "length_comp", "weight_at_age", and "age_to_length_conversion". The "weight_at_age" and "age_to_length_conversion" types are not currently treated as data in the sense that they are not included in the likelihood, just used within the population dynamics calculations. However, the input data is currently the most streamlined way to provide this information to the model.

The individual data types are described in more detail below after introducing the example data set that is included in the package.

## `data_big`

A sample data frame for a catch-at-age model with both ages and lengths is stored in the package as `data_big`. This data set is based on data that was used in [Li et al.](https://www.doi.org/10.7755/FB.119.2-3.5) for the Model Comparison Project ([github site](https://github.com/NOAA-FIMS/Age_Structured_Stock_Assessment_Model_Comparison)). The length data have since been added [data-raw/data_big.R](https://github.com/NOAA-FIMS/FIMS/blob/main/data-raw/data_big.R) based on an age-length conversion matrix. To see how this example data frame was created, see the R script here [R/data_big.R](https://github.com/NOAA-FIMS/FIMS/blob/main/R/data_big.R).

Here we explore each of the data types in turn.

### Landings (`type == "landings"`)

The landings data type contains the total catch in weight for each time step. FIMS does not yet include the option to model discards internally, so "landings" should account for all catch. The landings are not associated with an age or length so those columns should only contain `NA` values. The `unit` column should always be `mt` because this value is not yet processed so no conversion would be applied for other units. The `uncertainty` column should be the standard deviation of the logged observations if you are using the lognormal distribution to fit your data. `log_Fmort` parameters for each fleet and time step with catch can be estimated to fit the model to the landings data, informed by the uncertainty units. Common practice is to set the uncertainty to a small number, like 0.001.

```{r data_big-landings}
FIMS::data_big |>
  dplyr::filter(type == "landings") |>
  head(3)
```

```{r data_big-landings-plot}
#| fig.alt: >
#|   Barplot shows an increase in landings over the timing values 1 to 8, then noisy but stable values until the final bar at timing equals 30.

library(ggplot2)
FIMS::data_big |>
  dplyr::filter(type == "landings") |>
  ggplot(aes(x = timing, y = value)) +
  geom_bar(stat = "identity")
```

### Indices (`type == "index"`)
The index data type contains the relative or absolute abundance index for each time step. The inputs are similar to landings, in having `NA` values for age and length, `mt` for `unit`, and index uncertainty as the standard deviation of the logged observations. A `log_q` parameter can be fixed or estimated for each index.

```{r data_big-indices}
FIMS::data_big |>
  dplyr::filter(type == "index") |>
  head(3)
```

```{r data_big-indices-plot}
#| fig.alt: >
#|   Points with uncertainty shows a steadily decreasing index over time steps 1 to 30 with increasing precision for the later observations.
FIMS::data_big |>
  dplyr::filter(type == "index") |>
  ggplot(aes(x = timing, y = value)) +
  geom_point() +
  geom_pointrange(aes(
    ymin = qlnorm(p = 0.025, meanlog = log(value), sdlog = uncertainty),
    ymax = qlnorm(p = 0.975, meanlog = log(value), sdlog = uncertainty)
  )) +
  geom_hline(yintercept = 0)
```

### Age compositions (`type == "age_comp"`)
The age composition data type contains the age distribution of a fleet for a time step. The `age` column contains the age, which needs to start at age 0, and the `value` column contains either the number or the proportion of the catch or survey in each age bin. The `unit` column should be either "number" or "proportion". The `uncertainty` column should be the input sample size for the time step (equal across all ages within that time step).

```{r data_big-age_comp}
FIMS::data_big |>
  dplyr::filter(type == "age_comp") |>
  head(3)
```
```{r data_big-age_comp-plot}
#| fig.alt: >
#|   Barplot showing age comp by time step with a larger proportion of older ages (including a significant plus group at age 12 for the early time steps), and a truncated age distribution in the final time steps with relatively few fish over age 7.

FIMS::data_big |>
  dplyr::filter(type == "age_comp") |>
  ggplot2::ggplot(
    ggplot2::aes(x = age, y = value)
  ) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::xlab("Age") +
  ggplot2::ylab("Number") +
  ggplot2::facet_wrap(~timing) + 
  ggplot2::scale_x_continuous(breaks = 0:12)
```

### Length compositions (`type == "length_comp"`)
The length composition data are similar to the age compositions, except with values in the `length` column and `age = NA`. FIMS does not yet have growth implemented, so the length bins can be any values as long as the distribution of each age among those bins is provided by the `age_to_length_conversion` data type described below.

```{r data_big-length_comp}
FIMS::data_big |>
  dplyr::filter(type == "length_comp") |>
  head(3)
```
```{r data_big-length_comp-plot}
#| fig.alt: >
#|   Barplot showing length comp by time step with a mode around 450 mm which declines and becomes more sharply peaked over time.

FIMS::data_big |>
  dplyr::filter(type == "length_comp") |>
  ggplot2::ggplot(
    ggplot2::aes(x = length, y = value)
  ) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::xlab("Length") +
  ggplot2::ylab("Number") +
  ggplot2::facet_wrap(~timing)
```

### Weight-at-age data (`type == "weight_at_age"`)

### Age-to-length conversion (`type == "age_to_length_conversion`)


