---
title: "Recruitment phases and April surveys with direct Rcpp interfaces"
output: github_document
vignette: >
  %\VignetteIndexEntry{Recruitment phases and April surveys with direct Rcpp interfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
```

This walkthrough builds and fits a small FIMS catch-at-age model directly through
its exposed C++ modules and TMB. It combines two recruitment phases, annual
fishery observations, and April survey observations. Every data object and
likelihood connection is explicit; it does not use `FIMSFrame()`,
`setup_default_parameters()`, `initialize_fims()`, or `fit_fims()`.

For the higher-level interface, start with
[Multiple recruitment phases](fims-recruitment-phases.html). A standalone version
of this example is installed at
`system.file("examples", "recruitment-phases-without-wrappers.R", package = "FIMS")`.

## Start a fresh model

Use an installed FIMS version that includes recruitment phases, together with
Rcpp and TMB. Run the chunks in order in a fresh R session. The `FIMS::` prefix
resolves each exposed C++ class explicitly.

FIMS registers module objects globally. `clear()` resets that registry, so do
not call it while you still need to evaluate an existing TMB objective. Here
we clear it before constructing any modules.

```{r direct-setup}
library(FIMS)
# Use explicit namespace references so class lookup does not depend on the
# search path. These are the exposed C++ classes, not model setup wrappers.
FIMS::clear() # Resets the global FIMS registry; use a fresh R session for this example.

years <- 2027:2029
ages <- 1:5
ny <- length(years)
na <- length(ages)
```

## Annual fishery and April survey data

We model 2027–2029 with five annual age classes; the last class is a plus group.
The numbers below are hand-selected teaching data, not observations from an
assessment or a simulated parameter-recovery experiment.

| Data stream | Timing | Observation model | Sample size or uncertainty |
|:--|:--|:--|:--|
| Fishery biomass catch | Total over each year | Lognormal | Log-scale SD 0.20 |
| Fishery age composition | Catch over each year | Multinomial | 200 fish per year |
| Survey biomass index | April 1 | Lognormal | Log-scale SD 0.25 |
| Survey age composition | April 1 | Multinomial | 150 fish per survey |

Composition matrices have samples in rows and ages in columns. Their entries
are counts: row sums supply multinomial sample sizes. If starting from
proportions, multiply by the intended sample size before creating these data
objects.

Despite the table column name `log_sd`, its values are **SDs on the log-observation
scale**. The distribution module's `log_sd` parameter stores the logarithm of
that SD. If uncertainty is supplied as a coefficient of variation, first use
`sqrt(log(1 + CV^2))` to obtain the log-scale SD.

Catch and growth-derived weights must use consistent mass units. Survey
catchability converts vulnerable biomass into the relative index's units.

```{r direct-data}
# Illustrative observations: annual fishery totals and April 1 survey samples.
# Biomass uses the same units as growth-derived weight; the survey index is
# relative biomass (q scales it). Composition entries are COUNTS, not proportions.
annual_data <- data.frame(year = years, catch = c(500, 550, 600), log_sd = 0.2)
annual_age_counts <- rbind(
  c(10, 25, 45, 50, 70),
  c(12, 28, 43, 52, 65),
  c(15, 30, 45, 50, 60)
)
survey_data <- data.frame(
  year = years, date = as.Date(paste0(years, "-04-01")),
  index = c(220, 230, 240), log_sd = 0.25
)
survey_age_counts <- rbind(
  c(10, 25, 30, 35, 50),
  c(12, 28, 32, 35, 43),
  c(15, 30, 35, 32, 38)
)
colnames(annual_age_counts) <- colnames(survey_age_counts) <- ages
rownames(annual_age_counts) <- years
rownames(survey_age_counts) <- as.character(survey_data$date)
stopifnot(nrow(annual_age_counts) == ny, nrow(survey_age_counts) == ny,
          all(annual_age_counts >= 0), all(survey_age_counts >= 0))
```

## Split recruitment across January and July

Each year's recruitment budget is split 40% on January 1 and 60% on July 1.
Both groups enter at biological age 1 **on their own entry date**. On April 1,
that year's January recruits are present and have aged since entry; that year's
July recruits are not yet present. Surviving July recruits from earlier years
can still contribute to the April survey.

The Beverton–Holt module supplies the annual mean, and `LogDevsRecruitmentProcess`
applies annual deviations. Here the deviations are fixed at zero. Fractions
allocate the resulting annual total; they do not define separate recruitment
processes. The transformation below maps steepness from `(0.2, 1)` onto the
module's unconstrained scale.

`SetRecruitmentSchedule()` takes zero-based model-year indices, within-year
fractions, allocation fractions, and phase identifiers. Calculate fractions
from actual dates: 2028 has 366 days. Supply a complete, chronologically ordered
schedule with fractions summing to one in each year and consistent phase IDs.
No schedule row is supplied for the extra terminal reporting year.

```{r direct-recruitment}
# One annual recruitment total, allocated 40% in January and 60% in July.
# C++ year coordinates are ZERO-based. Fractions use actual calendar year length
# (2028 is a leap year); do not approximate July 1 as exactly half a year.
dates <- as.Date(paste0(rep(years, each = 2), rep(c("-01-01", "-07-01"), ny)))
year_start <- as.Date(paste0(rep(years, each = 2), "-01-01"))
year_end <- as.Date(paste0(rep(years + 1L, each = 2), "-01-01"))
phase <- rep(c("early", "late"), ny)
year_fraction <- as.numeric(dates - year_start) / as.numeric(year_end - year_start)

recruitment <- methods::new(FIMS::BevertonHoltRecruitment)
process <- methods::new(FIMS::LogDevsRecruitmentProcess)
recruitment$SetRecruitmentProcessID(process$get_id())
recruitment$n_years$set(ny)
recruitment$log_rzero[1]$value <- log(1e6)
h <- 0.75
recruitment$logit_steep[1]$value <- log((h - 0.2) / (1 - h))
recruitment$log_devs[] <- rep(0, ny - 1L)
recruitment$SetRecruitmentSchedule(
  rep(seq_len(ny) - 1L, each = 2), year_fraction,
  rep(c(0.4, 0.6), ny), phase
)
recruitment$SetRecruitmentEntryAges(rep(1, length(dates)))
# Both enter at biological age 1 on their own date. The July cohort is younger
# than the January cohort when observed together. All parameters default to
# constant unless explicitly marked below.
```

## Define shared biology and fishery selectivity

Growth and maturity parameters are shared across phases. The model evaluates
them at each phase cohort's biological age. These illustrative growth settings
use length in millimetres and weight in metric tonnes; the length–weight
coefficient supplies that conversion. The two length SDs describe variability
at the growth reference ages. This example has no length-composition data, so
it does not activate a fleet age-to-length mapping.

Mortality and selectivity remain annual age-class quantities. Logistic maturity
and fishery selectivity have separate inflection points and slopes. Parameters
remain constant unless their `estimation_type` is explicitly changed.

```{r direct-biology}
growth <- methods::new(FIMS::VonBertalanffySchnuteGrowth)
growth$n_ages$set(na)
growth$mean_length_young[1]$value <- 275
growth$mean_length_old[1]$value <- 725
growth$reference_age_for_length_young[1]$value <- min(ages)
growth$reference_age_for_length_old[1]$value <- max(ages)
growth$growth_coefficient[1]$value <- 0.18
growth$length_weight_a[1]$value <- 2.5e-11
growth$length_weight_b[1]$value <- 3
growth$length_at_age_sd_at_reference_ages[] <- c(28, 73)

maturity <- methods::new(FIMS::LogisticMaturity)
maturity$inflection_point[1]$value <- 3
maturity$slope[1]$value <- 1
selectivity <- methods::new(FIMS::LogisticSelectivity)
selectivity$inflection_point[1]$value <- 2.5
selectivity$slope[1]$value <- 1
```

## Connect annual fishery data and likelihoods

Creating a data object alone does not add a likelihood. Each stream needs three
connections: the fleet points to observed data, the distribution points to the
same data, and the distribution links to the fleet's expected-value vector.

For lognormal catch, that vector is `log_catch_expected`. For multinomial ages,
it is `agecomp_proportion`. `c(t(counts))` flattens a sample-by-age R matrix in
the row order expected by C++.

There is one estimated log fishing-mortality parameter per year. Annual catch
includes each cohort only after recruitment. Catch numbers use its remaining
exposure time; biomass catch integrates its changing biological weight. Fishery
age proportions describe this annual catch, not a January snapshot.

```{r direct-fishery}
# Annual fishing fleet: lognormal biomass catch and multinomial age counts.
catch_data <- methods::new(FIMS::Catch, ny)
catch_data$catch_data[] <- annual_data$catch
annual_age_data <- methods::new(FIMS::AgeComp, ny, na)
# Flatten sample-by-age matrices in row-major order for C++.
annual_age_data$age_comp_data[] <- c(t(annual_age_counts))
fleet <- methods::new(FIMS::Fleet)
fleet$n_years$set(ny)
fleet$n_ages$set(na)
fleet$log_Fmort[] <- rep(log(0.2), ny)
fleet$log_Fmort$set_estimation_types("fixed_effects")
fleet$log_q[1]$value <- 0
fleet$SetSelectivityID(selectivity$get_id())
fleet$SetObservedCatchDataID(catch_data$get_id())
fleet$SetObservedAgeCompDataID(annual_age_data$get_id())
catch_likelihood <- methods::new(FIMS::DlnormDistribution)
# log_sd stores log(SD on the log-observation scale), not log(CV).
catch_likelihood$log_sd[] <- log(annual_data$log_sd)
catch_likelihood$set_observed_data(catch_data$get_id())
catch_likelihood$set_distribution_links("data", fleet$log_catch_expected$get_id())

annual_age_likelihood <- methods::new(FIMS::DmultinomDistribution)
annual_age_likelihood$set_observed_data(annual_age_data$get_id())
annual_age_likelihood$set_distribution_links("data", fleet$agecomp_proportion$get_id())
```

## Connect the April survey and its likelihoods

A survey is represented by a second fleet with its own selectivity and one
estimated catchability parameter shared across years. Its fishing mortality is
fixed at `exp(-200)`, effectively zero. Its index is vulnerable biomass scaled
by catchability; its age composition describes the selected population at the
sample date.

Both survey streams need explicit timing. Otherwise low-level callers use the
legacy annual timing. The order of timing rows must match the corresponding
data rows. `SetObservationTiming()` takes the following coordinates:

| Argument | Meaning in this example |
|:--|:--|
| `type` | `"index"` or `"age_comp"` |
| `years` | Zero-based model years: 0, 1, 2 |
| `days` | Integer days since 1970-01-01, obtained from an R `Date` |
| `ids` | Zero-based sample indices within the stream |
| `observation_ids` | Distinct text identifiers for the samples |
| `dates` | ISO date strings |
| `fractions` | Elapsed days since January 1 divided by days in that year |

The model evaluates survival and phase-specific biology at those dates. An
observation on a recruitment date includes that day's newly recruited fish.
Adding or moving survey samples does not define the annual integration intervals
or change the population trajectory.

```{r direct-survey}
# April survey: a separate non-extractive fleet with its own selectivity and q.
# No annual integration for these samples: predictions are evaluated on April 1.
survey_index_data <- methods::new(FIMS::Index, ny)
survey_index_data$index_data[] <- survey_data$index
survey_age_data <- methods::new(FIMS::AgeComp, ny, na)
survey_age_data$age_comp_data[] <- c(t(survey_age_counts))
survey_selectivity <- methods::new(FIMS::LogisticSelectivity)
survey_selectivity$inflection_point[1]$value <- 2
survey_selectivity$slope[1]$value <- 1
survey <- methods::new(FIMS::Fleet)
survey$n_years$set(ny)
survey$n_ages$set(na)
survey$log_Fmort[] <- rep(-200, ny) # Fixed, effectively zero survey mortality.
survey$log_q[1]$value <- log(0.03)
survey$log_q[1]$estimation_type$set("fixed_effects")
survey$SetSelectivityID(survey_selectivity$get_id())
survey$SetObservedIndexDataID(survey_index_data$get_id())
survey$SetObservedAgeCompDataID(survey_age_data$get_id())

survey_start <- as.Date(paste0(years, "-01-01"))
survey_end <- as.Date(paste0(years + 1L, "-01-01"))
survey_fraction <- as.numeric(survey_data$date - survey_start) /
  as.numeric(survey_end - survey_start)
# days are absolute days since 1970-01-01, NOT day-of-year. Year and sample
# coordinates are zero-based. Each stream's rows must match its data row order.
for (stream in c("index", "age_comp")) {
  survey$SetObservationTiming(
    stream, seq_len(ny) - 1L, as.integer(survey_data$date), seq_len(ny) - 1L,
    paste0("survey1_", stream, "_", survey_data$date),
    as.character(survey_data$date), survey_fraction
  )
}
survey_index_likelihood <- methods::new(FIMS::DlnormDistribution)
survey_index_likelihood$log_sd[] <- log(survey_data$log_sd)
survey_index_likelihood$set_observed_data(survey_index_data$get_id())
survey_index_likelihood$set_distribution_links("data", survey$log_index_expected$get_id())
survey_age_likelihood <- methods::new(FIMS::DmultinomDistribution)
survey_age_likelihood$set_observed_data(survey_age_data$get_id())
survey_age_likelihood$set_distribution_links("data", survey$agecomp_proportion$get_id())
# Recruitment deviations remain fixed at zero, so no recruitment-process
# likelihood is needed here. Estimating random deviations requires a process
# distribution as well as registering those deviations as random effects.
```

## Assemble the population and model

Modules connect through the IDs returned by `get_id()`. The population links to
recruitment, growth, maturity, and both fleets; the catch-at-age model links to
the population. Keep the R module objects available while constructing the model.

Natural mortality is fixed at 0.2 for every year and age. The first element of
`log_init_naa` represents the **entire first-year recruitment budget**, not the
number already present on January 1. It therefore supplies 400,000 January
recruits and 600,000 July recruits. The older elements specify initial annual
age-class totals, which the model distributes among phases using reference
unfished survivorship.

Only the three annual fishing mortalities and survey catchability are estimated.
Initial abundance, recruitment parameters, biological parameters, selectivity,
observation SDs, and phase fractions remain fixed. This limited fit illustrates
the connections; it is not evidence that all these quantities can be estimated
jointly from three years of data.

```{r direct-population}
population <- methods::new(FIMS::Population)
population$n_years$set(ny)
population$n_ages$set(na)
population$n_fleets$set(2L)
population$ages[] <- ages
population$log_M[] <- rep(log(0.2), ny * na)
population$log_init_naa[] <- log(c(1e6, 8e5, 6e5, 4e5, 5e5))
# The first element is the FULL first-year recruitment budget: 400,000 early
# and 600,000 late. Older initial age totals are split internally among phases.
population$SetRecruitmentID(recruitment$get_id())
population$SetGrowthID(growth$get_id())
population$SetMaturityID(maturity$get_id())
population$AddFleet(fleet$get_id())
population$AddFleet(survey$get_id())
model <- methods::new(FIMS::CatchAtAge)
model$AddPopulation(population$get_id())
```

## Build the TMB objective and fit

`CreateTMBModel()` copies the registered interfaces into the numerical model.
`get_fixed()` and `get_random()` return its estimable parameter vectors. TMB
receives an empty data list because observations have already been registered
through the FIMS data modules.

There are no estimated random effects in this example, and therefore no
recruitment-process likelihood term. Adding random recruitment deviations would
also require their process distribution and `random = "re"` in `MakeADFun()`;
changing the estimation type alone is insufficient.

We use `nlminb()` directly. Its convergence code, finite objective, and maximum
absolute gradient are useful diagnostics. A zero convergence code alone does
not guarantee an adequate fit. Set `optimize` to `FALSE` to inspect predictions
at the starting values.

Reports below are evaluated explicitly at `parameters`. They do not depend on
copying optimized values back into the original Rcpp interface objects.

```{r direct-fit}
FIMS::CreateTMBModel()
obj <- TMB::MakeADFun(
  data = list(), parameters = list(p = FIMS::get_fixed(), re = FIMS::get_random()),
  DLL = "FIMS", silent = TRUE
)

# Fit annual fishing mortality and survey catchability using nlminb directly.
# Set FALSE for a forward evaluation at the initial parameter values.
optimize <- TRUE
parameters <- obj$par
if (optimize) {
  opt <- nlminb(parameters, obj$fn, obj$gr)
  stopifnot(opt$convergence == 0L)
  parameters <- opt$par
}
max_gc <- max(abs(obj$gr(parameters)))
nll <- obj$fn(parameters)
report <- obj$report(parameters)
stopifnot(is.finite(nll), all(is.finite(obj$gr(parameters))))
```

## Inspect phase allocations and predictions

Unlike the zero-based schedule inputs, raw report indices are one-based.
Here fleet 1 is the fishery and fleet 2 is the survey, following their creation
order. Phase indices follow the sorted identifiers (`early`, then `late`).

`recruitment_events` reports phase allocations. `recruitment_cohorts` gives
biological age and abundance at each row's `available_from` time: January 1 for
survivors or the entry time for new recruits. It is **not** an April survey
snapshot. Use the survey prediction vectors to inspect the April observations.

The terminal year reports only its January allocation of mean recruitment using
the final modeled year's phase pattern. The plus group retains a representative
biological age per phase rather than all individual ages. See
[Multiple recruitment phases](fims-recruitment-phases.html) for the initialization,
unfished-reference, and catch-quadrature conventions.

```{r direct-reports}
# Report coordinates are ONE-based (unlike schedule inputs); columns are unnamed.
events <- as.data.frame(report$recruitment_events)
names(events) <- c(
  "year_i", "phase_i", "year_fraction", "entry_age",
  "recruits", "unfished_recruits"
)
events$year <- c(years, max(years) + 1L)[events$year_i]
events$phase <- c("early", "late")[events$phase_i]
cohorts <- as.data.frame(report$recruitment_cohorts)
names(cohorts) <- c(
  "year_i", "phase_i", "age_i", "biological_age",
  "available_from", "numbers", "unfished_numbers"
)
# The extra terminal year reports January only, using the final year's phase
# pattern and mean recruitment; later terminal-year pulses are not inserted.
# Cohort ages/numbers describe their available_from date, not a shared survey
# date. The plus group uses a representative biological age for each phase.
cat("\nRecruitment events:\n")
print(events)
cat("\nInitial-year cohorts:\n")
print(subset(cohorts, year_i == 1))
cat("\nAnnual catch predictions:\n")
print(data.frame(annual_data, predicted = as.numeric(report$catch_expected[[1]])))
cat("\nApril survey index predictions:\n")
survey_predictions <- data.frame(survey_data,
  predicted = as.numeric(report$index_expected[[2]]))
print(survey_predictions)
annual_age_predictions <- matrix(report$agecomp_proportion[[1]], ny, na, byrow = TRUE)
survey_age_predictions <- matrix(report$agecomp_proportion[[2]], ny, na, byrow = TRUE)
cat("\nApril survey age probabilities (rows are dates):\n")
rownames(survey_age_predictions) <- as.character(survey_data$date)
colnames(survey_age_predictions) <- ages
print(survey_age_predictions)
```

The panels below compare observations (points) with fitted predictions (lines).
Annual catch and the April index have different units and temporal meanings, so
they are plotted separately. The model need not pass through every point: one
shared survey catchability and the age-composition likelihoods constrain the fit.

```{r direct-prediction-plots, fig.width=8, fig.height=3.5, fig.alt="Annual fishery catch and April survey index observations compared with fitted predictions in separate panels."}
local({
  old <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
  on.exit(par(old))
  plot(years, annual_data$catch, pch = 16, xlab = "Year",
       ylab = "Annual catch (tonnes)", main = "Annual fishery",
       ylim = range(c(annual_data$catch, report$catch_expected[[1]])))
  lines(years, report$catch_expected[[1]], col = "steelblue", lwd = 2)
  legend("topleft", c("Observed", "Predicted"), pch = c(16, NA),
         lty = c(NA, 1), col = c("black", "steelblue"), bty = "n")
  plot(years, survey_data$index, pch = 16, xlab = "Year",
       ylab = "Relative biomass index", main = "April 1 survey",
       ylim = range(c(survey_data$index, survey_predictions$predicted)))
  lines(years, survey_predictions$predicted, col = "steelblue", lwd = 2)
})
```

## Verify the likelihood components

The total negative log likelihood is the sum of four observation components:

\[
\mathrm{NLL} = \mathrm{NLL}_{\mathrm{catch}} +
\mathrm{NLL}_{\mathrm{fishery\ ages}} +
\mathrm{NLL}_{\mathrm{survey\ index}} +
\mathrm{NLL}_{\mathrm{survey\ ages}}.
\]

For catch and index, FIMS uses `meanlog = log(prediction)`, including the
lognormal density's Jacobian term. Thus the prediction is the lognormal median;
there is no additional mean-bias adjustment in this example. For age counts,
multinomial probabilities are normalized within each sample.

The following calculations independently reconstruct the four components with
base R distributions. Their sum must agree with TMB's objective, including
normalizing constants. This checks that all four streams are connected to the
intended prediction vectors. It does not assess model adequacy or parameter
identifiability.

```{r direct-likelihood-check}
# Independent R calculations make each likelihood link inspectable. FIMS uses
# log(prediction) as meanlog, with no lognormal mean-bias adjustment.
nll_by_component <- c(
  annual_catch = -sum(dlnorm(annual_data$catch,
    log(report$catch_expected[[1]]), annual_data$log_sd, log = TRUE)),
  annual_age = -sum(vapply(seq_len(ny), function(i)
    dmultinom(annual_age_counts[i, ], prob = annual_age_predictions[i, ], log = TRUE), numeric(1))),
  april_index = -sum(dlnorm(survey_data$index,
    log(survey_predictions$predicted), survey_data$log_sd, log = TRUE)),
  april_age = -sum(vapply(seq_len(ny), function(i)
    dmultinom(survey_age_counts[i, ], prob = survey_age_predictions[i, ], log = TRUE), numeric(1)))
)
stopifnot(isTRUE(all.equal(sum(nll_by_component), nll, tolerance = 1e-7)),
          all(abs(rowSums(survey_age_predictions) - 1) < 1e-8))
cat("\nNegative log likelihood by component:\n")
print(nll_by_component)
cat("\nNegative log likelihood:", nll, "\n")
cat("Maximum gradient component:", max_gc, "\n")
```

## Adapting the example

For known-parameter simulations, exact-data recovery, noisy replicates, and a
January-only timing comparison, see [Simulation and recovery with recruitment
phases](fims-recruitment-recovery.html).

Replace the illustrative tables with your observations, keeping sample ordering,
units, dates, and sample sizes consistent. Additional survey samples require
larger data objects and matching timing vectors for each affected stream; they
do not require additional modeled years. For projections, extend the annual
inputs and recruitment schedule through every future modeled year.

Length compositions require `LengthComp` data, a multinomial likelihood linked
to `lengthcomp_proportion`, and an enabled age-to-length mapping with explicit
fleet length bins. The present example deliberately uses age compositions only.

After collecting numerical results, release the TMB objective before resetting
the global FIMS registry. Do not evaluate an old objective after `clear()`.

```{r direct-cleanup, include=FALSE}
TMB::FreeADFun(obj)
rm(obj)
FIMS::clear()
```
