---
title: "Introducing the Fisheries Integrated Modeling System (FIMS)"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fisheries Integrated Modeling System
The NOAA Fisheries Integrated Modeling System (FIMS) is a new modeling framework for fisheries modeling. FIMS is a software system designed and architected to support next-generation fisheries stock assessment, ecosystem, and socioeconomic modeling. It's important to note that FIMS itself is not a model, but rather a framework for creating models. The framework is made up of many modules that come together to create a "the best model" that suites the needs of the end-user. What follows is a demo of creating a catch-at-age assessment model using FIMS. 

## Creating Models in FIMS
To begin, we import the FIMS and TMB libraries. We then create a new Rcpp fims module using `Rcpp::Module` and call `fims$clear` to ensure memory from any previous fims model run is cleared out. 
```{r fims1, warning=FALSE, message=FALSE}
#install from test-estimation branch
remotes::install_github("NOAA-FIMS/FIMS@integration-test-estimation-R-clear-FIMS")
library(FIMS)
library(TMB)

#load the module
fims <- Rcpp::Module("fims", PACKAGE = "FIMS")
fims$clear()
```

First let's set up the dimensions of the model based on the Model Comparison project:
```{r fims-dims}
nyears <- 30 # the number of years which we have data for.
nseanons<- 1 # the number of seasons in each year. FIMS currently defaults to 1
ages <- c(1,2,3,4,5,6,7,8,9,10,11,12) # age vector.
nages <- 12 # the number of age groups.
```


Data and variable values are taken from the Model Comparison project. See [R/data_mile1.R](https://github.com/NOAA-FIMS/FIMS/blob/main/R/data_mile1.R) and [tests/testthat/test-fims-estimation.R](https://github.com/NOAA-FIMS/FIMS/blob/integration-test-estimation-R-clear-FIMS/tests/testthat/test-fims-estimation.R) for details on how data and variable values are read in from the Model Comparison project. We will be reading data into the model using the FIMSFrame and FIMSFrameAge S4 R classes set up in [R/fimsframe.R](https://github.com/NOAA-FIMS/FIMS/blob/main/R/fimsframe.R)

```{r fimsframe}
#use FIMS data frame
data(package = "FIMS")
fims_frame <- FIMSFrame(data_mile1)
```

The `fims_frame` object contains a `@data` slot that holds a long data frame with catch data for the fishery and index data for the survey:

```{r ageframe}
str(fims_frame)
fims_frame@data %>% dplyr::filter(type == "landings") %>% head()
fims_frame@data %>% dplyr::filter(type == "index") %>% head()
```


Using this data frame, we will start setting up the fims data objects. This example from the Model Comparison project sets up a single fishery fleet with age composition data and catch and a single survey with age composition data and an index. Data are read into fims as long vectors, regardless of their original dimension, hence the motivation behind the long data frames that are created with the fimsframe S4 classes. 


```{r data}
#fishery data
fishery_catch <- fims_frame@data %>% 
  dplyr::filter( type == "landings" ) %>%
  .$value 
fishery_agecomp <-c(
  0.065, 0.110, 0.120, 0.115, 0.100, 0.065, 0.075, 0.065, 0.070, 0.040, 0.035, 0.140, 
  0.130, 0.095, 0.120, 0.140, 0.080, 0.080, 0.055, 0.075, 0.030, 0.045, 0.030, 0.120, 
  0.050, 0.175, 0.110, 0.150, 0.090, 0.090, 0.075, 0.050, 0.025, 0.025, 0.025, 0.135, 
  0.035, 0.075, 0.215, 0.095, 0.095, 0.060, 0.065, 0.060, 0.055, 0.055, 0.045, 0.145, 
  0.170, 0.040, 0.125, 0.190, 0.090, 0.075, 0.065, 0.060, 0.045, 0.025, 0.030, 0.085, 
  0.085, 0.155, 0.075, 0.100, 0.155, 0.070, 0.095, 0.060, 0.035, 0.040, 0.040, 0.090, 
  0.070, 0.190, 0.260, 0.070, 0.060, 0.110, 0.045, 0.050, 0.030, 0.035, 0.015, 0.065, 
  0.075, 0.105, 0.190, 0.240, 0.065, 0.035, 0.065, 0.030, 0.030, 0.045, 0.025, 0.095, 
  0.075, 0.120, 0.100, 0.170, 0.220, 0.045, 0.050, 0.055, 0.030, 0.035, 0.015, 0.085, 
  0.120, 0.110, 0.175, 0.135, 0.145, 0.115, 0.025, 0.030, 0.030, 0.025, 0.015, 0.075, 
  0.075, 0.130, 0.155, 0.130, 0.050, 0.130, 0.135, 0.025, 0.045, 0.050, 0.020, 0.055, 
  0.050, 0.160, 0.175, 0.120, 0.115, 0.110, 0.050, 0.080, 0.020, 0.020, 0.025, 0.075, 
  0.085, 0.105, 0.120, 0.185, 0.110, 0.105, 0.075, 0.050, 0.065, 0.015, 0.005, 0.080, 
  0.090, 0.120, 0.130, 0.140, 0.150, 0.060, 0.100, 0.025, 0.050, 0.025, 0.030, 0.080, 
  0.105, 0.200, 0.120, 0.100, 0.115, 0.100, 0.075, 0.045, 0.010, 0.035, 0.020, 0.075, 
  0.130, 0.125, 0.195, 0.185, 0.080, 0.065, 0.020, 0.020, 0.050, 0.015, 0.050, 0.065, 
  0.120, 0.170, 0.165, 0.160, 0.070, 0.050, 0.060, 0.045, 0.030, 0.020, 0.020, 0.090, 
  0.075, 0.180, 0.230, 0.130, 0.095, 0.040, 0.040, 0.050, 0.045, 0.015, 0.020, 0.080, 
  0.115, 0.095, 0.230, 0.190, 0.100, 0.090, 0.025, 0.020, 0.040, 0.035, 0.015, 0.045, 
  0.075, 0.145, 0.130, 0.200, 0.180, 0.045, 0.045, 0.060, 0.020, 0.025, 0.010, 0.065, 
  0.180, 0.140, 0.175, 0.085, 0.165, 0.075, 0.055, 0.035, 0.015, 0.020, 0.010, 0.045, 
  0.130, 0.255, 0.100, 0.120, 0.050, 0.120, 0.080, 0.040, 0.030, 0.015, 0.020, 0.040, 
  0.095, 0.255, 0.270, 0.075, 0.080, 0.040, 0.065, 0.035, 0.050, 0.005, 0.010, 0.020, 
  0.060, 0.195, 0.290, 0.220, 0.055, 0.055, 0.015, 0.045, 0.020, 0.010, 0.010, 0.025, 
  0.180, 0.175, 0.175, 0.205, 0.140, 0.020, 0.050, 0.020, 0.015, 0.010, 0.000, 0.010, 
  0.100, 0.195, 0.180, 0.160, 0.170, 0.085, 0.045, 0.015, 0.015, 0.020, 0.015, 0.000, 
  0.155, 0.170, 0.220, 0.115, 0.080, 0.085, 0.080, 0.010, 0.025, 0.020, 0.005, 0.035, 
  0.180, 0.225, 0.125, 0.220, 0.100, 0.070, 0.040, 0.035, 0.005, 0.000, 0.000, 0.000, 
  0.230, 0.250, 0.125, 0.085, 0.125, 0.045, 0.030, 0.025, 0.045, 0.015, 0.010, 0.015, 
  0.155, 0.290, 0.215, 0.105, 0.085, 0.090, 0.020, 0.010, 0.010, 0.000, 0.005, 0.015)

#survey data
survey_index <- fims_frame@data %>% 
  dplyr::filter( type == "index" ) %>%
  .$value 
#survey_agecomp <-  
```

## Creating a Population
Each module in the FIMS-R interface is made of S4 objects. These S4 objects serve as a interface between R and the underlining C++ code that defines FIMS. Before setting up the population module, we first need to set up each component of the population (e.g. recruitment, growth, etc.). 

### Recruitment
We'll start with setting up a Beverton Holt recruitment module. We first instantiate a module using the `methods::new()` function. We can use `methods::show()` to view all the fields (i.e. variables) and methods (i.e. functions) available in `fims$BevertonHoltRecruitment` module.

```{r recruitment}
# Recruitment
recruitment <- methods::new(fims$BevertonHoltRecruitment)
methods::show(fims$BevertonHoltRecruitment)
```
There are three parameters we need to set-up: *log_sigma_recruit*, *log_rzero*, and *logit_steep*. Each variable of [Parameter class](https://github.com/NOAA-FIMS/FIMS/blob/main/inst/include/interface/rcpp/rcpp_objects/rcpp_interface_base.hpp) has three additional fields: *value*, *is_random_effect*, and *estimated*. Currently, FIMS is not set up to run random effects. The default value for this field and the *estimate* field is currently set to `FALSE`.

```{r set-up-recruitment}
recruitment$log_sigma_recruit$value <- log(0.4)
recruitment$log_rzero$value <- log(1e+06)
recruitment$log_rzero$is_random_effect <- FALSE
recruitment$log_rzero$estimated <- TRUE
recruitment$logit_steep$value <- -log(1.0 - 0.75) + log(0.75 - 0.2)
recruitment$logit_steep$is_random_effect <- FALSE
recruitment$logit_steep$estimated <- FALSE
```

We also need to set up recruitment deviations. FIMS recruitment modules have a boolean, *estimate_deviations* to specify whether or not deviations are estimated; and a vector, *deviations* to set the deviation values. The *deviations* vector takes values greater than 0 as these will be logged in the likelihood.

```{r rec-devs}
recruitment$estimate_deviations <- FALSE
recruitment$deviations <- c(1.0931337, 1.5494153, 0.8754735, 0.6488721,
                            1.9128839, 1.6593211, 0.9327825, 1.3531871, 
                            0.9207434, 1.2304792, 1.1652038, 0.8048559, 
                            0.8752845, 1.1187967, 0.8989675, 1.3083559, 
                            1.2724463, 0.5799534, 0.7891447, 0.5571984, 
                            1.3515173, 1.2452116, 0.8002613, 0.5983474, 
                            1.1704664, 0.5828168, 0.8223697, 1.2225558,
                            1.4513402, 0.9308739)
```
### Growth
Now, we'll define the growth module for our population using an empirical weight at age model.
```{r growth}
# Growth
ewaa_growth <- methods::new(fims$EWAAgrowth)
ewaa_growth$ages <- ages
ewaa_growth$weights <- c(0.5306555, 1.1963283, 2.0582654, 
                         3.0349873, 4.0552124, 5.0646975,
                         6.0262262, 6.9169206,  7.7248909, 
                         8.4461128, 9.0818532, 9.6366950)

```   
### Maturity
Each population will also need a maturity model. Here we define a logistic maturity model. 
```{r maturity}
# Maturity
maturity <- new(fims$LogisticMaturity)
maturity$median$value <- 2.25
maturity$median$is_random_effect <- FALSE
maturity$median$estimated <- FALSE
maturity$slope$value <- 3
maturity$slope$is_random_effect <- FALSE
maturity$slope$estimated <- FALSE
```

Now that our life history sub-models are defined, lets define the actual population.

```{r population}
# Population
population <- new(fims$Population)
population$log_M <- rep(log(0.2), nages)
population$log_init_naa <- log(c(993947.5,811707.8, 661434.4, 
                                 537804.8, 436664.0, 354303.4, 
                                 287397.0, 233100.2, 189054.0, 
                                 153328.4, 124353.2, 533681.3))
population$estimate_init_naa <- TRUE
population$nages <- nages
population$ages <- ages
population$nfleets <- 1 # 1 fleet and 1 survey
population$nseasons <- 1
population$nyears <- nyears
```

Now we need to link up the recruitment, growth, and maturity modules we set above with this new population module. We do this by calling `get_id()` from each respective module and passing that unique ID into each respective `Set` function from population. 

```{r set-pop-modules}
population$SetMaturity(maturity$get_id())
population$SetGrowth(ewaa_growth$get_id())
population$SetRecruitment(recruitment$get_id())

```


## Creating a Fleet
Now that we've defined our population, lets create a fleet.
### Fleet Data
Each fleet is required to have data in order to evaluate the objective function. For this example, we'll create both index and age composition data for the fleet to use. 

```{r fleet_data}
#fleet index data
fishing_fleet_index <- new(fims$Index, nyears)
fishing_fleet_index$index_data <- fishery_catch
#fleet age composition data
fishing_fleet_age_comp <- new(fims$AgeComp, nyears, nages)
# Effective sampling size is 200
fishing_fleet_age_comp$age_comp_data <- fishery_agecomp * 200

```
### Fleet Selectivity
```{r fleet_selectivity}
fishing_fleet_selectivity <- new(fims$LogisticSelectivity)
fishing_fleet_selectivity$median$value <- 2.0
fishing_fleet_selectivity$median$is_random_effect <- FALSE
fishing_fleet_selectivity$median$estimated <- TRUE
fishing_fleet_selectivity$slope$value <- 1.0
fishing_fleet_selectivity$slope$is_random_effect <- FALSE
fishing_fleet_selectivity$slope$estimated <- TRUE
```
### Creating the Fleet Object
Now that we've created everything that a fleet needs, lets create the actual fleet object.

```{r fleet}
fishing_fleet <- new(fims$Fleet)
fishing_fleet$nages <- nages
fishing_fleet$nyears <- nyears
fishing_fleet$log_Fmort <- log(c(0.009459165, 0.02728886, 0.04506364, 
                                 0.06101782, 0.04860075, 0.08742055, 
                                 0.0884472, 0.1866079, 0.109009, 0.1327043, 
                                 0.1506155, 0.161243, 0.1166402, 0.1693461, 
                                 0.1801919, 0.1612405, 0.3145732, 0.2572476, 
                                 0.2548873, 0.2514621, 0.3491014, 0.2541077, 
                                 0.4184781, 0.3457212, 0.3436855, 0.3141712, 
                                 0.3080268, 0.4317453, 0.3280309, 0.4996754))
fishing_fleet$estimate_F <- TRUE
fishing_fleet$random_F <- FALSE
fishing_fleet$log_q <- 1.0
fishing_fleet$estimate_q <- FALSE
fishing_fleet$random_q <- FALSE
fishing_fleet$SetAgeCompLikelihood(1)
fishing_fleet$SetIndexLikelihood(1)
fishing_fleet$SetSelectivity(fishing_fleet_selectivity$get_id())
fishing_fleet$SetObservedIndexData(fishing_fleet_index$get_id())
fishing_fleet$SetObservedAgeCompData(fishing_fleet_age_comp$get_id())
```

## Creating a Survey
A survey object is essentially the same as a fleet object with a catachabilty (q) variable.

### Survey Selectivity
```{r survey_selectivity}
  # survey_fleet_selectivity <- new(fims$LogisticSelectivity)
  # survey_fleet_selectivity$median$value <- 2.0
  # survey_fleet_selectivity$median$is_random_effect <- FALSE
  # survey_fleet_selectivity$median$estimated <- TRUE
  # survey_fleet_selectivity$slope$value <- 1.0
  # survey_fleet_selectivity$slope$is_random_effect <- FALSE
  # survey_fleet_selectivity$slope$estimated <- TRUE
```

### Creating the Survey Object
```{r survey}
 # survey_fleet <- new(fims$Fleet)
 # survey_fleet$is_survey<-TRUE
 # #survey_fleet$nages <- om_input$nages
 # survey_fleet$nyears <- om_input$nyr
 #  # survey_fleet$log_Fmort <- rep(log(0.0000000000000000000000000001), om_input$nyr) #-Inf?
 # survey_fleet$estimate_F <- FALSE
 # survey_fleet$random_F <- FALSE
 # survey_fleet$log_q <- log(om_output$survey_q$survey1)
 # survey_fleet$estimate_q <- TRUE
 # survey_fleet$random_q <- FALSE
 # survey_fleet$log_obs_error$value <- log(sqrt(log(em_input$cv.survey$survey1^2+1)))
 # survey_fleet$log_obs_error$estimated <- FALSE
 # survey_fleet$SetAgeCompLikelihood(1)
 # survey_fleet$SetIndexLikelihood(1)
 # survey_fleet$SetSelectivity(survey_fleet_selectivity$get_id())
 # survey_fleet$SetObservedIndexData(survey_fleet_index$get_id())
 # survey_fleet$SetObservedAgeCompData(survey_fleet_age_comp$get_id())
  
```  
## Putting It All Together 
## Creating the FIMS Model and Making the TMB Function
```{r model}
  sucess<-fims$CreateTMBModel()
  parameters <- list(p = fims$get_fixed())
  obj <- MakeADFun(data=list(), parameters, DLL="FIMS", silent = TRUE)

```
## Fitting the Model
```{r fit_model}
  
  opt<-optim(obj$par, obj$fn, obj$gr, method = "BFGS",
                       control = list(maxit=1000000, reltol = 1e-15))

  print(opt)
```

### TMB Reporting
```{r tmb_report}

 # sdr <- TMB::sdreport(obj)
  # sdr_fixed <- summary(sdr, "fixed")
  report <- obj$report()
  
  # print(sdr_fixed)
   # print(report)
  
```


## Plotting Results
```{r plots}
print(fishing_fleet_index$index_data)
print(report$expected_index)

plot(1:nyears, fishing_fleet_index$index_data, xlab="Year", ylab="Index")
lines( report$expected_index,lty=3,lwd=2,col="blue")

```

## Clear C++ objects from memory
```{r clear}
fims$clear()
```
