---
title: "Reference Points From a CatchAtAge Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reference Points From a CatchAtAge Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(FIMS)
```

`get_reference_points()` extracts a biological snapshot from a live `CatchAtAge`
model and calculates SPR targets and Beverton-Holt MSY. This vignette builds a
small configured model so every step can run without an assessment data set.
It does not fit that model. See [SPR and MSY From a Biological Snapshot](fims-reference-points.html)
for the equations, conventions, and solver diagnostics.

## Build a small model

Run this example in a fresh R session. `clear()` resets the FIMS object registry.
The example has two modeled years, three annual ages, and two fishing fleets.
The final age is a plus group.

```{r growth-maturity}
clear()

growth <- methods::new(EWAAGrowth)
growth$n_years$set(2L)
growth$ages[] <- 1:3
# Include weights for the beginning of the year after the terminal year.
growth$weights[] <- c(1, 2, 3, 2, 3, 4, 2, 3, 4)

maturity <- methods::new(LogisticMaturity)
maturity$inflection_point[] <- 2
maturity$slope[] <- 1
```

Use Beverton-Holt recruitment with zero recruitment deviations. Reference
points use the mean stock-recruitment relationship, not the deviations.

```{r recruitment}
recruitment <- methods::new(BevertonHoltRecruitment)
process <- methods::new(LogDevsRecruitmentProcess)
recruitment$SetRecruitmentProcessID(process$get_id())
recruitment$n_years$set(2L)
recruitment$log_rzero[] <- log(1000)
recruitment$logit_steep[] <- log((0.75 - 0.2) / (1 - 0.75))
recruitment$log_devs[] <- 0
```

Natural mortality and weight change between the two years, allowing us to
examine the effect of choosing a reference year.

```{r population}
population <- methods::new(Population)
population$n_years$set(2L)
population$n_ages$set(3L)
population$n_fleets$set(2L)
population$ages[] <- 1:3
population$log_M[] <- log(c(0.2, 0.2, 0.2, 0.3, 0.4, 0.5))
population$log_init_naa[] <- log(c(1000, 500, 500))
population$proportion_female[] <- 0.5
population$SetGrowthID(growth$get_id())
population$SetMaturityID(maturity$get_id())
population$SetRecruitmentID(recruitment$get_id())
```

The fleets have the same selectivity here. Their annual F coefficients imply
a 25%/75% allocation. No observations or likelihoods are needed for this example.

```{r fleets-model}
selectivity <- methods::new(LogisticSelectivity)
selectivity$inflection_point[] <- 2
selectivity$slope[] <- 1

for (share in c(0.25, 0.75)) {
  fleet <- methods::new(Fleet)
  fleet$n_years$set(2L)
  fleet$n_ages$set(3L)
  fleet$n_lengths$set(0L)
  fleet$log_Fmort[] <- log(c(0.2, 0.4) * share)
  fleet$log_q[] <- 0
  fleet$SetSelectivityID(selectivity$get_id())
  population$AddFleet(fleet$get_id())
}

model <- methods::new(CatchAtAge)
model$AddPopulation(population$get_id())
CreateTMBModel()
```

## Select the reference year

`year` is a **one-based index into the modeled years**, not a calendar year.
The population argument is its FIMS object ID. Use `population$get_id()` rather
than assuming that ID equals its position in a list.

```{r reference-year}
reference <- get_reference_points(
  model, population$get_id(), year = 2L,
  spr_targets = c(0.4, 0.35)
)
knitr::kable(reference$spr, digits = 6)
reference$msy[c("fishing_mortality", "yield", "spawning_biomass", "status")]
```

The call refreshes derived quantities from the current internal double
parameters, as `model$get_output()` does. For a fitted model, first synchronize
its fitted fixed effects with `set_fixed()` and its fitted random effects with
`set_random()`, in the respective parameter orders returned by `get_fixed()`
and `get_random()`. Passing a live model does not automatically retrieve
estimates from a saved `FIMSFit`. The calculation does not refit the model or
change its parameters.

Inspect the returned snapshot before interpreting a result:

```{r inspect-snapshot}
reference$inputs[c("population_id", "year_index", "ages", "fleet_ids")]
knitr::kable(data.frame(
  age = reference$inputs$ages,
  natural_mortality = reference$inputs$natural_mortality,
  weight = reference$inputs$weight,
  maturity = reference$inputs$maturity
), digits = 4)
vapply(reference$inputs$fleets, `[[`, numeric(1), "share")
```

The adapter copies mortality, maturity, female proportions, biological weight,
and fleet selectivity for the selected year. Catch weights follow the
population weights used by `CatchAtAge`. The live adapter requires at least
two consecutive annual ages.

## Preserve the recruitment baseline

SPR uses the selected year's unfished spawning biomass per recruit. For MSY,
the Beverton-Holt relationship retains the model's initial `phi0` baseline.
Choosing a later reference year does not redefine the fitted recruitment
relationship.

```{r compare-years}
first_year <- get_reference_points(model, population$get_id(), year = 1L)
knitr::kable(data.frame(
  year_index = c(1L, 2L),
  reference_year_SBPR0 = c(
    first_year$unfished_per_recruit$spawning_biomass,
    reference$unfished_per_recruit$spawning_biomass
  ),
  recruitment_phi0 = c(
    first_year$inputs$recruitment$phi0,
    reference$inputs$recruitment$phi0
  )
), digits = 4)
```

Because biology differs between years, the second year's unfished equilibrium
recruitment need not equal the original R0. This distinction matters when
comparing reference years or changing assumptions in a saved snapshot.

## Override the fishing allocation

Default fleet shares are the reference year's F coefficients normalized to sum
to one. To prescribe shares, supply one per fleet in ascending fleet ID order,
as recorded in `inputs$fleet_ids`. Use zero for fleets to exclude, such as
surveys. If all fitted F coefficients are zero, explicit shares are required.

```{r allocation}
first_fleet_only <- get_reference_points(
  model, population$get_id(), year = 2L,
  fleet_shares = c(1, 0)
)
first_fleet_only$msy$fleet_yield
```

Since these two example fleets have identical selectivities and weights,
changing their allocation changes the partition of yield but leaves the total
reference points unchanged. Different fishing patterns can change both SPR
and MSY reference points.

## Reuse a snapshot after clearing the live model

The result holds ordinary R data. Save the entire result with `saveRDS()` to
retain both inputs and solver settings. Recalculation needs no live FIMS objects.

```{r reuse-snapshot}
snapshot <- reference$inputs
clear()
recalculated <- calculate_reference_points(snapshot, spr_targets = c(0.4, 0.35))
isTRUE(all.equal(reference, recalculated))
```

```{r validate-workflow, include=FALSE}
stopifnot(
  all(reference$spr$status == "converged"),
  reference$msy$status == "converged",
  isTRUE(all.equal(reference$inputs$natural_mortality, c(0.3, 0.4, 0.5))),
  first_year$inputs$recruitment$phi0 == reference$inputs$recruitment$phi0,
  first_fleet_only$msy$fleet_yield[2] == 0,
  isTRUE(all.equal(first_fleet_only$msy$yield, reference$msy$yield)),
  isTRUE(all.equal(reference, recalculated))
)
```

Inspect solver status for every calculation. A converged bounded MSY search
still requires sensible biological assumptions and checks of the search bounds.
The current implementation supports Beverton-Holt MSY and does not calculate
standard errors; use `msy = FALSE` when only SPR targets are needed.
