---
title: "Multi-population modeling in the Fisheries Integrated Modeling System (FIMS)"
output: github_document
vignette: >
  %\VignetteIndexEntry{Multi-population modeling in the Fisheries Integrated Modeling System (FIMS)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates how to configure a FIMS catch-at-age model with two
population objects. The example intentionally keeps the two populations simple:
they share the same recruitment, growth, maturity, fishing fleet, and survey
fleet modules. This is useful for demonstrating how multiple populations are
registered with the top-level `CatchAtAge` model.

The workflow is:

1. load operating-model data,
2. create observed-data objects for landings, survey indices, and age compositions,
3. configure fleet and survey modules,
4. define biological modules shared by both populations,
5. create two population modules,
6. add both populations to a `CatchAtAge` model, and
7. build and fit the TMB objective function.

> **Important:** because the two populations in this example share the same fleet,
> survey, recruitment, growth, and maturity modules, this vignette should be read
> as a minimal multi-population setup example. A more realistic application may
> require population-specific fleets, surveys, recruitment processes, movement, or
> biological parameters.

## Load FIMS and example data

```{r load-data}
library(FIMS)

# Set the iteration ID to 1 for accessing specific input/output lists.
iter_id <- 1

# Toggle estimation. If FALSE, the model reports values at the input parameters.
estimation_mode <- TRUE

# Optional random-effects configuration. Set to NULL for the default example.
random_effects <- NULL
map <- list()

c_case <- 1
github_dir <- paste0(
  "https://github.com/NOAA-FIMS/",
  "Age_Structured_Stock_Assessment_Model_Comparison/raw/main/",
  "FIMS_integration_test_data/FIMS_C",
  c_case,
  "/output/OM/"
)

r_data_file <- "OM1.RData"
load(url(paste0(github_dir, r_data_file)))

# Clear any previously registered FIMS modules before building the model.
clear()
```

## Create observed data objects

The fleet module uses observed landings and age-composition data. The survey
module uses an abundance index and age-composition data.

```{r observed-data}
# Observed fishery landings.
landings <- em_input[["L.obs"]][["fleet1"]]
fishing_fleet_landings <- methods::new(Landings, om_input[["nyr"]])

purrr::walk(
  seq_len(om_input[["nyr"]]),
  \(i) fishing_fleet_landings$landings_data$set(i - 1, landings[i])
)

# Observed fishery age compositions, scaled by sample size for likelihood weighting.
fishing_fleet_age_comp <- methods::new(
  AgeComp,
  om_input[["nyr"]],
  om_input[["nages"]]
)

fishery_age_comp_counts <-
  c(t(em_input[["L.age.obs"]][["fleet1"]])) * em_input[["n.L"]][["fleet1"]]

purrr::walk(
  seq_along(fishery_age_comp_counts),
  \(i) fishing_fleet_age_comp$age_comp_data$set(i - 1, fishery_age_comp_counts[i])
)

# Observed survey index.
survey_index <- em_input[["survey.obs"]][["survey1"]]
survey_fleet_index <- methods::new(Index, om_input[["nyr"]])

purrr::walk(
  seq_len(om_input[["nyr"]]),
  \(i) survey_fleet_index$index_data$set(i - 1, survey_index[i])
)

# Observed survey age compositions, scaled by sample size.
survey_fleet_age_comp <- methods::new(
  AgeComp,
  om_input[["nyr"]],
  om_input[["nages"]]
)

survey_age_comp_counts <-
  c(t(em_input[["survey.age.obs"]][["survey1"]])) * em_input[["n.survey"]][["survey1"]]

purrr::walk(
  seq_along(survey_age_comp_counts),
  \(i) survey_fleet_age_comp$age_comp_data$set(i - 1, survey_age_comp_counts[i])
)
```

## Configure the fishery fleet

```{r fishery-fleet}
fishing_fleet_selectivity <- methods::new(LogisticSelectivity)
fishing_fleet_selectivity$inflection_point[1]$value <-
  om_input[["sel_fleet"]][["fleet1"]][["A50.sel1"]]
fishing_fleet_selectivity$inflection_point[1]$estimation_type$set("fixed_effects")

fishing_fleet_selectivity$slope[1]$value <-
  om_input[["sel_fleet"]][["fleet1"]][["slope.sel1"]]
fishing_fleet_selectivity$slope[1]$estimation_type$set("fixed_effects")

fishing_fleet <- methods::new(Fleet)
fishing_fleet$n_years$set(om_input[["nyr"]])
fishing_fleet$n_ages$set(om_input[["nages"]])

fishing_fleet$log_Fmort$resize(om_input[["nyr"]])
for (y in seq_len(om_input[["nyr"]])) {
  fishing_fleet$log_Fmort[y]$value <- log(om_output[["f"]][y])
}
fishing_fleet$log_Fmort$set_all_estimable(TRUE)

# Catchability is fixed at 1 for the fishery landings example.
fishing_fleet$log_q[1]$value <- log(1.0)
fishing_fleet$log_q[1]$estimation_type$set("constant")

fishing_fleet$SetSelectivityID(fishing_fleet_selectivity$get_id())
fishing_fleet$SetObservedLandingsDataID(fishing_fleet_landings$get_id())
fishing_fleet$SetObservedAgeCompDataID(fishing_fleet_age_comp$get_id())

# Landings likelihood: lognormal observation model.
fishing_fleet_landings_distribution <- methods::new(DlnormDistribution)
fishing_fleet_landings_distribution$log_sd$resize(om_input[["nyr"]])

for (y in seq_len(om_input[["nyr"]])) {
  # Convert coefficient of variation to lognormal standard deviation,
  # then store it on the log scale required by DlnormDistribution.
  cv <- em_input[["cv.L"]][["fleet1"]]
  fishing_fleet_landings_distribution$log_sd[y]$value <- log(sqrt(log(cv^2 + 1)))
}
fishing_fleet_landings_distribution$log_sd$set_all_estimable(FALSE)
fishing_fleet_landings_distribution$set_observed_data(
  fishing_fleet$GetObservedLandingsDataID()
)
fishing_fleet_landings_distribution$set_distribution_links(
  "data",
  fishing_fleet$log_landings_expected$get_id()
)

# Age-composition likelihood: multinomial observation model.
fishing_fleet_agecomp_distribution <- methods::new(DmultinomDistribution)
fishing_fleet_agecomp_distribution$set_observed_data(
  fishing_fleet$GetObservedAgeCompDataID()
)
fishing_fleet_agecomp_distribution$set_distribution_links(
  "data",
  fishing_fleet$agecomp_proportion$get_id()
)
```

## Configure the survey fleet

The survey is represented with a fleet object that has negligible fishing
mortality and an estimable survey catchability parameter.

```{r survey-fleet}
survey_fleet_selectivity <- methods::new(LogisticSelectivity)
survey_fleet_selectivity$inflection_point[1]$value <-
  om_input[["sel_survey"]][["survey1"]][["A50.sel1"]]
survey_fleet_selectivity$inflection_point[1]$estimation_type$set("fixed_effects")

survey_fleet_selectivity$slope[1]$value <-
  om_input[["sel_survey"]][["survey1"]][["slope.sel1"]]
survey_fleet_selectivity$slope[1]$estimation_type$set("fixed_effects")

survey_fleet <- methods::new(Fleet)
survey_fleet$n_years$set(om_input[["nyr"]])
survey_fleet$n_ages$set(om_input[["nages"]])

survey_fleet$log_Fmort$resize(om_input[["nyr"]])
for (y in seq_len(om_input[["nyr"]])) {
  survey_fleet$log_Fmort[y]$value <- -200
}
survey_fleet$log_Fmort$set_all_estimable(FALSE)

survey_fleet$log_q[1]$value <- log(om_output[["survey_q"]][["survey1"]])
survey_fleet$log_q[1]$estimation_type$set("fixed_effects")

survey_fleet$SetSelectivityID(survey_fleet_selectivity$get_id())
survey_fleet$SetObservedIndexDataID(survey_fleet_index$get_id())
survey_fleet$SetObservedAgeCompDataID(survey_fleet_age_comp$get_id())

# Survey index likelihood: lognormal observation model.
survey_fleet_index_distribution <- methods::new(DlnormDistribution)
survey_fleet_index_distribution$log_sd$resize(om_input[["nyr"]])

for (y in seq_len(om_input[["nyr"]])) {
  cv <- em_input[["cv.survey"]][["survey1"]]
  survey_fleet_index_distribution$log_sd[y]$value <- log(sqrt(log(cv^2 + 1)))
}
survey_fleet_index_distribution$log_sd$set_all_estimable(FALSE)
survey_fleet_index_distribution$set_observed_data(
  survey_fleet$GetObservedIndexDataID()
)
survey_fleet_index_distribution$set_distribution_links(
  "data",
  survey_fleet$log_index_expected$get_id()
)

# Survey age-composition likelihood.
survey_fleet_agecomp_distribution <- methods::new(DmultinomDistribution)
survey_fleet_agecomp_distribution$set_observed_data(
  survey_fleet$GetObservedAgeCompDataID()
)
survey_fleet_agecomp_distribution$set_distribution_links(
  "data",
  survey_fleet$agecomp_proportion$get_id()
)

# Age-to-length conversion matrix used by the survey fleet.
for (i in seq_along(em_input[["age_to_length_conversion"]])) {
  survey_fleet$age_to_length_conversion[i]$value <-
    c(t(em_input[["age_to_length_conversion"]]))[i]
}
survey_fleet$age_to_length_conversion$set_all_estimable(FALSE)
survey_fleet$age_to_length_conversion$set_all_random(FALSE)
```

## Optional selectivity random effects

```{r selectivity-random-effects}
if (!is.null(random_effects) && "selectivity" %in% names(random_effects)) {
  if (random_effects[["selectivity"]] == "log_devs") {
    fishing_fleet_selectivity$log_devs$set_all_random(TRUE)
    survey_fleet_selectivity$log_devs$set_all_random(TRUE)
  }

  if (random_effects[["selectivity"]] == "log_sel") {
    fishing_fleet_selectivity$log_sel$set_all_random(TRUE)
    survey_fleet_selectivity$log_sel$set_all_random(TRUE)
  }

  if (random_effects[["selectivity"]] == "pars") {
    fishing_fleet_selectivity$inflection_point$estimation_type$set("random_effects")
    fishing_fleet_selectivity$slope$estimation_type$set("random_effects")
    survey_fleet_selectivity$inflection_point$estimation_type$set("random_effects")
    survey_fleet_selectivity$slope$estimation_type$set("random_effects")
  }
}
```

## Define shared recruitment

Both populations use the same Beverton-Holt recruitment module in this example.
Sharing the module means both populations reference the same recruitment
parameters and process.

```{r recruitment}
recruitment <- methods::new(BevertonHoltRecruitment)

if (is.null(random_effects) || random_effects[["recruitment"]] == "log_devs") {
  recruitment_process <- methods::new(LogDevsRecruitmentProcess)
} else {
  recruitment_process <- methods::new(LogRRecruitmentProcess)
}

recruitment$SetRecruitmentProcessID(recruitment_process$get_id())
recruitment$n_years$set(om_input[["nyr"]])

# Equilibrium recruitment.
recruitment$log_rzero[1]$value <- log(om_input[["R0"]])
recruitment$log_rzero[1]$estimation_type$set("fixed_effects")

# Beverton-Holt steepness on the FIMS logit scale.
recruitment$logit_steep[1]$value <-
  -log(1.0 - om_input[["h"]]) + log(om_input[["h"]] - 0.2)
recruitment$logit_steep[1]$estimation_type$set("constant")

if (is.null(random_effects) || random_effects[["recruitment"]] == "log_devs") {
  recruitment$log_devs$resize(om_input[["nyr"]] - 1)
  for (y in seq_len(om_input[["nyr"]] - 1)) {
    recruitment$log_devs[y]$value <- om_input[["logR.resid"]][y + 1]
  }
}

if (!is.null(random_effects) && "recruitment" %in% names(random_effects)) {
  if (random_effects[["recruitment"]] == "log_devs") {
    recruitment$log_devs$set_all_random(TRUE)
  }

  if (random_effects[["recruitment"]] == "log_r") {
    recruitment$log_r$resize(om_input[["nyr"]] - 1)
    recruits_true <- matrix(
      c(t(om_output[["N.age"]])),
      om_input[["nyr"]],
      om_input[["nages"]],
      byrow = TRUE
    )[, 1]

    for (y in seq_len(om_input[["nyr"]] - 1)) {
      recruitment$log_r[y]$value <- if (estimation_mode) 0 else recruits_true[y + 1]
    }
    recruitment$log_r$set_all_random(TRUE)
  }
}

if (is.null(random_effects)) {
  recruitment$log_devs$set_all_estimable(TRUE)
}

recruitment_distribution <- methods::new(DnormDistribution)
recruitment_distribution$log_sd$resize(1)
recruitment_distribution$log_sd[1]$value <- log(om_input[["logR_sd"]])

recruitment_distribution$observed_values$resize(om_input[["nyr"]] - 1)
recruitment_distribution$expected_values$resize(om_input[["nyr"]] - 1)

for (i in seq_len(om_input[["nyr"]] - 1)) {
  recruitment_distribution$observed_values[i]$value <- 0
  recruitment_distribution$expected_values[i]$value <- 0
}

if (!is.null(random_effects) && "recruitment" %in% names(random_effects)) {
  if (random_effects[["recruitment"]] == "log_devs") {
    recruitment_distribution$log_sd[1]$estimation_type$set("fixed_effects")
    recruitment_distribution$set_distribution_links(
      "random_effects",
      recruitment$log_devs$get_id()
    )
  }

  if (random_effects[["recruitment"]] == "log_r") {
    recruitment_distribution$log_sd[1]$value <- log(1)
    recruitment_distribution$log_sd[1]$estimation_type$set("fixed_effects")
    recruitment_distribution$set_distribution_links(
      "random_effects",
      c(
        recruitment$log_r$get_id(),
        recruitment$log_expected_recruitment$get_id()
      )
    )
  }
} else {
  recruitment_distribution$set_distribution_links(
    "random_effects",
    recruitment$log_devs$get_id()
  )
}
```

## Define shared growth and maturity

```{r growth-maturity}
ewaa_growth <- methods::new(EWAAGrowth)
ewaa_growth$n_years$set(om_input[["nyr"]])
ewaa_growth$ages$resize(om_input[["nages"]])

purrr::walk(
  seq_along(om_input[["ages"]]),
  \(i) ewaa_growth$ages$set(i - 1, om_input[["ages"]][i])
)

ewaa_growth$weights$resize((om_input[["nyr"]] + 1) * om_input[["nages"]])

purrr::walk(
  seq_len(ewaa_growth$weights$size()),
  \(i) ewaa_growth$weights$set(
    i - 1,
    om_input[["W.mt"]][((i - 1) %% om_input[["nages"]]) + 1]
  )
)

maturity <- methods::new(LogisticMaturity)
maturity$inflection_point[1]$value <- om_input[["A50.mat"]]
maturity$inflection_point[1]$estimation_type$set("constant")
maturity$slope[1]$value <- om_input[["slope.mat"]]
maturity$slope[1]$estimation_type$set("constant")
```

## Define two population modules

The helper below keeps the population setup consistent and avoids duplicating the
same block of code for each population. Each population receives its own
`Population` object but references the same biological and fleet modules.

```{r populations}
make_population <- function() {
  population <- methods::new(Population)

  population$log_M$resize(om_input[["nyr"]] * om_input[["nages"]])
  for (i in seq_len(om_input[["nyr"]] * om_input[["nages"]])) {
    population$log_M[i]$value <- log(om_input[["M.age"]][1])
  }
  population$log_M$set_all_estimable(FALSE)

  population$log_init_naa$resize(om_input[["nages"]])
  for (i in seq_len(om_input[["nages"]])) {
    population$log_init_naa[i]$value <- log(om_output[["N.age"]][1, i])
  }
  population$log_init_naa$set_all_estimable(TRUE)

  population$n_ages$set(om_input[["nages"]])
  population$ages$resize(om_input[["nages"]])
  purrr::walk(
    seq_along(om_input[["ages"]]),
    \(i) population$ages$set(i - 1, om_input[["ages"]][i])
  )

  population$n_fleets$set(sum(om_input[["fleet_num"]], om_input[["survey_num"]]))
  population$n_years$set(om_input[["nyr"]])

  population$SetRecruitmentID(recruitment$get_id())
  population$SetGrowthID(ewaa_growth$get_id())
  population$SetMaturityID(maturity$get_id())
  population$AddFleet(fishing_fleet$get_id())
  population$AddFleet(survey_fleet$get_id())

  population
}

population1 <- make_population()
population2 <- make_population()
```

## Create and fit the catch-at-age model

```{r fit-model}
caa <- methods::new(CatchAtAge)
caa$AddPopulation(population1$get_id())
caa$AddPopulation(population2$get_id())

CreateTMBModel()

parameters <- list(
  p = get_fixed(),
  re = get_random()
)

obj <- TMB::MakeADFun(
  data = list(),
  parameters = parameters,
  DLL = "FIMS",
  silent = TRUE,
  map = map,
  random = "re"
)

if (estimation_mode) {
  opt <- stats::nlminb(
    obj[["par"]],
    obj[["fn"]],
    obj[["gr"]],
    control = list(eval.max = 10000, iter.max = 10000, trace = 0)
  )

  FIMS::set_fixed(opt$par)
  fims_finalized <- caa$get_output()

  sdr <- TMB::sdreport(obj)
  sdr_report <- summary(sdr, "report")
  sdr_fixed <- summary(sdr, "fixed")
  sdr_random <- summary(sdr, "random")
} else {
  opt <- NULL
  sdr <- list()
  sdr_report <- list()
  sdr_fixed <- matrix(
    obj[["env"]]$parList()[["p"]],
    ncol = 1,
    dimnames = list(NULL, "Estimate")
  )
  sdr_random <- matrix(
    obj[["env"]]$parList()[["re"]],
    ncol = 1,
    dimnames = list(NULL, "Estimate")
  )
}

row.names(sdr_fixed) <- names(FIMS:::get_parameter_names(sdr_fixed[, 1]))
row.names(sdr_random) <- names(FIMS:::get_random_names(sdr_random[, 1]))

report <- obj[["report"]](obj[["env"]][["last.par.best"]])
```

## Inspect results

```{r inspect-results}
# Gradient at the final parameter values. Values near zero indicate that the
# optimizer has likely reached a local optimum.
obj$gr()

# Optimizer diagnostics and convergence code.
opt

# Reported quantities from the model.
head(sdr_report)
```

## Clean up

```{r cleanup}
clear()
```

## Notes for extending this example

This example demonstrates the mechanics of adding more than one population to a
single catch-at-age model. Extensions may include:

- population-specific recruitment modules,
- population-specific growth or maturity parameters,
- separate fleets or surveys for each population,
- movement between populations or areas, and
- population-specific likelihood components.

When adapting this example, check whether a module should be shared across
populations or instantiated separately. Sharing a module means the populations use
the same registered object and therefore the same parameterization.
