---
title: "FIMS Demo"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fisheries Integrated Modeling System
The NOAA Fisheries Integrated Modeling System (FIMS) is a new modeling framework for fisheries modeling. FIMS is a software system designed and architected to support next-generation fisheries stock assessment, ecosystem, and socioeconomic modeling. It's important to note that FIMS itself is not a model, but rather a framework for creating models. The framework is made up of many modules that come together to create a "the best model" that suites the needs of the end-user. What follows is a demo of creating an assessment model using FIMS.

## Creating Models in FIMS
To begin, we import the FIMS library and define some important variables that describe the dimensions of our data.
```{r fims1}
library(FIMS)

nyears<-30  # the number of years which we have data for.
nseanons<-1 # the number of seasons in each year.
ages<-c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)   # age vector.
nages<-12   # the number of age groups.

```


## Creating a Population
Each module in the FIMS/R interfqace is made of S4 objects. These S4 objects serve as a interface between R and the underlining C++ code that defines FIMS. Let's now define a population.

```{r fims2}

```
### Recruitment
We'll start with a recruitment module.

### Growth

### Maturity

## Creating a Fleet
### Selectivity
### Adding Data
## Creating a Survey
### Selectivity
### Adding Data
## Putting It All Together 
## Creating a TMB Model



