---
title: "FIMS Path: Growth (VonBertalanffy)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FIMS Path: Growth (VonBertalanffy)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, echo=FALSE}
.rchunk {
  border: solid #2e75b6;
}
.rcppchunk {
  border: solid #c55a11;
}
.fimschunk {
  border: solid #548235;
}
.infochunk {
  border: solid #262626;
}
```

<style type="text/css">
code span.co{
  color: #8c8c8c;
}
</style>

This vignette describes the hierarchical structure of FIMS by tracing the path
from R to C++ for the **Growth v1** module using the fixed-parameter
**VonBertalanffy** implementation. The goal is to show how values set in R
propagate through the Rcpp interface and into the C++ core where
length-at-age (LAA), SD LAA, and weight-at-age (WAA) are computed and cached.
It also shows how the current branch now carries those growth outputs into the
shared population size subsystem used by migrated length-based consumers. The
VonB parameterization used here is the SS3-style **two-reference-age** form
(`length_at_ref_age_1`, `length_at_ref_age_2` at
`reference_age_for_length_1`, `reference_age_for_length_2`), sometimes
referred to informally as a "two-point" or "length-at-two-ages"
parameterization.

## Modules in R

Growth parameters are created through the standard configuration/parameter workflow. The Growth module type is switched to `VonBertalanffy`, default parameters are created, and optional overrides are applied.

```{r, class.source = "rchunk", eval = FALSE}
library(FIMS)
data("data_big")
fims_frame <- FIMSFrame(data_big)

# Switch Growth to VonBertalanffy
config <- create_default_configurations(data = fims_frame) |>
  tidyr::unnest(cols = data)
config$module_type[config$module_name == "Growth"] <- "VonBertalanffy"
config <- tidyr::nest(config, .by = c(model_family, module_name, fleet_name))

# Create default parameters (Growth defaults live here)
params <- create_default_parameters(configurations = config, data = fims_frame) |>
  tidyr::unnest(cols = data)

# Optional: override Growth values
params <- dplyr::rows_update(
  params,
  tibble::tibble(
    module_name = "Growth",
    label = c("length_at_ref_age_1", "length_at_ref_age_2",
              "growth_coefficient_K", "length_weight_a", "length_weight_b",
              "reference_age_for_length_1", "reference_age_for_length_2"),
    value = c(8, 60, 0.2, 1e-5, 3.0, 1, 12)
  ),
  by = c("module_name", "label")
)

# Initialize Growth module (usually happens inside initialize_fims)
growth <- FIMS:::initialize_growth(parameters = params, data = fims_frame)
```

The defaults for growth parameters (including `reference_age_for_length_1`, `reference_age_for_length_2`, and `length_at_age_sd_at_ref_ages`) are created in `R/create_default_parameters.R`. Missing reference ages are treated as an error during initialization so defaults remain centralized.

## Rcpp Interface

The Growth Rcpp interface exposes a `VonBertalanffyGrowth` class with parameter fields. These are registered in `src/fims_modules.hpp` and implemented in `inst/include/interface/rcpp/rcpp_objects/rcpp_growth.hpp`.

```{Rcpp, class.source = "rcppchunk", eval = FALSE}
Rcpp::class_<VonBertalanffyGrowthInterface>("VonBertalanffyGrowth")
  .constructor()
  .field("length_at_ref_age_1",
         &VonBertalanffyGrowthInterface::length_at_ref_age_1)
  .field("length_at_ref_age_2",
         &VonBertalanffyGrowthInterface::length_at_ref_age_2)
  .field("growth_coefficient_K",
         &VonBertalanffyGrowthInterface::growth_coefficient_K)
  .field("reference_age_for_length_1",
         &VonBertalanffyGrowthInterface::reference_age_for_length_1)
  .field("reference_age_for_length_2",
         &VonBertalanffyGrowthInterface::reference_age_for_length_2)
  .field("length_weight_a",
         &VonBertalanffyGrowthInterface::length_weight_a)
  .field("length_weight_b",
         &VonBertalanffyGrowthInterface::length_weight_b)
  .field("length_at_age_sd_at_ref_ages",
         &VonBertalanffyGrowthInterface::length_at_age_sd_at_ref_ages)
  .field("n_ages", &VonBertalanffyGrowthInterface::n_ages)
  .method("get_id", &VonBertalanffyGrowthInterface::get_id)
  .method("evaluate", &VonBertalanffyGrowthInterface::evaluate);
```

These fields mirror the fixed parameters used by the C++ growth model: `length_at_ref_age_1`, `length_at_ref_age_2`, `growth_coefficient_K`, `reference_age_for_length_1`, `reference_age_for_length_2`, `length_weight_a`, `length_weight_b`, and `length_at_age_sd_at_ref_ages`.

## C++ Core

Growth v1 is implemented as a small stack of C++ types:

- **Functor:** `VonBertalanffyGrowth` in `inst/include/population_dynamics/growth/functors/von_bertalanffy.hpp`  
- **Model:** `GrowthModel` in `inst/include/population_dynamics/growth/growth_model.hpp`  
- **Products:** `GrowthProducts` in `inst/include/population_dynamics/growth/growth_products.hpp`  
- **Adapter/capability layer:** `VonBertalanffyGrowthModelAdapter` in `inst/include/population_dynamics/growth/growth_model_adapter.hpp`

The functor defines the core formula:

```{Rcpp, class.source = "fimschunk", eval = FALSE}
// file: inst/include/population_dynamics/growth/functors/von_bertalanffy.hpp
// L(a) = length_at_ref_age_1 + (length_at_ref_age_2 - length_at_ref_age_1) * (1 - exp(-growth_coefficient_K * (a - reference_age_for_length_1))) /
//        (1 - exp(-growth_coefficient_K * (reference_age_for_length_2 - reference_age_for_length_1)))
```

The model wraps the functor and computes cached products (mean LAA, SD LAA,
mean WAA) across years, ages, and sexes. The data structures are already
dimensioned as `(year, age, sex)` even though v1 uses a single year and a
single sex:

```{Rcpp, class.source = "fimschunk", eval = FALSE}
// file: inst/include/population_dynamics/growth/growth_model.hpp
GrowthModel::Prepare() {
  // validate parameters
  // compute LAA/SD/WAA and store in GrowthProducts
}
```

## How Growth Connects To The Shared Size Path

In the current branch, growth products are no longer the end of the story for
migrated length-based consumers. Instead:

- growth computes and caches upstream biological products such as `MeanLAA`,
  `SdLAA`, and `MeanWAA`
- population owns the canonical biological size support through
  `Population::size_grid`
- a `GrowthDerivedSizeProvider` reads the prepared growth products and builds
  `ProbSize` on that biological grid
- fleet observation bins remain fleet-owned, and runtime ALK helpers map the
  population size products into fleet observation space when needed

This keeps biological size support population-level while leaving fleet bins as
observation geometry only.

The provider is the bridge between prepared growth products and the migrated
size path:

```{Rcpp, class.source = "fimschunk", eval = FALSE}
// file: inst/include/population_dynamics/size/growth_derived_size_provider.hpp
class GrowthDerivedSizeProvider : public SizeDistributionProviderBase<Type> {
  // forwards MeanLAA and SdLAA from prepared GrowthProducts
  // prepares ProbSize on Population::size_grid
};
```

The runtime ALK helpers then consume that provider rather than rebuilding
fleet-bin probabilities directly from growth products:

```{Rcpp, class.source = "fimschunk", eval = FALSE}
// file: inst/include/population_dynamics/alk/functors/alk_runtime.hpp
std::make_shared<GrowthDerivedALK<Type>>(
    fleet,
    growth_observation,
    population->size_distribution_provider);
```

## Wiring into Information and Population

Growth objects are stored in `Information` and attached to populations. The
growth adapter still bridges the product-based model to the legacy
`evaluate(age)` interface, but `Information` now also configures the shared
population size path when the growth object exposes the
`GrowthDerivedObservationBase` capability.

```{Rcpp, class.source = "infochunk", eval = FALSE}
// file: inst/include/common/information.hpp
if (std::shared_ptr<fims_popdy::GrowthDerivedObservationBase<Type>>
        growth_observation = std::dynamic_pointer_cast<
            fims_popdy::GrowthDerivedObservationBase<Type>>(p->growth)) {
  growth_observation->Initialize(p->n_years, p->n_ages, 1);
  growth_observation->SetAgeOffset(min_age);
  EnsurePopulationSizeGrid(valid_model, p);
  ConfigureGrowthDerivedSizeProvider(p, growth_observation);
}
```

The adapter still uses cached WAA for the direct `evaluate(age)` path:

```{Rcpp, class.source = "fimschunk", eval = FALSE}
// file: inst/include/population_dynamics/growth/growth_model_adapter.hpp
return p.MeanWAA(0, age_index, 0);
```

Population now holds both a generic growth pointer and the population-owned
size objects used by the migrated length-based path:

```{Rcpp, class.source = "fimschunk", eval = FALSE}
// file: inst/include/population_dynamics/population/population.hpp
std::shared_ptr<fims_popdy::GrowthBase<Type>> growth;
SizeGrid size_grid;
std::shared_ptr<fims_popdy::SizeDistributionProviderBase<Type>>
    size_distribution_provider;
```

This keeps the population code generic and allows new growth variants to plug
in without changing downstream logic. It also leaves room for future expansion
to sex-specific or time-varying growth by populating `n_sexes` and `n_years`
when those models are added. In the current pooled implementation,
`Information` initializes the growth-derived capability with `n_sexes = 1`,
uses the minimum modeled age as the age offset, prepares a valid population
biological size grid, and configures the provider that downstream ALK and WAA
logic can reuse.

## Summary

Growth v1 is a fixed-parameter VonBertalanffy implementation with a clean
R -> Rcpp -> C++ pipeline. Parameters are defined in R defaults, passed
through Rcpp, and used in C++ to compute LAA, SD LAA, and WAA products. In the
current branch, those prepared growth products also feed a shared population
size subsystem: the growth-derived size provider prepares population-grid size
probabilities, the population owns the biological size grid, fleets keep their
observation bins, and migrated downstream consumers such as the runtime ALK
path use that shared structure instead of rebuilding fleet-bin logic directly
from growth outputs.
