---
title: "Empirical Dynamic Modeling (EDM) Workflow in FIMS"
output: github_document
vignette: >
  %\VignetteIndexEntry{Empirical Dynamic Modeling (EDM) Workflow in FIMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
withr::local_options(list(
  cli.num_colors   = 1,
  crayon.enabled   = FALSE,
  pillar.subtle    = FALSE
))
```

## Overview

Empirical Dynamic Modeling (EDM) is a nonparametric framework for analyzing
nonlinear dynamical systems from time series data. Unlike parametric stock
assessment models, EDM makes no assumptions about functional forms — it
reconstructs the attractor geometry of the underlying system directly from
lagged coordinates of the observed time series (Takens' theorem).

FIMS implements three EDM prediction algorithms through a high-level R
interface backed by C++ functors:

| Method | Function | Best suited for |
|--------|----------|----------------|
| **Simplex Projection** | `fit_edm(method = "simplex")` | Identifying embedding dimension, nonlinearity testing |
| **S-Map** | `fit_edm(method = "smap")` | Quantifying nonlinearity strength, state-dependent coefficients |
| **GP-EDM** | `fit_edm(method = "gp")` | Probabilistic prediction, uncertainty quantification |

This vignette walks through the complete univariate EDM workflow — from
embedding dimension selection to multi-step forecasting — using the built-in
`data_big` dataset.

## Setup

```{r library}
library(FIMS)
library(ggplot2)

data("data_big")
ff <- FIMSFrame(data_big)
```

## Step 1 — Select the Embedding Dimension E

The first step in any EDM analysis is finding the optimal embedding dimension
`E`. `select_embedding_dimension()` sweeps `E` from `E_min` to `E_max`,
runs Simplex Projection with `n_neighbors = E + 1` for each candidate, and
returns the Pearson correlation (ρ) and RMSE between in-library predictions
and observations.

```{r select-E}
edim <- select_embedding_dimension(
  ff,
  series_type = "catch",
  series_name = "fleet1",
  E_min = 1L,
  E_max = 10L
)
print(edim)

optimal_E <- edim$E[edim$optimal]
cat("Optimal E:", optimal_E, "\n")
```

```{r plot-E, fig.alt="Prediction skill vs embedding dimension E"}
ggplot(edim, aes(x = E, y = rho, colour = optimal)) +
  geom_line(colour = "grey60") +
  geom_point(size = 3) +
  scale_colour_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick")) +
  labs(title = "Simplex prediction skill vs embedding dimension",
       x = "Embedding dimension (E)", y = expression(rho)) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Step 2 — Build the Delay Embedding

Once the optimal `E` is identified, use `create_edm_embedding()` to store the
delay-embedding matrix inside the `FIMSFrame`. This is required before calling
`select_n_neighbors()` or `fit_edm()`.

```{r create-embedding}
ff <- create_edm_embedding(
  ff,
  series_type = "catch",
  series_name = "fleet1",
  E           = optimal_E,
  tau         = 1L
)

# Inspect the stored embedding
get_edm_embeddings(ff) |> names()
```

## Step 3 — Select Nearest Neighbours (Simplex only)

For Simplex Projection, the number of nearest neighbours `n_neighbors`
controls the degree of local averaging. `select_n_neighbors()` evaluates
`k = E+1` to `k = min(n_library/2, E+10)` and identifies the best value.

```{r select-k}
skill_k <- select_n_neighbors(ff)
print(skill_k)

optimal_k <- skill_k$n_neighbors[skill_k$optimal]
cat("Optimal n_neighbors:", optimal_k, "\n")
```

```{r plot-k, fig.alt="Prediction skill vs number of nearest neighbours"}
ggplot(skill_k, aes(x = n_neighbors, y = rho, colour = optimal)) +
  geom_line(colour = "grey60") +
  geom_point(size = 3) +
  scale_colour_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick")) +
  labs(title = "Simplex prediction skill vs n_neighbors",
       x = "Number of nearest neighbours (k)", y = expression(rho)) +
  theme_minimal() +
  theme(legend.position = "none")
```

Alternatively, `fit_edm()` can auto-select `n_neighbors` internally by
setting `auto_select = TRUE`.

## Step 4 — Fit an EDM Model

### Simplex Projection

```{r fit-simplex}
fit_s <- fit_edm(
  ff,
  method      = "simplex",
  n_neighbors = optimal_k
)
print(fit_s)
```

### S-Map

S-Map includes a nonlinearity parameter `theta`. Larger `theta` values weight
library points closer to the query state more heavily. A `theta` near zero
approximates a linear AR model.

```{r fit-smap}
fit_m <- fit_edm(
  ff,
  method = "smap",
  theta  = 2.0
)
print(fit_m)
```

### GP-EDM

GP-EDM fits a Gaussian process to the delay-embedding manifold, using MAP
optimisation to learn ARD inverse length-scales `phi`, signal variance
`sigma2`, and process noise `ve`. User-supplied initial values allow control
over the optimisation starting point.

```{r fit-gp}
fit_g <- fit_edm(
  ff,
  method         = "gp",
  gp_sigma2_init = 1.0,
  gp_ve_init     = 0.1
)
print(fit_g)

# Fitted GP hyperparameters
get_parameters(fit_g)[c("phi", "sigma2", "ve")]
```

## Step 5 — Evaluate Prediction Skill

`edm_skill()` computes ρ, RMSE, and MAE for any fitted `EDMFit` object
against the stored target values in the `FIMSFrame`.

```{r skill-compare}
skill_s <- edm_skill(fit_s, ff)
skill_m <- edm_skill(fit_m, ff)
skill_g <- edm_skill(fit_g, ff)

skill_tbl <- dplyr::bind_rows(skill_s, skill_m, skill_g)
print(skill_tbl)
```

```{r plot-skill, fig.alt="In-library prediction skill comparison across EDM methods"}
ggplot(skill_tbl, aes(x = method, y = rho, fill = method)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c(
    simplex = "steelblue", smap = "darkorange", gp = "forestgreen"
  )) +
  labs(title = "In-library prediction skill by EDM method",
       x = "Method", y = expression(rho)) +
  ylim(0, 1) +
  theme_minimal()
```

## Step 6 — Generate Forecasts

`edm_forecast()` generates out-of-sample empirical forecasts from any fitted
`EDMFit` object. For `n_ahead > 1`, each predicted value is iteratively
appended to the series and used as the query state for the next step.

```{r forecast}
fc_s <- edm_forecast(fit_s, ff, n_ahead = 5L)
fc_m <- edm_forecast(fit_m, ff, n_ahead = 5L)
fc_g <- edm_forecast(fit_g, ff, n_ahead = 5L)

forecasts <- dplyr::bind_rows(
  dplyr::mutate(fc_s, method = "simplex"),
  dplyr::mutate(fc_m, method = "smap"),
  dplyr::mutate(fc_g, method = "gp")
)
print(forecasts)
```

```{r plot-forecast, fig.alt="Five-step ahead EDM forecasts from three methods"}
# Observed catch series
obs <- dplyr::filter(get_data(ff),
                     type == "catch", fleet == "fleet1") |>
  dplyr::arrange(timing) |>
  dplyr::select(time = timing, value)

ggplot() +
  geom_line(data = obs, aes(x = time, y = value), colour = "grey40") +
  geom_point(data = obs, aes(x = time, y = value), colour = "grey40") +
  geom_line(data = forecasts,
            aes(x = time, y = forecast, colour = method, group = method),
            linetype = "dashed", linewidth = 1) +
  geom_point(data = forecasts,
             aes(x = time, y = forecast, colour = method), size = 2) +
  scale_colour_manual(values = c(
    simplex = "steelblue", smap = "darkorange", gp = "forestgreen"
  )) +
  labs(title = "Catch: observed series and 5-step EDM forecasts",
       x = "Year", y = "Catch", colour = "Method") +
  theme_minimal()
```

## Multivariate Embedding (Preview)

When multiple time series are available, `create_multivariate_edm_embedding()`
builds a joint k×E delay-embedding matrix. This is useful for visualising
cross-variable relationships and will support multivariate prediction in a
future FIMS release once the C++ `construct_multivariate()` interface is
merged.

```{r multivariate}
ff_mv <- create_multivariate_edm_embedding(
  ff,
  series_specs = list(
    catch = list(series_type = "catch", series_name = "fleet1"),
    index = list(series_type = "index", series_name = "survey1")
  ),
  E             = optimal_E,
  tau           = 1L,
  target_series = "catch"
)

# Retrieve as a named matrix
mat <- model_edm_matrix_multivariate(ff_mv, include_target = TRUE)
head(mat)
```

## Summary

The table below shows the complete set of EDM functions added in Phase 4 of
the GSoC 2026 EDM integration, together with their roles in the workflow.

| Function | Role |
|----------|------|
| `select_embedding_dimension()` | Find optimal `E` via Simplex sweep |
| `create_edm_embedding()` | Build and store univariate delay embedding |
| `select_n_neighbors()` | Find optimal `k` for Simplex via in-library CV |
| `fit_edm()` | Fit Simplex, S-Map, or GP-EDM model |
| `edm_skill()` | Compute ρ, RMSE, MAE for a fitted model |
| `edm_forecast()` | Generate iterative multi-step out-of-sample forecasts |
| `create_multivariate_edm_embedding()` | Build joint k×E embedding from multiple series |
| `model_edm_matrix_multivariate()` | Retrieve multivariate embedding matrix |

## References

* Sugihara, G. & May, R. (1990). Nonlinear forecasting as a way of
  distinguishing chaos from measurement error in time series. *Nature*,
  344, 734–741.
* Sugihara, G. et al. (2012). Detecting Causality in Complex Ecosystems.
  *Science*, 338(6106), 496–500.
* Ye, H. et al. (2015). Equation-free mechanistic ecosystem forecasting
  using empirical dynamic modeling. *PNAS*, 112(13), E1569–E1576.
