---
title: "Introducing FIMS input data"
output: github_document
vignette: >
  %\VignetteIndexEntry{Introducing FIMS input data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set local options so vignette output is clean and consistent
# (avoids color codes and subtle formatting in rendered docs)
withr::local_options(list(
  cli.num_colors = 1,
  crayon.enabled = FALSE,
  pillar.subtle = FALSE
))
library(FIMS)
```

## Other sources of information

This vignette is an expansion of the ["Data" section](https://noaa-fims.github.io/FIMS/articles/fims-demo.html#data) in the [FIMS overview vignette](https://noaa-fims.github.io/FIMS/articles/fims-demo.html). It provides more details about the data structure needed for a FIMS model and how to prepare your data for use in a FIMS model using the `FIMSFrame()` function.

The [FIMS cheatsheet](https://github.com/NOAA-FIMS/FIMS/tree/main/pkgdown/assets/fims-cheatsheet.pdf) includes a quick summary of the input data format.

The help page for the `FIMS::data_big` data set (https://noaa-fims.github.io/FIMS/reference/data_big.html or via `?FIMS::data_big`) provides detailed information about the input data format.

Note: this vignette does not include anything about model configuration, which is covered in the overview vignette linked above.

## Data format

The input data for a FIMS model is a single long data frame with the following columns: `r glue::glue_collapse(colnames(FIMS::data_big), sep = ", ", last = ", and ")`. Both the single table and long format are in contrast to the input format for some legacy stock assessment models where the data are provided in multiple tables, each with a different format. The SS3 data file, for instance, has separate tables for catch, indices, age compositions, length compositions, with wide tables for the composition data (each age or length bin has a different column). The use of a single, long table for data input used by FIMS makes it easier to summarize and filter the data across data types (e.g., by fleet, timing, age, etc.), as well as being better suited to the widely-used tidyverse collection of packages in R.

Two drawbacks of the long format are (1) some information is duplicated, and (2) the long format is harder to read and understand by looking at the raw data frame. To meet this second challenge, the FIMS team is working on functions to summarize, visualize, and check for errors in the input data.

The currently available values for the `type` column are `r glue::glue_collapse(glue::glue('\"{FIMS::fims_input_types}\"'), sep = ", ", last = ", and ")`. The "weight_at_age", "ageing_error", and "age_to_length_conversion" types are not currently treated as data in the sense that they are not included in the likelihood. They are used in the population dynamics calculations or, for "ageing_error", in the expected age compositions.

The individual data types are described in more detail below after introducing the example data set that is included in the package.

## `data_big`

A sample data frame for a catch-at-age model with both ages and lengths is stored in the package as `data_big`. This data set is based on data that was used in [Li et al. (2021)](https://www.doi.org/10.7755/FB.119.2-3.5) for the Model Comparison Project ([github site](https://github.com/NOAA-FIMS/Age_Structured_Stock_Assessment_Model_Comparison)). The length data have since been added [data-raw/data_big.R](https://github.com/NOAA-FIMS/FIMS/blob/main/data-raw/data_big.R) based on an age--length conversion matrix. To see how this example data frame was created, see the R script here [R/data_big.R](https://github.com/NOAA-FIMS/FIMS/blob/main/R/data_big.R).

## `FIMSFrame()`

`FIMSFrame()` must be used to create a data object that can be passed to a FIMS model. This function returns an object that has the `FIMSFrame` class. This class contains a data slot that stores the input data frame and several other slots that store summary information about the object, e.g., the fleet names, the number of age bins, etc. All of this information can be accessed using `get_*()` functions, e.g., `get_n_years()`, `get_fleets()`. Every fleet named in the data must have catch, index, or composition data. `FIMSFrame()` returns an error otherwise, so a misspelled fleet name in other rows does not add a fleet to the model.

When you execute `FIMSFrame()` on a long data frame, such as `data_big`, the function executes several error checks to help ensure that your data is ready for a FIMS model. For example, until the internal estimation of growth is possible, all catch-at-age models require weight-at-age data. If your data frame does not have weight-at-age data, then `FIMSFrame()` will error and return helpful information on what you are missing.

The output of `FIMSFrame()` also contains summary information about your data and has an internal plotting function. See the [FIMS overview vignette](https://noaa-fims.github.io/FIMS/articles/fims-demo.html) for more information.

```{r fims-frame}
data_4_model <- FIMSFrame(data_big)
data_4_model
```

## Catch (`type == "catch"`)

The catch data type contains the total catch in weight for each time step. FIMS does not yet include the option to model discards internally, so "catch" should account for all catch. The catch are not associated with an age or length so those columns should only contain `NA` values. The `unit` column should currently always be `mt`. The `uncertainty` column should be the standard deviation of the logged observations if you are using the lognormal distribution to fit your data. `log_Fmort` parameters for each fleet and time step with catch can be estimated to fit the model to the catch data, informed by the uncertainty values. Common practice is to set the uncertainty to a small number, like 0.001, so the catch data are fit closely.

```{r data_big-catch}
# first three rows of the catch data
FIMS::data_big |>
  dplyr::filter(type == "catch") |>
  head(3)
```

```{r data_big-catch-plot}
#| fig.alt: >
#|   Barplot shows an increase in catch over the timing values 1 to 8, then noisy but stable values until the final bar at timing equals 30.
# time series plot of catch data
library(ggplot2)
FIMS::data_big |>
  dplyr::filter(type == "catch") |>
  ggplot(aes(x = timing, y = observed)) +
  geom_bar(stat = "identity") +
  ggtitle("Catch time series in data_big example data set")
```

## Indices (`type == "index"`)

The index data type contains the relative or absolute abundance index for each time step. The inputs are similar to catch, in having `NA` values for age and length, `mt` for `unit`, and index uncertainty as the standard deviation of the logged observations.

```{r data_big-indices}
# first three rows of the index data
FIMS::data_big |>
  dplyr::filter(type == "index") |>
  head(3)
```

```{r data_big-indices-plot}
#| fig.alt: >
#|   Points with uncertainty shows a steadily decreasing index over time steps 1 to 30 with increasing precision for the later observations.
# time series plot of index data
FIMS::data_big |>
  dplyr::filter(type == "index") |>
  ggplot2::ggplot(ggplot2::aes(x = timing, y = observed)) +
  ggplot2::geom_point() +
  ggplot2::geom_pointrange(ggplot2::aes(
    ymin = stats::qlnorm(
      p = 0.025,
      meanlog = log(observed),
      sdlog = unlist(
        FIMS:::parse_data_distribution(uncertainty)$sdlog,
        use.names = FALSE
      )
    ),
    ymax = stats::qlnorm(
      p = 0.975,
      meanlog = log(observed),
      sdlog = unlist(
        FIMS:::parse_data_distribution(uncertainty)$sdlog,
        use.names = FALSE
      )
    )
  )) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::ggtitle("Index time series in data_big example data set")
```

## Age compositions (`type == "age_comp"`)

The age-composition data type contains the age distribution of a fleet for a time step. The `age` column contains integer ages defining the age bins. While standard FIMS models typically assume age bins start at zero, the data_big example currently uses ages ranging from 1--12, and thus, recruitment will be age-1 fish not age-0 fish. The `observed` column contains either the number or the proportion of the catch or survey in each age bin. The `unit` column should be either "number" or "proportion". If "proportion" is used, the observations are checked at a later step to confirm that they sum to 1.0. The `uncertainty` column should be the input sample size for the time step (equal across all ages within that time step).

```{r data_big-age_comp}
# first three rows of the age composition data
FIMS::data_big |>
  dplyr::filter(type == "age_comp") |>
  head(3)
```
```{r data_big-age_comp-plot}
#| fig.alt: >
#|   Barplot showing age comp by time step with a larger proportion of older ages (including a significant plus group at age 12 for the early time steps) and a truncated age distribution in the final time steps with relatively few fish over age 7.
# plot of age compositions by time step
FIMS::data_big |>
  dplyr::filter(type == "age_comp") |>
  ggplot2::ggplot(
    ggplot2::aes(x = age, y = observed)
  ) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::xlab("Age") +
  ggplot2::ylab("Number") +
  ggplot2::facet_wrap(~timing) +
  ggplot2::scale_x_continuous(breaks = get_ages(data_4_model)) +
  ggtitle("Age compositions by time step in data_big example data set")
```

## Length compositions (`type == "length_comp"`)

The length-composition data are similar to the age compositions, except with values in the `length` column and `age = NA`. The `length` column identifies the length bin for each observation. The `observed` column contains either the number or the proportion of fish in each length bin. The `unit` column should be either "number" or "proportion". The `uncertainty` column should be the input sample size for the time step, repeated across all length bins in that time step.

When length compositions are included, FIMS needs a way to convert numbers at age into numbers at length. This can come from fixed `age_to_length_conversion` rows or from a growth-derived age-to-length conversion using the VonB-Schnute Growth module.

```{r data_big-length_comp}
# first three rows of the length composition data
FIMS::data_big |>
  dplyr::filter(type == "length_comp") |>
  head(3)
```
```{r data_big-length_comp-plot}
#| fig.alt: >
#|   Barplot showing length comp by time step with a mode around 450 mm which declines and becomes more sharply peaked over time.
# plot of length compositions by time step
FIMS::data_big |>
  dplyr::filter(type == "length_comp") |>
  ggplot2::ggplot(
    ggplot2::aes(x = length, y = observed)
  ) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::xlab("Length") +
  ggplot2::ylab("Number") +
  ggplot2::facet_wrap(~timing) +
  ggtitle("Length compositions by time step in data_big example data set")
```

## Length bins (`type == "length_bin"`)

The `length_bin` data type defines length bins without providing length-composition observations. These rows are useful when a model needs length-bin geometry for a growth-derived age-to-length conversion, but the user is not fitting length-composition data for that fleet.

For `length_bin` rows, the `length` column contains the length-bin values. The `observed`, `age`, and `uncertainty` columns should usually be `NA`. The `fleet` column identifies which fleet or survey the length bins apply to, and `timing` can be `NA` unless fleet-specific timing is needed.

## Weight-at-age data (`type == "weight_at_age"`)

The weight-at-age data are not currently included in the likelihood and are just used for converting from numbers to biomass in the population dynamics. The `unit` column should be `mt` and the uncertainty should be `NA`.

Two key features of the weight-at-age data are 

  1. these data are required for the one additional time step (final observation for timing + 1), and
  2. the observations are currently shared among fleets and throughout the dynamics (so only need to be provided once).

The final timing + 1 observations are required because FIMS reports spawning biomass at the beginning of the year for every year, including the year after the last catches are removed from the population.

The weight-at-age observations for the `data_big` are identical across time steps, and are provided for timing = 1 to 31 and associated with name = "fleet1".

```{r data_big-weight_at_age}
# first and last entries  of the weight-at-age data
FIMS::data_big |>
  dplyr::filter(type == "weight_at_age") |>
  head(3)
FIMS::data_big |>
  dplyr::filter(type == "weight_at_age") |>
  tail(3)

# check that the observations are identical across time steps within each age
FIMS::data_big |>
  dplyr::filter(type == "weight_at_age") |>
  dplyr::group_by(age) |>
  dplyr::summarize(n_distinct = dplyr::n_distinct(observed))
```

```{r data_big-weight_at_age-plot}
#| fig.alt: >
#|   Line plot showing weight at age showing identical monotonically increasing weight-at-age observations ages that are the same across time steps.
# plot of weight-at-age by time step
FIMS::data_big |>
  dplyr::filter(type == "weight_at_age") |>
  ggplot2::ggplot(
    ggplot2::aes(x = age, y = observed)
  ) +
  ggplot2::geom_line(color = "blue") +
  ggplot2::xlab("Age") +
  ggplot2::ylab("Weight (mt)") +
  ggplot2::facet_wrap(~timing) +
  ggplot2::scale_x_continuous(breaks = get_ages(data_4_model)) +
  geom_hline(yintercept = 0) +
  ggtitle("Weight-at-age by time step in data_big example data set")
```

## Ageing error (`type == "ageing_error"`)

Ageing error data are optional. Age readers do not always assign the true age, which spreads a strong year class across neighboring ages. With ageing error rows, each fleet's expected age composition is moved from true age to observed age before it is compared to the data. Length compositions, catch, and indices still use true age, and ageing error is not part of the likelihood.

Each row gives the probability that a fish of one true age is read as one observed age. The observed age goes in `age`, the true age in `uncertainty`, and the probability in `observed`, with `unit = "proportion"`. The `uncertainty` column holds formulas for other data types, so true ages are stored as text, e.g., `uncertainty = as.character(true_age)`.

Each set of rows is a full matrix over the model ages, and the probabilities for each true age must sum to 1. Sums within 1e-3 of 1 are rescaled to exactly 1, with a warning, and larger differences are an error. If a matrix was rounded to fewer than 3 decimal places, first check that `uncertainty` holds the true age and `age` the observed age, then divide each true age's probabilities by their total before adding it. Probability on observed ages below the youngest model age is added to that age, and above the oldest model age to the plus group, with a warning. Every observed age needs a total probability, summed across true ages, of at least 1e-3, because an expected proportion of 0 makes the likelihood undefined.

Every model age needs rows as a true age, in `uncertainty`. Rows for true ages outside the model ages are not used, with a warning. Each row is the probability of each observed age given one true age, so probabilities for observed ages within a row can be added, but combining rows for different true ages needs the number of fish at each true age, which is unknown. The model has no fish below the youngest age, e.g., true age 0 in a model that starts at age 1, so dropping those rows changes nothing. Above the plus group, the row for the plus-group true age is used for every fish in the plus group.

Ageing error can vary by fleet and year. For each fleet and year, the most specific rows are used:

  1. rows for that fleet and that `timing`,
  2. rows for that fleet with `timing = NA`,
  3. rows with `fleet = NA` and that `timing`, and
  4. rows with `fleet = NA` and `timing = NA`.

Rows with `fleet = NA` apply to every fleet with age-composition data. A `timing` applies to that year only. Without a `timing = NA` default, every model year needs its own rows, including projection years. A fleet with no rows that apply to it has no ageing error, which is also the default.

With ageing error, the reported `agecomp_proportion` and `agecomp_expected` are at observed age, while `catch_numbers_at_age` and `index_numbers_at_age` stay at true age.

`data_big` does not include ageing error. The example below adds a matrix where each age is read 1 year too young or too old with probability 0.1.

```{r data_big-ageing_error}
ageing_error_ages <- get_ages(data_4_model)
n_ageing_error_ages <- length(ageing_error_ages)
ageing_error_matrix <- diag(0.8, n_ageing_error_ages)
ageing_error_matrix[cbind(
  1:(n_ageing_error_ages - 1),
  2:n_ageing_error_ages
)] <- 0.1
ageing_error_matrix[cbind(
  2:n_ageing_error_ages,
  1:(n_ageing_error_ages - 1)
)] <- 0.1
ageing_error_matrix[1, 1] <- 0.9
ageing_error_matrix[n_ageing_error_ages, n_ageing_error_ages] <- 0.9
ageing_error <- tibble::tibble(
  type = "ageing_error",
  fleet = NA_character_,
  age = rep(ageing_error_ages, times = n_ageing_error_ages),
  length = NA_real_,
  timing = NA_real_,
  # Rows of the matrix are true ages, so t() lists observed ages within each
  # true age.
  observed = as.vector(t(ageing_error_matrix)),
  unit = "proportion",
  uncertainty = as.character(rep(ageing_error_ages, each = n_ageing_error_ages))
)
data_with_ageing_error <- FIMSFrame(dplyr::bind_rows(data_big, ageing_error))
get_data(data_with_ageing_error) |>
  dplyr::filter(type == "ageing_error") |>
  head(3)
```

[Stock Synthesis](https://nmfs-ost.github.io/ss3-website/) and the [AgeingError package](https://pfmc-assessments.github.io/AgeingError/) describe ageing error by the mean and standard deviation of observed age at each true age, from age 0 to the population maximum age. In that convention an observed age covers the year from that age to the next, so unbiased readers have a mean of true age + 0.5 (entered as -1 in Stock Synthesis), and a standard deviation of 0.001 means almost no ageing error. The probability of each observed age is the normal probability over its year. The first and last observed ages collect the tails, as FIMS does when it folds observed ages outside the model ages, so each row sums to 1. The example below uses this convention, so vectors from Stock Synthesis or the AgeingError package can be used directly.

```{r ageing_error-from-mean-sd}
# Rows are true ages and columns are observed ages. Here readers are unbiased
# and the standard deviation grows with age.
n_ages <- n_ageing_error_ages
mean_observed_age <- matrix(ageing_error_ages + 0.5, n_ages, n_ages)
sd_observed_age <- matrix(0.1 * ageing_error_ages, n_ages, n_ages)
observed_age <- matrix(ageing_error_ages, n_ages, n_ages, byrow = TRUE)
# The first observed age starts at -Inf and the last ends at Inf so they
# collect the tails.
lower_bound <- observed_age
lower_bound[, 1] <- -Inf
upper_bound <- observed_age + 1
upper_bound[, n_ages] <- Inf
ageing_error_matrix <-
  stats::pnorm(upper_bound, mean_observed_age, sd_observed_age) -
  stats::pnorm(lower_bound, mean_observed_age, sd_observed_age)
round(ageing_error_matrix[1:4, 1:4], 3)
```

The [AgeingError package](https://pfmc-assessments.github.io/AgeingError/) estimates ageing error from double reads and returns it as `ageing_error` rows. `AgeingError::run()` stores 1 table per reader in `ageing_error_fims`, and `AgeingError::ageing_error_matrix_to_fims()` converts any matrix with true ages as rows. Bind the table for 1 reader, not `ageing_error_fims` itself, which holds every reader. The tables use `fleet = NA` and `timing = NA`, so to use different readers for different fleets or years, set `fleet` or `timing` in each table first. The tables start at true age 0 and run to the oldest age read, so FIMS drops true ages outside the model ages and adds observed ages outside them to the end ages, with warnings.

```{r ageing_error-from-AgeingError, eval = FALSE}
# out is the result of AgeingError::run() on double-read data.
reader1 <- out[["output"]][["ageing_error_fims"]][["reader1"]]
reader2 <- out[["output"]][["ageing_error_fims"]][["reader2"]] |>
  dplyr::mutate(fleet = "survey1")
data_with_ageing_error <- FIMSFrame(
  dplyr::bind_rows(data_big, reader1, reader2)
)
```

## Age-to-length conversion (`type == "age_to_length_conversion"`)

The age-to-length conversion data type contains the distribution of lengths for each age. The `age` and `length` columns contain all the combinations of age and length bins, and the `observed` column contains the proportion of fish at each age that are in each length bin. The `unit` column should be "proportion". A single set of the age-to-length conversion data is required, rather than separate copies for each fleet (name) or time step. The age-to-length conversion data are not currently included in the likelihood but are used to convert numbers at age to numbers at length for the population dynamics calculations when length compositions are included in the model.

Note: purely age-based models which have no length-composition data can omit the age-to-length conversion data as well.

```{r data_big-age_to_length_conversion}
# first and last entries in the age-to-length conversion data type
FIMS::data_big |>
  dplyr::filter(type == "age_to_length_conversion") |>
  head(3)
FIMS::data_big |>
  dplyr::filter(type == "age_to_length_conversion") |>
  tail(3)

# show that the number of rows in the age-to-length conversion data type is equal to the product of the number of unique combinations of age and length
FIMS::data_big |>
  dplyr::filter(type == "age_to_length_conversion") |>
  nrow()
# 12 ages, 23 length bins:
length(unique(na.omit(FIMS::data_big$age))) * length(unique(na.omit(FIMS::data_big$length)))
```

```{r data_big-age_to_length_conversion-plot}
#| fig.alt: >
#|   Scatter plot showing the age-to-length conversion with point size representing the proportion of fish at each age-length combination.
# plot of age-to-length conversion observations for the data_big example data set
FIMS::data_big |>
  dplyr::filter(type == "age_to_length_conversion") |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = age,
      y = length,
      size = observed
    )
  ) +
  ggplot2::geom_point(color = gray(0, alpha = 0.3)) +
  ggplot2::scale_size_continuous(range = c(0.1, 12)) +
  ggplot2::xlab("Age") +
  ggplot2::ylab("Length (mm)") +
  ggplot2::scale_x_continuous(breaks = get_ages(data_4_model)) +
  geom_hline(yintercept = 0) +
  ggtitle("Age-to-length conversion for the data_big example data set")
```
