---
title: "Using the Likelihood Module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Likelihood Module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The likelihood module provides explicit likelihood components that can be
connected to data, model parameters, random effects, derived quantities, or
penalty terms. It lives alongside the legacy distribution module, but uses a
more direct object model for linking inputs and expected values.

The available likelihood classes are:

- `NormalLikelihood`
- `LognormalLikelihood`
- `GammaLikelihood`
- `InvGammaLikelihood`
- `MultinomialLikelihood`

All likelihood classes inherit from `LikelihoodBase` and are exposed through
the FIMS Rcpp module.

# Setting Values

The likelihood interface uses `RealVector` objects for fixed numeric values.
Use whole-vector replacement to fill `RealVector` and `ParameterVector`
objects from R.

```{r vector-replacement, eval = FALSE}
real_values <- methods::new(RealVector)
real_values[] <- c(1, 2, 3)

parameter_values <- methods::new(ParameterVector)
parameter_values[] <- c(1, 2, 3)
```

# Direct Evaluation

Likelihoods can be evaluated directly from R. Direct evaluation is useful for
unit tests and for checking that a likelihood has been parameterized correctly
before adding it to a full model.

```{r direct-normal, eval = FALSE}
library(FIMS)

normal_likelihood <- methods::new(NormalLikelihood)

normal_likelihood$observed_values[] <- c(1, 2, 3)
normal_likelihood$expected_values[] <- c(1.1, 1.9, 3.2)
normal_likelihood$log_sd[] <- log(0.5)

normal_likelihood$evaluate()
normal_likelihood$nll_components$toRVector()

clear()
```

The normal likelihood evaluates the negative log likelihood equivalent to:

```{r normal-equivalent, eval = FALSE}
-sum(stats::dnorm(
  x = c(1, 2, 3),
  mean = c(1.1, 1.9, 3.2),
  sd = exp(log(0.5)),
  log = TRUE
))
```

# Linking Inputs

Likelihood objects can own their observed values directly, or they can link to
another FIMS vector. Linked inputs are useful when the input is a model
parameter, a random effect, or a derived quantity created by another module.

Use `set_real_input()` for a `RealVector` input.

```{r linked-real-input, eval = FALSE}
input <- methods::new(RealVector)
input[] <- c(0.1, -0.2, 0.3)

penalty <- methods::new(NormalLikelihood)
penalty$expected_values[] <- 0
penalty$log_sd[] <- log(1)
penalty$set_real_input(input, "penalty")
```

Use `set_parameter_input()` for a `ParameterVector` input. This is the usual
pattern for priors and random effects.

```{r linked-parameter-input, eval = FALSE}
selectivity <- methods::new(LogisticSelectivity)
selectivity$inflection_point[1]$value <- 1.75
selectivity$inflection_point[1]$estimation_type$set("fixed_effects")

prior <- methods::new(NormalLikelihood)
prior$expected_values[] <- 1.75
prior$log_sd[] <- log(3)
prior$set_parameter_input(selectivity$inflection_point, "prior")
```

The role argument can be one of:

- `"data"`
- `"prior"`
- `"random_effect"`, `"random_effects"`, or `"re"`
- `"penalty"`

# Linking Expected Values

Expected values can also be owned by the likelihood or linked from another
module. Linking expected values is important for catch-at-age models because
expected landings, expected index values, and composition probabilities are
usually derived quantities.

Use `set_real_expected_input()` to link a `RealVector` expected value and
`set_parameter_expected_input()` to link a `ParameterVector` expected value.

```{r linked-expected-values, eval = FALSE}
landings_likelihood <- methods::new(LognormalLikelihood)

landings_likelihood$observed_values[] <- observed_landings
landings_likelihood$log_sd[] <- log_landings_sd

landings_likelihood$set_parameter_expected_input(
  fishing_fleet$log_landings_expected
)
```

For multinomial likelihoods, `observed_values` and `expected_values` are stored
as flat vectors, and `dims` stores the row and column dimensions.

```{r multinomial, eval = FALSE}
agecomp_likelihood <- methods::new(MultinomialLikelihood)

agecomp_likelihood$observed_values[] <- agecomp_counts
agecomp_likelihood$dims[] <- c(n_years, n_ages)

agecomp_likelihood$set_parameter_expected_input(
  fishing_fleet$agecomp_proportion
)
```

# Derived Quantity Example

In a catch-at-age model, many likelihood expected values are not fixed inputs.
They are derived during model evaluation after the population and fleet modules
have been linked. The Rcpp fleet object exposes handles to those derived
quantities, and `CreateTMBModel()` connects those handles to the backend
derived vectors through `Information::variable_map`.

For example, the expected log landings for a fishing fleet are calculated by
the catch-at-age model and stored in the fleet derived quantity
`log_landings_expected`. The likelihood owns the observed landings and
observation error, but links its expected values to that derived quantity.

```{r derived-log-landings, eval = FALSE}
fishing_fleet <- methods::new(Fleet)
fishing_fleet$n_years$set(n_years)
fishing_fleet$n_ages$set(n_ages)
fishing_fleet$SetSelectivityID(fishing_selectivity$get_id())
fishing_fleet$SetObservedLandingsDataID(fishing_landings$get_id())

landings_likelihood <- methods::new(LognormalLikelihood)
landings_likelihood$observed_values[] <- observed_landings
landings_likelihood$log_sd[] <- log_landings_sd

landings_likelihood$set_parameter_expected_input(
  fishing_fleet$log_landings_expected
)
```

The same pattern works for composition likelihoods. Here the expected
age-composition probabilities are derived by the model, while observed counts
and dimensions are owned by the likelihood.

```{r derived-agecomp, eval = FALSE}
agecomp_likelihood <- methods::new(MultinomialLikelihood)

agecomp_likelihood$observed_values[] <- agecomp_counts
agecomp_likelihood$dims[] <- c(n_years, n_ages)

agecomp_likelihood$set_parameter_expected_input(
  fishing_fleet$agecomp_proportion
)
```

The likelihood can be created before the backend derived vector exists. It
stores the frontend id from `fishing_fleet$log_landings_expected` or
`fishing_fleet$agecomp_proportion`; then `CreateTMBModel()` initializes the
derived quantities and resolves that id to the backend vector.

# Priors And Random Effects

Priors and random effects are represented by ordinary likelihood objects with
different roles. The role tells FIMS how the component should be reported and
how its input is interpreted by the backend model.

```{r recruitment-random-effects, eval = FALSE}
recruitment_likelihood <- methods::new(NormalLikelihood)

recruitment_likelihood$expected_values[] <- 0
recruitment_likelihood$log_sd[] <- log(recruitment_sd)

recruitment_likelihood$set_parameter_input(
  recruitment$log_devs,
  "random_effects"
)
```

This pattern keeps priors, random effects, data likelihoods, and penalties in
one common system. It also avoids duplicating distribution-specific linking
code across model modules.

# Model Integration

When `CreateTMBModel()` is called, each live likelihood object is added to the
backend model. Linked frontend vector ids are resolved through the backend
`Information::variable_map`.

The basic workflow is:

1. Create the model modules and set their parameter values.
2. Create likelihood objects and link their inputs or expected values.
3. Call `CreateTMBModel()`.
4. Use `CreateTMBModelParameterList()` and `CreateTMBModelRandomEffects()` to
   build the inputs for `TMB::MakeADFun()`.

```{r tmb-workflow, eval = FALSE}
CreateTMBModel()

parameters <- CreateTMBModelParameterList()
random_effects <- CreateTMBModelRandomEffects()

obj <- TMB::MakeADFun(
  data = list(),
  parameters = parameters,
  DLL = "FIMS",
  random = random_effects
)
```

Likelihood components are evaluated after the legacy distribution components
and contribute to the model's joint negative log likelihood. Their
observation-level contributions are stored in `nll_components`, which can be
used in tests and reports.
