---
title: "Running a retrospective analysis on a FIMS model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running a retrospective analysis on a FIMS model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 7,
  out.width = "100%"
)
```

```{r setup}
library(FIMS)
packageVersion("FIMS")
clear()
```


## Set up FIMS model

Following the [minimal FIMS demo vignette](https://noaa-fims.github.io/FIMS/articles/fims-demo-minimal.html), we will set up a FIMS model and run it. This model will contain the all years of data and will be our reference model (see section below for more information about reference models). For this example we are using the dataset included in the FIMS package and creating a set of default parameters. 

```{r run_fims_model}
# Load sample data
data("data_big")
# Prepare data for FIMS model
data_4_model <- FIMSFrame(data_big)

# Create parameters
parameters <- data_4_model |>
  create_default_configurations() |>
  create_default_parameters(data = data_4_model)

# Run the  model with optimization
base_model <- parameters |>
  initialize_fims(data = data_4_model) |>
  fit_fims(optimize = TRUE)

# Clear memory post-run
clear()
```

## Retrospective analysis

A retrospective analysis is a common model diagnostic used to assess the stability of estimates of parameters and derived quantities, such as fishing mortality and spawning biomass, and if there is a consistent pattern in the estimates as years of data are removed. The process involves removing *n* years of data from the end of the time series and refitting the model. This is an iterative process where the user typically runs 5--10 models, with each run (often called "peels") progressively removing another year of data. Estimates from these peels are then compared to the model run with the full dataset (reference model). The Mohn's rho statistic can be used to compare the relative difference between the reference model and each peel. Mohn's rho statistic is calculated as: 

$$\rho = \frac{1}{n} \sum_{i=1}^{n} \frac{x_{p,y} - x_{base,y}}{x_{base,y}}$$

where $n$ is the number of retrospective peels, $x_{p, y}$ is the estimated value (e.g., spawning biomass, fishing mortality, etc.) from peel $p$ at its terminal year $y$, and $x_{base,y}$ is the value from the reference model, $base$, at that same year $y$.

Some things to note about `run_fims_retrospective()` is that you can provide data that is already a `FIMSFrame` object or you can give it one that is not. In the example below, we are using the `FIMSFrame` data object (`data_4_model`) created above. Additionally, it expects a vector of positive values of `years_to_remove` from the base model, starting from 0. If you are running many peels, it may be more efficient to use multiple cores, which can be specified with the `n_cores` argument. 

> NOTE: If left empty, the default for `n_cores` is one less than the number of cores on your machine so this could unexpectedly use up a lot of your computing power. 

```{r, retro}
retro_fit <- run_fims_retrospective(
  years_to_remove = 0:2,
  data_4_model,
  parameters,
  n_cores = 1
)
```

## Mohn's Rho

Once you have run a retrospective, you can then calculate the Mohn's rho statistic for spawning biomass by: 

```{r, mohns_rho}
rho_sb <- calculate_mohns_rho(retro_fit, quantity = "spawning_biomass")
```

## Visualize the results

To visualize the results of the retrospective analysis, we can use `plot_retrospective()` to display the quantity of interest. Below, we want to assess the impact on spawning biomass. The function returns a `ggplot` object, which you can customize by adding add layers as you normally would to a `ggplot`. For example, below, we specify the x and y axes labels and the legend title.

```{r, plot-spawning-biomass, message = FALSE, warning = FALSE}
#| fig.alt: >
#|   Plot of spawning biomass showing a small negative bias with removed years.
fig_retrospective_ssb <- stockplotr::plot_spawning_biomass(
  retro_fit$estimates |>
    dplyr::rename(
      year = year_i,
      estimate = estimated
    ) |>
    dplyr::mutate(
      uncertainty_label = "se"
    ),
  group = "retrospective_peel"
)
print(fig_retrospective_ssb)
```

We can compare the spawning biomass estimates from the reference model and each peel for the last six years. 
```{r}
retro_fit[["estimates"]] |>
  dplyr::filter(label == "spawning_biomass") |>
  dplyr::select(label, year_i, estimated, retrospective_peel) |>
  tidyr::pivot_wider(
    names_from = retrospective_peel,
    values_from = estimated
  ) |>
  dplyr::rename(
    "Year" = year_i,
    "Reference model" = `0`,
    "Peel 1" = `1`,
    "Peel 2" = `2`
  ) |>
  tail()
```

## Clean up

Once you are finished running a FIMS model, it is always good practice to clear the C++ memory from one FIMS model run to the next.
```{r}
clear()
```
