---
title: "Things To Try With Von Bertalanffy Growth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Things To Try With Von Bertalanffy Growth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Purpose

This vignette walks through small changes to the Von Bertalanffy growth example.
Each section changes one part of the setup, reruns the full model, and saves the
outputs for comparison.

The goal is to show how growth inputs move through the full model workflow.

The examples focus on:

- starting values for estimable growth parameters,
- fleet length-bin spacing,
- length-at-age variability,
- the second reference age in the two-reference-age growth curve,
- reporting the full realized growth-derived ALK tensor,
- downstream outputs that respond to the growth-derived pathway.


## Shared Setup

This companion vignette uses the same generated-data helper functions as the
main Von Bertalanffy growth vignette.

It assumes these helper functions have already been defined earlier in the
session:

- `make_vonb_demo_data()`
- `vonb_two_reference_age()`
- `vonb_implied_linf()`

If you want to knit this file on its own, copy those helper definitions into
this vignette before the setup chunk below.

```{r setup}
library(FIMS)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stockplotr)

core_growth_labels <- c(
  "length_at_ref_age_1",
  "length_at_ref_age_2",
  "growth_coefficient_K"
)
```

## Bring Your Own Growth Data

### What users can do right now

Right now, the Von Bertalanffy growth parameters can be informed by your fleet
length-composition data.

In this workflow, users do not yet pass a separate table of raw age-length
observations into a direct growth likelihood. Instead, growth is estimated as
part of the full model while the model tries to match the length-composition
data and the rest of the assessment inputs.

### What to bring

If you want to estimate Von Bertalanffy growth with your own data, the most
important pieces are:

- `length_comp` rows for the fleets and years you want the growth path to help explain,
- `length_bin` rows if you want to control the fleet observation bins explicitly,
- `weight_at_age` rows because the current FIMS user path still requires them,
- the usual supporting data for the rest of the model, such as landings, index,
  and age compositions when those are part of the assessment.

### If you start from raw age-length samples

If your starting point is a raw table of fish ages and lengths, a simple
workflow is:

1. summarize those observations into fleet-year `length_comp` inputs,
2. use the raw age-length data outside FIMS to choose reasonable starting
   values for the growth curve and the SD-at-age anchors,
3. fit the full FIMS model with `Growth = "VonBertalanffy"` and review the
   fitted growth estimates.

### Minimal setup pattern

```{r byog-growth-template, eval = FALSE}
my_frame <- FIMS::FIMSFrame(my_long_data)

my_configurations <- FIMS::create_default_configurations(data = my_frame) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

my_parameters <- FIMS::create_default_parameters(
  configurations = my_configurations,
  data = my_frame
) |>
  tidyr::unnest(cols = data)

# Then update the key Growth values to match your stock as closely as you can:
# - length_at_ref_age_1
# - length_at_ref_age_2
# - growth_coefficient_K
# - reference_age_for_length_1
# - reference_age_for_length_2
# - length_at_age_sd_at_ref_ages
```

### What this means for the examples below

The examples in this vignette are generated-data checks, but the same basic
workflow is what a user would follow with real data: prepare the long FIMS
table, switch Growth to `VonBertalanffy`, set reasonable starting values, and
then see how the fitted growth results respond.

## Baseline Reference Run

### What this run does

The baseline run uses the generated data and model settings from the main
Von Bertalanffy growth vignette. The sections that follow change one piece of
this setup at a time.

### Run the baseline model

```{r baseline-run}
# Start fresh.
FIMS::clear()

# Create the generated data using the baseline settings.
baseline_demo <- make_vonb_demo_data()

# Replace the default empirical weight-at-age Growth module with
# Von Bertalanffy growth.
baseline_configurations <- FIMS::create_default_configurations(
  data = baseline_demo$frame
) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

# Create the parameter table.
baseline_parameters <- FIMS::create_default_parameters(
  configurations = baseline_configurations,
  data = baseline_demo$frame
) |>
  tidyr::unnest(cols = data)

# Pull out the default VonB SD anchor rows.
baseline_growth_sd_rows <- baseline_parameters |>
  dplyr::filter(
    module_name == "Growth",
    label == "length_at_age_sd_at_ref_ages"
  ) |>
  dplyr::mutate(
    # Move the interpolation anchors to the ages used in this demo.
    age = c(1, 12),
    # Match the generated-truth SD values for the example.
    value = c(30, 73)
  )

# Replace the generic defaults with the demo-specific growth settings.
baseline_parameters <- baseline_parameters |>
  dplyr::filter(
    !(module_name == "Growth" &
        label == "length_at_age_sd_at_ref_ages")
  ) |>
  dplyr::bind_rows(baseline_growth_sd_rows) |>
  dplyr::rows_update(
    tibble::tibble(
      module_name = "Growth",
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      value = c(150, 800, 0.22, 1, 12)
    ),
    by = c("module_name", "label")
  )

# Initialize and fit the model.
baseline_input <- baseline_parameters |>
  FIMS::initialize_fims(data = baseline_demo$frame)

# Request the full realized ALK tensor for the comparison plots.
baseline_input$model$ReportGrowthDerivedALKTensor(TRUE)

baseline_fit <- FIMS::fit_fims(
  input = baseline_input,
  optimize = TRUE,
  number_of_loops = 5,
  number_of_newton_steps = 0,
  get_sd = FALSE
)

baseline_growth_estimates <- FIMS::get_estimates(baseline_fit) |>
  dplyr::filter(
    module_name == "Growth",
    label %in% core_growth_labels
  )

# Pull the report only after the fitted growth estimates are in hand.
baseline_report <- FIMS::get_report(baseline_fit)

baseline_growth_comparison <- baseline_growth_estimates |>
  dplyr::select(label, input, estimated) |>
  dplyr::left_join(
    baseline_demo$true_parameters |>
      dplyr::filter(label %in% core_growth_labels),
    by = "label"
  )

baseline_growth_comparison

baseline_diagnostics <- tibble::tibble(
  scenario = "baseline",
  convergence = FIMS::get_opt(baseline_fit)$convergence,
  max_gradient = FIMS::get_max_gradient(baseline_fit)
)

baseline_diagnostics
```

## Try 1: Change Growth Starting Values

### What this changes

This run changes only the starting values for the three estimable
Von Bertalanffy growth parameters.

### Why this matters

If the growth parameters are connected to the objective and the data are
informative, the model should move away from the changed starting values and
toward the generated truth.

### Run the model

```{r try-1-starting-values}
FIMS::clear()

starting_values_demo <- make_vonb_demo_data()

starting_values_configurations <- FIMS::create_default_configurations(
  data = starting_values_demo$frame
) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

starting_values_parameters <- FIMS::create_default_parameters(
  configurations = starting_values_configurations,
  data = starting_values_demo$frame
) |>
  tidyr::unnest(cols = data)

starting_values_growth_sd_rows <- starting_values_parameters |>
  dplyr::filter(
    module_name == "Growth",
    label == "length_at_age_sd_at_ref_ages"
  ) |>
  dplyr::mutate(
    age = c(1, 12),
    value = c(30, 73)
  )

# CHANGED: starting values are farther from the generated truth.
starting_values_parameters <- starting_values_parameters |>
  dplyr::filter(
    !(module_name == "Growth" &
        label == "length_at_age_sd_at_ref_ages")
  ) |>
  dplyr::bind_rows(starting_values_growth_sd_rows) |>
  dplyr::rows_update(
    tibble::tibble(
      module_name = "Growth",
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      value = c(100, 700, 0.35, 1, 12)
    ),
    by = c("module_name", "label")
  )

starting_values_input <- starting_values_parameters |>
  FIMS::initialize_fims(data = starting_values_demo$frame)

starting_values_input$model$ReportGrowthDerivedALKTensor(TRUE)

starting_values_fit <- FIMS::fit_fims(
  input = starting_values_input,
  optimize = TRUE,
  number_of_loops = 5,
  number_of_newton_steps = 0,
  get_sd = FALSE
)

starting_values_growth_estimates <- FIMS::get_estimates(starting_values_fit) |>
  dplyr::filter(
    module_name == "Growth",
    label %in% core_growth_labels
  )

# Pull the report only after the fitted growth estimates are in hand.
starting_values_report <- FIMS::get_report(starting_values_fit)

starting_values_growth_comparison <- starting_values_growth_estimates |>
  dplyr::select(label, input, estimated) |>
  dplyr::left_join(
    starting_values_demo$true_parameters |>
      dplyr::filter(label %in% core_growth_labels),
    by = "label"
  )

starting_values_growth_comparison

starting_values_diagnostics <- tibble::tibble(
  scenario = "changed starting values",
  convergence = FIMS::get_opt(starting_values_fit)$convergence,
  max_gradient = FIMS::get_max_gradient(starting_values_fit)
)

starting_values_diagnostics
```

### What should happen

The `input` column should show the changed starting values. The estimated
values should move toward the generated truth.

## Try 2: Change Fleet Length Bins

### What this changes

This run changes the fleet observation-bin spacing from 50 to 25.

### Why this matters

The mean growth curve should be similar, but the realized growth-derived ALK
should be mapped onto a finer fleet observation grid.

### Run the model

```{r try-2-finer-length-bins}
FIMS::clear()

# CHANGED: use a finer fleet observation-bin grid.
finer_bins_demo <- make_vonb_demo_data(
  lengths = seq(0, 1400, 25)
)

finer_bins_configurations <- FIMS::create_default_configurations(
  data = finer_bins_demo$frame
) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

finer_bins_parameters <- FIMS::create_default_parameters(
  configurations = finer_bins_configurations,
  data = finer_bins_demo$frame
) |>
  tidyr::unnest(cols = data)

finer_bins_growth_sd_rows <- finer_bins_parameters |>
  dplyr::filter(
    module_name == "Growth",
    label == "length_at_age_sd_at_ref_ages"
  ) |>
  dplyr::mutate(
    age = c(1, 12),
    value = c(30, 73)
  )

finer_bins_parameters <- finer_bins_parameters |>
  dplyr::filter(
    !(module_name == "Growth" &
        label == "length_at_age_sd_at_ref_ages")
  ) |>
  dplyr::bind_rows(finer_bins_growth_sd_rows) |>
  dplyr::rows_update(
    tibble::tibble(
      module_name = "Growth",
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      value = c(150, 800, 0.22, 1, 12)
    ),
    by = c("module_name", "label")
  )

finer_bins_input <- finer_bins_parameters |>
  FIMS::initialize_fims(data = finer_bins_demo$frame)

finer_bins_input$model$ReportGrowthDerivedALKTensor(TRUE)

finer_bins_fit <- FIMS::fit_fims(
  input = finer_bins_input,
  optimize = TRUE,
  number_of_loops = 5,
  number_of_newton_steps = 0,
  get_sd = FALSE
)

finer_bins_growth_estimates <- FIMS::get_estimates(finer_bins_fit) |>
  dplyr::filter(
    module_name == "Growth",
    label %in% core_growth_labels
  )

# Pull the report only after the fitted growth estimates are in hand.
finer_bins_report <- FIMS::get_report(finer_bins_fit)

finer_bins_growth_comparison <- finer_bins_growth_estimates |>
  dplyr::select(label, input, estimated) |>
  dplyr::left_join(
    finer_bins_demo$true_parameters |>
      dplyr::filter(label %in% core_growth_labels),
    by = "label"
  )

finer_bins_growth_comparison

finer_bins_diagnostics <- tibble::tibble(
  scenario = "finer fleet bins",
  convergence = FIMS::get_opt(finer_bins_fit)$convergence,
  max_gradient = FIMS::get_max_gradient(finer_bins_fit)
)

finer_bins_diagnostics
```

### What should happen

The fitted growth curve should remain close to the generated truth. The ALK
comparison plots later in the vignette should show probability mapped onto
narrower length bins.

## Try 2b: Keep the Biological Size Grid Separate

### What this changes

This small check does not change the growth fit. It makes the user-facing grid
input more concrete by showing how the finer-bin example above enters FIMS
through explicit `length_bin` rows.

### Why this matters

Fleet observation bins are the piece a user controls directly. The biological
size grid used inside the growth-derived path is a separate population-side
object that FIMS checks against that fleet geometry.

### Run the check

```{r try-2b-size-grid-contract}
# The finer-bin example above is already a user-supplied grid example.
# The `lengths` argument passed to `make_vonb_demo_data()` becomes explicit
# `length_bin` rows in the FIMS input.
requested_user_grid <- finer_bins_demo$lengths

user_grid_from_rows <- finer_bins_demo$raw_data |>
  dplyr::filter(type == "length_bin") |>
  dplyr::pull(length) |>
  unique() |>
  sort()

resolved_grid_from_frame <- FIMS::get_lengths(finer_bins_demo$frame)

tibble::tibble(
  check = c(
    "explicit length_bin rows match the requested grid",
    "FIMS keeps the user-supplied fleet grid",
    "growth-derived ALK reporting still uses the path"
  ),
  result = c(
    isTRUE(all.equal(user_grid_from_rows, requested_user_grid)),
    isTRUE(all.equal(resolved_grid_from_frame, requested_user_grid)),
    length(finer_bins_report[["growth_derived_age_to_length_conversion"]][[1]]) > 0
  )
)

tibble::tibble(
  source = c("requested user grid", "resolved grid in FIMSFrame"),
  n_bins = c(length(requested_user_grid), length(resolved_grid_from_frame)),
  first_bin = c(min(requested_user_grid), min(resolved_grid_from_frame)),
  last_bin = c(max(requested_user_grid), max(resolved_grid_from_frame))
)
```

### What should happen

The first table should return `TRUE` in all three rows. The second table should
show that the grid the user supplied in the finer-bin example is the same grid
FIMS carries forward in the data object. This is the user-facing grid input,
while the population-side biological grid remains a separate internal part of
the growth-derived path.

## Try 3: Change Length-at-Age Variability

### What this changes

This run increases the SD-at-age anchors used by the interpolation-based
variability pathway.

### Why this matters

The mean growth curve and the variability around length-at-age are different
pieces of the growth pathway. The mean curve may remain similar, while the ALK
spreads probability across more length bins.

### Run the model

```{r try-3-wider-sd}
FIMS::clear()

# CHANGED: generate data with wider SD length-at-age.
wider_sd_demo <- make_vonb_demo_data(
  sd_at_ref_ages = c(45, 110)
)

wider_sd_configurations <- FIMS::create_default_configurations(
  data = wider_sd_demo$frame
) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

wider_sd_parameters <- FIMS::create_default_parameters(
  configurations = wider_sd_configurations,
  data = wider_sd_demo$frame
) |>
  tidyr::unnest(cols = data)

# CHANGED: use matching wider SD anchors in the model.
wider_sd_growth_sd_rows <- wider_sd_parameters |>
  dplyr::filter(
    module_name == "Growth",
    label == "length_at_age_sd_at_ref_ages"
  ) |>
  dplyr::mutate(
    age = c(1, 12),
    value = c(45, 110)
  )

wider_sd_parameters <- wider_sd_parameters |>
  dplyr::filter(
    !(module_name == "Growth" &
        label == "length_at_age_sd_at_ref_ages")
  ) |>
  dplyr::bind_rows(wider_sd_growth_sd_rows) |>
  dplyr::rows_update(
    tibble::tibble(
      module_name = "Growth",
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      value = c(150, 800, 0.22, 1, 12)
    ),
    by = c("module_name", "label")
  )

wider_sd_input <- wider_sd_parameters |>
  FIMS::initialize_fims(data = wider_sd_demo$frame)

wider_sd_input$model$ReportGrowthDerivedALKTensor(TRUE)

wider_sd_fit <- FIMS::fit_fims(
  input = wider_sd_input,
  optimize = TRUE,
  number_of_loops = 5,
  number_of_newton_steps = 0,
  get_sd = FALSE
)

wider_sd_growth_estimates <- FIMS::get_estimates(wider_sd_fit) |>
  dplyr::filter(
    module_name == "Growth",
    label %in% core_growth_labels
  )

# Pull the report
wider_sd_report <- FIMS::get_report(wider_sd_fit)

wider_sd_growth_comparison <- wider_sd_growth_estimates |>
  dplyr::select(label, input, estimated) |>
  dplyr::left_join(
    wider_sd_demo$true_parameters |>
      dplyr::filter(label %in% core_growth_labels),
    by = "label"
  )

wider_sd_growth_comparison

wider_sd_diagnostics <- tibble::tibble(
  scenario = "wider SD",
  convergence = FIMS::get_opt(wider_sd_fit)$convergence,
  max_gradient = FIMS::get_max_gradient(wider_sd_fit)
)

wider_sd_diagnostics
```

### What should happen

The fitted mean growth curve should remain close to the generated truth. The
ALK plots should show probability spread across more length bins.

## Try 4: Change the Second Reference Age

### What this changes

This run moves the second reference age from 12 to 15 and changes the generated
length at that reference age.

### Why this matters

`length_at_ref_age_2` is the length at the second reference age. It is not
`Linf`.

### Run the model

```{r try-4-second-reference-age}
FIMS::clear()

# CHANGED: move the second reference age from 12 to 15.
second_ref_demo <- make_vonb_demo_data(
  reference_age_for_length_2 = 15,
  length_at_ref_age_2 = 900,
  sd_at_ref_ages = c(30, 85)
)

second_ref_configurations <- FIMS::create_default_configurations(
  data = second_ref_demo$frame
) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

second_ref_parameters <- FIMS::create_default_parameters(
  configurations = second_ref_configurations,
  data = second_ref_demo$frame
) |>
  tidyr::unnest(cols = data)

# CHANGED: move the second SD anchor to age 15.
second_ref_growth_sd_rows <- second_ref_parameters |>
  dplyr::filter(
    module_name == "Growth",
    label == "length_at_age_sd_at_ref_ages"
  ) |>
  dplyr::mutate(
    age = c(1, 15),
    value = c(30, 85)
  )

# CHANGED: set the model's second reference age to 15.
second_ref_parameters <- second_ref_parameters |>
  dplyr::filter(
    !(module_name == "Growth" &
        label == "length_at_age_sd_at_ref_ages")
  ) |>
  dplyr::bind_rows(second_ref_growth_sd_rows) |>
  dplyr::rows_update(
    tibble::tibble(
      module_name = "Growth",
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      value = c(150, 850, 0.22, 1, 15)
    ),
    by = c("module_name", "label")
  )

second_ref_input <- second_ref_parameters |>
  FIMS::initialize_fims(data = second_ref_demo$frame)

second_ref_input$model$ReportGrowthDerivedALKTensor(TRUE)

second_ref_fit <- FIMS::fit_fims(
  input = second_ref_input,
  optimize = TRUE,
  number_of_loops = 5,
  number_of_newton_steps = 0,
  get_sd = FALSE
)

second_ref_growth_estimates <- FIMS::get_estimates(second_ref_fit) |>
  dplyr::filter(
    module_name == "Growth",
    label %in% core_growth_labels
  )

# Pull the report
second_ref_report <- FIMS::get_report(second_ref_fit)

second_ref_growth_comparison <- second_ref_growth_estimates |>
  dplyr::select(label, input, estimated) |>
  dplyr::left_join(
    second_ref_demo$true_parameters |>
      dplyr::filter(label %in% core_growth_labels),
    by = "label"
  )

second_ref_growth_comparison

second_ref_diagnostics <- tibble::tibble(
  scenario = "second reference age 15",
  convergence = FIMS::get_opt(second_ref_fit)$convergence,
  max_gradient = FIMS::get_max_gradient(second_ref_fit)
)

second_ref_diagnostics
```

### What should happen

The fitted second reference length should be close to the generated length at
age 15. The implied asymptote should still be above that reference length.

## Try 5: Turn Off the Full ALK Report

### What this changes

This run uses the same growth-derived pathway, but does not request the full
realized ALK tensor in the report.

### Why this matters

Using the growth-derived ALK pathway and reporting the full realized ALK tensor
are separate choices. This run is useful for showing that the model can use the
derived path without materializing the full tensor for inspection.

### Run the model

```{r try-5-alk-report-off}
FIMS::clear()

alk_report_off_demo <- make_vonb_demo_data()

alk_report_off_configurations <- FIMS::create_default_configurations(
  data = alk_report_off_demo$frame
) |>
  tidyr::unnest(cols = data) |>
  dplyr::mutate(
    module_type = dplyr::if_else(
      module_name == "Growth",
      "VonBertalanffy",
      module_type
    )
  ) |>
  tidyr::nest(.by = c(model_family, module_name, fleet))

alk_report_off_parameters <- FIMS::create_default_parameters(
  configurations = alk_report_off_configurations,
  data = alk_report_off_demo$frame
) |>
  tidyr::unnest(cols = data)

alk_report_off_growth_sd_rows <- alk_report_off_parameters |>
  dplyr::filter(
    module_name == "Growth",
    label == "length_at_age_sd_at_ref_ages"
  ) |>
  dplyr::mutate(
    age = c(1, 12),
    value = c(30, 73)
  )

alk_report_off_parameters <- alk_report_off_parameters |>
  dplyr::filter(
    !(module_name == "Growth" &
        label == "length_at_age_sd_at_ref_ages")
  ) |>
  dplyr::bind_rows(alk_report_off_growth_sd_rows) |>
  dplyr::rows_update(
    tibble::tibble(
      module_name = "Growth",
      label = c(
        "length_at_ref_age_1",
        "length_at_ref_age_2",
        "growth_coefficient_K",
        "reference_age_for_length_1",
        "reference_age_for_length_2"
      ),
      value = c(150, 800, 0.22, 1, 12)
    ),
    by = c("module_name", "label")
  )

alk_report_off_input <- alk_report_off_parameters |>
  FIMS::initialize_fims(data = alk_report_off_demo$frame)

# CHANGED: do not request the full realized ALK tensor.
# alk_report_off_input$model$ReportGrowthDerivedALKTensor(TRUE)

alk_report_off_fit <- FIMS::fit_fims(
  input = alk_report_off_input,
  optimize = TRUE,
  number_of_loops = 5,
  number_of_newton_steps = 0,
  get_sd = FALSE
)

alk_report_off_growth_estimates <- FIMS::get_estimates(alk_report_off_fit) |>
  dplyr::filter(
    module_name == "Growth",
    label %in% core_growth_labels
  )

# Pull the report only after the fitted growth estimates are in hand.
alk_report_off_report <- FIMS::get_report(alk_report_off_fit)

alk_report_off_growth_comparison <- alk_report_off_growth_estimates |>
  dplyr::select(label, input, estimated) |>
  dplyr::left_join(
    alk_report_off_demo$true_parameters |>
      dplyr::filter(label %in% core_growth_labels),
    by = "label"
  )

alk_report_off_growth_comparison

alk_report_off_diagnostics <- tibble::tibble(
  scenario = "ALK report off",
  convergence = FIMS::get_opt(alk_report_off_fit)$convergence,
  max_gradient = FIMS::get_max_gradient(alk_report_off_fit)
)

alk_report_off_diagnostics

# The model can still use the derived pathway, but the full tensor is not
# materialized in the report.
alk_report_off_realized_alk <-
  alk_report_off_report[["growth_derived_age_to_length_conversion"]][[1]]

tibble::tibble(
  realized_alk_length = length(alk_report_off_realized_alk)
)
```

### What should happen

The fitted growth values should still be close to the generated truth. The full
`growth_derived_age_to_length_conversion` object should have length zero because
the tensor was not requested for reporting.

## Compare Growth Estimates Across Runs

```{r compare-growth-estimates}
growth_parameter_comparison <- dplyr::bind_rows(
  baseline_growth_comparison |>
    dplyr::mutate(scenario = "baseline"),
  starting_values_growth_comparison |>
    dplyr::mutate(scenario = "changed starting values"),
  finer_bins_growth_comparison |>
    dplyr::mutate(scenario = "finer fleet bins"),
  wider_sd_growth_comparison |>
    dplyr::mutate(scenario = "wider SD"),
  second_ref_growth_comparison |>
    dplyr::mutate(scenario = "second reference age 15"),
  alk_report_off_growth_comparison |>
    dplyr::mutate(scenario = "ALK report off")
) |>
  dplyr::select(scenario, label, input, estimated, true_value)

growth_parameter_comparison

dplyr::bind_rows(
  baseline_diagnostics,
  starting_values_diagnostics,
  finer_bins_diagnostics,
  wider_sd_diagnostics,
  second_ref_diagnostics,
  alk_report_off_diagnostics
)
```

## Compare Estimated Growth Curves Across Runs

```{r prepare-growth-curve-comparison}
#| fig.width: 9
#| fig.height: 6
plot_ages <- seq(1, 50, by = 0.25)

baseline_estimated_values <- baseline_growth_estimates |>
  dplyr::select(label, estimated) |>
  tidyr::pivot_wider(names_from = label, values_from = estimated)

starting_values_estimated_values <- starting_values_growth_estimates |>
  dplyr::select(label, estimated) |>
  tidyr::pivot_wider(names_from = label, values_from = estimated)

finer_bins_estimated_values <- finer_bins_growth_estimates |>
  dplyr::select(label, estimated) |>
  tidyr::pivot_wider(names_from = label, values_from = estimated)

wider_sd_estimated_values <- wider_sd_growth_estimates |>
  dplyr::select(label, estimated) |>
  tidyr::pivot_wider(names_from = label, values_from = estimated)

second_ref_estimated_values <- second_ref_growth_estimates |>
  dplyr::select(label, estimated) |>
  tidyr::pivot_wider(names_from = label, values_from = estimated)

alk_report_off_estimated_values <- alk_report_off_growth_estimates |>
  dplyr::select(label, estimated) |>
  tidyr::pivot_wider(names_from = label, values_from = estimated)

growth_curve_comparison <- dplyr::bind_rows(
  tibble::tibble(
    scenario = "baseline",
    age = plot_ages,
    mean_length = vonb_two_reference_age(
      age = plot_ages,
      length_at_ref_age_1 = baseline_estimated_values$length_at_ref_age_1,
      length_at_ref_age_2 = baseline_estimated_values$length_at_ref_age_2,
      growth_coefficient_K = baseline_estimated_values$growth_coefficient_K,
      reference_age_for_length_1 = 1,
      reference_age_for_length_2 = 12
    )
  ),
  tibble::tibble(
    scenario = "changed starting values",
    age = plot_ages,
    mean_length = vonb_two_reference_age(
      age = plot_ages,
      length_at_ref_age_1 =
        starting_values_estimated_values$length_at_ref_age_1,
      length_at_ref_age_2 =
        starting_values_estimated_values$length_at_ref_age_2,
      growth_coefficient_K =
        starting_values_estimated_values$growth_coefficient_K,
      reference_age_for_length_1 = 1,
      reference_age_for_length_2 = 12
    )
  ),
  tibble::tibble(
    scenario = "finer fleet bins",
    age = plot_ages,
    mean_length = vonb_two_reference_age(
      age = plot_ages,
      length_at_ref_age_1 =
        finer_bins_estimated_values$length_at_ref_age_1,
      length_at_ref_age_2 =
        finer_bins_estimated_values$length_at_ref_age_2,
      growth_coefficient_K =
        finer_bins_estimated_values$growth_coefficient_K,
      reference_age_for_length_1 = 1,
      reference_age_for_length_2 = 12
    )
  ),
  tibble::tibble(
    scenario = "wider SD",
    age = plot_ages,
    mean_length = vonb_two_reference_age(
      age = plot_ages,
      length_at_ref_age_1 = wider_sd_estimated_values$length_at_ref_age_1,
      length_at_ref_age_2 = wider_sd_estimated_values$length_at_ref_age_2,
      growth_coefficient_K = wider_sd_estimated_values$growth_coefficient_K,
      reference_age_for_length_1 = 1,
      reference_age_for_length_2 = 12
    )
  ),
  tibble::tibble(
    scenario = "second reference age 15",
    age = plot_ages,
    mean_length = vonb_two_reference_age(
      age = plot_ages,
      length_at_ref_age_1 = second_ref_estimated_values$length_at_ref_age_1,
      length_at_ref_age_2 = second_ref_estimated_values$length_at_ref_age_2,
      growth_coefficient_K = second_ref_estimated_values$growth_coefficient_K,
      reference_age_for_length_1 = 1,
      reference_age_for_length_2 = 15
    )
  ),
  tibble::tibble(
    scenario = "ALK report off",
    age = plot_ages,
    mean_length = vonb_two_reference_age(
      age = plot_ages,
      length_at_ref_age_1 =
        alk_report_off_estimated_values$length_at_ref_age_1,
      length_at_ref_age_2 =
        alk_report_off_estimated_values$length_at_ref_age_2,
      growth_coefficient_K =
        alk_report_off_estimated_values$growth_coefficient_K,
      reference_age_for_length_1 = 1,
      reference_age_for_length_2 = 12
    )
  )
)
```

```{r plot-growth-curve-comparison, fig.alt = "Estimated Von Bertalanffy growth curves across scenarios."}
#| fig.width: 9
#| fig.height: 6
ggplot2::ggplot(
  growth_curve_comparison,
  ggplot2::aes(
    x = age,
    y = mean_length,
    color = scenario,
    linetype = scenario
  )
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(
    title = "Estimated VonB growth curves across scenarios",
    x = "Age",
    y = "Mean length",
    color = NULL,
    linetype = NULL
  ) +
  stockplotr::theme_noaa()
```

## Compare Growth-Derived Weight-at-Age

```{r compare-growth-derived-waa}
growth_waa_comparison <- dplyr::bind_rows(
  tibble::tibble(
    scenario = "baseline",
    year = rep(baseline_demo$years, each = length(baseline_demo$ages)),
    age = rep(baseline_demo$ages, times = length(baseline_demo$years)),
    growth_derived_mean_WAA =
      baseline_report[["growth_derived_mean_WAA"]][[1]]
  ),
  tibble::tibble(
    scenario = "changed starting values",
    year = rep(
      starting_values_demo$years,
      each = length(starting_values_demo$ages)
    ),
    age = rep(
      starting_values_demo$ages,
      times = length(starting_values_demo$years)
    ),
    growth_derived_mean_WAA =
      starting_values_report[["growth_derived_mean_WAA"]][[1]]
  ),
  tibble::tibble(
    scenario = "finer fleet bins",
    year = rep(finer_bins_demo$years, each = length(finer_bins_demo$ages)),
    age = rep(finer_bins_demo$ages, times = length(finer_bins_demo$years)),
    growth_derived_mean_WAA =
      finer_bins_report[["growth_derived_mean_WAA"]][[1]]
  ),
  tibble::tibble(
    scenario = "wider SD",
    year = rep(wider_sd_demo$years, each = length(wider_sd_demo$ages)),
    age = rep(wider_sd_demo$ages, times = length(wider_sd_demo$years)),
    growth_derived_mean_WAA =
      wider_sd_report[["growth_derived_mean_WAA"]][[1]]
  ),
  tibble::tibble(
    scenario = "second reference age 15",
    year = rep(second_ref_demo$years, each = length(second_ref_demo$ages)),
    age = rep(second_ref_demo$ages, times = length(second_ref_demo$years)),
    growth_derived_mean_WAA =
      second_ref_report[["growth_derived_mean_WAA"]][[1]]
  ),
  tibble::tibble(
    scenario = "ALK report off",
    year = rep(
      alk_report_off_demo$years,
      each = length(alk_report_off_demo$ages)
    ),
    age = rep(
      alk_report_off_demo$ages,
      times = length(alk_report_off_demo$years)
    ),
    growth_derived_mean_WAA =
      alk_report_off_report[["growth_derived_mean_WAA"]][[1]]
  )
)
```

```{r plot-growth-derived-waa, fig.alt = "Growth-derived mean weight-at-age across scenarios."}
#| fig.width: 9
#| fig.height: 6
ggplot2::ggplot(
  growth_waa_comparison,
  ggplot2::aes(
    x = age,
    y = growth_derived_mean_WAA,
    color = scenario,
    linetype = scenario
  )
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::facet_wrap(~ year) +
  ggplot2::labs(
    title = "Growth-derived mean weight-at-age across scenarios",
    x = "Age",
    y = "Mean weight-at-age",
    color = NULL,
    linetype = NULL
  ) +
  stockplotr::theme_noaa()
```

## Compare Whole-Model Outputs Across Runs

These plots use values from the report object. They are whole-model outputs,
so they are useful for seeing how changes move through the model, but they are
not one-to-one tests of the Growth module alone.

### Compare Expected Landings

```{r plot-landings-across-runs, fig.alt = "Expected landings across growth scenarios."}
#| fig.width: 9
#| fig.height: 6
landings_comparison <- dplyr::bind_rows(
  tibble::tibble(
    scenario = "baseline",
    year = baseline_demo$years,
    landings_expected =
      baseline_report[["landings_expected"]][[1]]
  ),
  tibble::tibble(
    scenario = "changed starting values",
    year = starting_values_demo$years,
    landings_expected =
      starting_values_report[["landings_expected"]][[1]]
  ),
  tibble::tibble(
    scenario = "finer fleet bins",
    year = finer_bins_demo$years,
    landings_expected =
      finer_bins_report[["landings_expected"]][[1]]
  ),
  tibble::tibble(
    scenario = "wider SD",
    year = wider_sd_demo$years,
    landings_expected =
      wider_sd_report[["landings_expected"]][[1]]
  ),
  tibble::tibble(
    scenario = "second reference age 15",
    year = second_ref_demo$years,
    landings_expected =
      second_ref_report[["landings_expected"]][[1]]
  ),
  tibble::tibble(
    scenario = "ALK report off",
    year = alk_report_off_demo$years,
    landings_expected =
      alk_report_off_report[["landings_expected"]][[1]]
  )
)

landings_target <- baseline_demo$raw_data |>
  dplyr::filter(type == "landings") |>
  dplyr::pull(value) |>
  dplyr::first()

landings_plot_limits <- range(
  c(
    landings_comparison$landings_expected,
    landings_target - 0.5,
    landings_target + 0.5
  )
)

ggplot2::ggplot(
  landings_comparison,
  ggplot2::aes(
    x = year,
    y = landings_expected,
    color = scenario,
    linetype = scenario
  )
) +
  ggplot2::geom_hline(
    yintercept = landings_target,
    color = "grey40",
    linetype = "dotted"
  ) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point() +
  ggplot2::coord_cartesian(ylim = landings_plot_limits) +
  ggplot2::labs(
    title = "Expected landings across growth scenarios",
    x = "Year",
    y = "Expected landings",
    color = NULL,
    linetype = NULL
  ) +
  stockplotr::theme_noaa()
```

### Compare Index Weight

```{r plot-index-weight-across-runs, fig.alt = "Index weight across growth scenarios."}
#| fig.width: 9
#| fig.height: 6
index_weight_comparison <- dplyr::bind_rows(
  tibble::tibble(
    scenario = "baseline",
    year = baseline_demo$years,
    index_weight =
      baseline_report[["index_weight"]][[1]]
  ),
  tibble::tibble(
    scenario = "changed starting values",
    year = starting_values_demo$years,
    index_weight =
      starting_values_report[["index_weight"]][[1]]
  ),
  tibble::tibble(
    scenario = "finer fleet bins",
    year = finer_bins_demo$years,
    index_weight =
      finer_bins_report[["index_weight"]][[1]]
  ),
  tibble::tibble(
    scenario = "wider SD",
    year = wider_sd_demo$years,
    index_weight =
      wider_sd_report[["index_weight"]][[1]]
  ),
  tibble::tibble(
    scenario = "second reference age 15",
    year = second_ref_demo$years,
    index_weight =
      second_ref_report[["index_weight"]][[1]]
  ),
  tibble::tibble(
    scenario = "ALK report off",
    year = alk_report_off_demo$years,
    index_weight =
      alk_report_off_report[["index_weight"]][[1]]
  )
)

ggplot2::ggplot(
  index_weight_comparison,
  ggplot2::aes(
    x = year,
    y = index_weight,
    color = scenario,
    linetype = scenario
  )
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point() +
  ggplot2::labs(
    title = "Index weight across growth scenarios",
    x = "Year",
    y = "Index weight",
    color = NULL,
    linetype = NULL
  ) +
  stockplotr::theme_noaa()
```

### Compare Mean Fishing Mortality by Year

```{r plot-mean-f-across-runs, fig.alt = "Mean fishing mortality across growth scenarios."}
#| fig.width: 9
#| fig.height: 6
baseline_mortality_F <- matrix(
  baseline_report[["mortality_F"]][[1]],
  nrow = length(baseline_demo$ages),
  ncol = length(baseline_demo$years)
)

starting_values_mortality_F <- matrix(
  starting_values_report[["mortality_F"]][[1]],
  nrow = length(starting_values_demo$ages),
  ncol = length(starting_values_demo$years)
)

finer_bins_mortality_F <- matrix(
  finer_bins_report[["mortality_F"]][[1]],
  nrow = length(finer_bins_demo$ages),
  ncol = length(finer_bins_demo$years)
)

wider_sd_mortality_F <- matrix(
  wider_sd_report[["mortality_F"]][[1]],
  nrow = length(wider_sd_demo$ages),
  ncol = length(wider_sd_demo$years)
)

second_ref_mortality_F <- matrix(
  second_ref_report[["mortality_F"]][[1]],
  nrow = length(second_ref_demo$ages),
  ncol = length(second_ref_demo$years)
)

alk_report_off_mortality_F <- matrix(
  alk_report_off_report[["mortality_F"]][[1]],
  nrow = length(alk_report_off_demo$ages),
  ncol = length(alk_report_off_demo$years)
)

mean_f_comparison <- dplyr::bind_rows(
  tibble::tibble(
    scenario = "baseline",
    year = baseline_demo$years,
    mean_F = colMeans(baseline_mortality_F)
  ),
  tibble::tibble(
    scenario = "changed starting values",
    year = starting_values_demo$years,
    mean_F = colMeans(starting_values_mortality_F)
  ),
  tibble::tibble(
    scenario = "finer fleet bins",
    year = finer_bins_demo$years,
    mean_F = colMeans(finer_bins_mortality_F)
  ),
  tibble::tibble(
    scenario = "wider SD",
    year = wider_sd_demo$years,
    mean_F = colMeans(wider_sd_mortality_F)
  ),
  tibble::tibble(
    scenario = "second reference age 15",
    year = second_ref_demo$years,
    mean_F = colMeans(second_ref_mortality_F)
  ),
  tibble::tibble(
    scenario = "ALK report off",
    year = alk_report_off_demo$years,
    mean_F = colMeans(alk_report_off_mortality_F)
  )
)

ggplot2::ggplot(
  mean_f_comparison,
  ggplot2::aes(
    x = year,
    y = mean_F,
    color = scenario,
    linetype = scenario
  )
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point() +
  ggplot2::labs(
    title = "Mean fishing mortality across growth scenarios",
    x = "Year",
    y = "Mean fishing mortality across ages",
    color = NULL,
    linetype = NULL
  ) +
  stockplotr::theme_noaa()
```

### Compare Spawning-Biomass Proxy

This generated example does not need to rely on an exported `spawning_biomass`
output. Instead, this plot calculates a simple spawning-biomass proxy from
numbers-at-age, growth-derived mean weight-at-age, and proportion mature-at-age.

```{r plot-spawning-biomass-proxy-across-runs, fig.alt = "Spawning biomass proxy across growth scenarios."}
# Use only the modeled years that match the WAA vector.
#| fig.width: 9
#| fig.height: 6
baseline_numbers_matrix <- matrix(
  baseline_report[["numbers_at_age"]][[1]],
  nrow = length(baseline_demo$ages)
)[, seq_along(baseline_demo$years)]

baseline_waa_matrix <- matrix(
  baseline_report[["growth_derived_mean_WAA"]][[1]],
  nrow = length(baseline_demo$ages),
  ncol = length(baseline_demo$years)
)

baseline_maturity_matrix <- matrix(
  baseline_report[["proportion_mature_at_age"]][[1]],
  nrow = length(baseline_demo$ages)
)[, seq_along(baseline_demo$years)]

starting_values_numbers_matrix <- matrix(
  starting_values_report[["numbers_at_age"]][[1]],
  nrow = length(starting_values_demo$ages)
)[, seq_along(starting_values_demo$years)]

starting_values_waa_matrix <- matrix(
  starting_values_report[["growth_derived_mean_WAA"]][[1]],
  nrow = length(starting_values_demo$ages),
  ncol = length(starting_values_demo$years)
)

starting_values_maturity_matrix <- matrix(
  starting_values_report[["proportion_mature_at_age"]][[1]],
  nrow = length(starting_values_demo$ages)
)[, seq_along(starting_values_demo$years)]

finer_bins_numbers_matrix <- matrix(
  finer_bins_report[["numbers_at_age"]][[1]],
  nrow = length(finer_bins_demo$ages)
)[, seq_along(finer_bins_demo$years)]

finer_bins_waa_matrix <- matrix(
  finer_bins_report[["growth_derived_mean_WAA"]][[1]],
  nrow = length(finer_bins_demo$ages),
  ncol = length(finer_bins_demo$years)
)

finer_bins_maturity_matrix <- matrix(
  finer_bins_report[["proportion_mature_at_age"]][[1]],
  nrow = length(finer_bins_demo$ages)
)[, seq_along(finer_bins_demo$years)]

wider_sd_numbers_matrix <- matrix(
  wider_sd_report[["numbers_at_age"]][[1]],
  nrow = length(wider_sd_demo$ages)
)[, seq_along(wider_sd_demo$years)]

wider_sd_waa_matrix <- matrix(
  wider_sd_report[["growth_derived_mean_WAA"]][[1]],
  nrow = length(wider_sd_demo$ages),
  ncol = length(wider_sd_demo$years)
)

wider_sd_maturity_matrix <- matrix(
  wider_sd_report[["proportion_mature_at_age"]][[1]],
  nrow = length(wider_sd_demo$ages)
)[, seq_along(wider_sd_demo$years)]

second_ref_numbers_matrix <- matrix(
  second_ref_report[["numbers_at_age"]][[1]],
  nrow = length(second_ref_demo$ages)
)[, seq_along(second_ref_demo$years)]

second_ref_waa_matrix <- matrix(
  second_ref_report[["growth_derived_mean_WAA"]][[1]],
  nrow = length(second_ref_demo$ages),
  ncol = length(second_ref_demo$years)
)

second_ref_maturity_matrix <- matrix(
  second_ref_report[["proportion_mature_at_age"]][[1]],
  nrow = length(second_ref_demo$ages)
)[, seq_along(second_ref_demo$years)]

alk_report_off_numbers_matrix <- matrix(
  alk_report_off_report[["numbers_at_age"]][[1]],
  nrow = length(alk_report_off_demo$ages)
)[, seq_along(alk_report_off_demo$years)]

alk_report_off_waa_matrix <- matrix(
  alk_report_off_report[["growth_derived_mean_WAA"]][[1]],
  nrow = length(alk_report_off_demo$ages),
  ncol = length(alk_report_off_demo$years)
)

alk_report_off_maturity_matrix <- matrix(
  alk_report_off_report[["proportion_mature_at_age"]][[1]],
  nrow = length(alk_report_off_demo$ages)
)[, seq_along(alk_report_off_demo$years)]

spawning_biomass_proxy_comparison <- dplyr::bind_rows(
  tibble::tibble(
    scenario = "baseline",
    year = baseline_demo$years,
    spawning_biomass_proxy = colSums(
      baseline_numbers_matrix *
        baseline_waa_matrix *
        baseline_maturity_matrix
    )
  ),
  tibble::tibble(
    scenario = "changed starting values",
    year = starting_values_demo$years,
    spawning_biomass_proxy = colSums(
      starting_values_numbers_matrix *
        starting_values_waa_matrix *
        starting_values_maturity_matrix
    )
  ),
  tibble::tibble(
    scenario = "finer fleet bins",
    year = finer_bins_demo$years,
    spawning_biomass_proxy = colSums(
      finer_bins_numbers_matrix *
        finer_bins_waa_matrix *
        finer_bins_maturity_matrix
    )
  ),
  tibble::tibble(
    scenario = "wider SD",
    year = wider_sd_demo$years,
    spawning_biomass_proxy = colSums(
      wider_sd_numbers_matrix *
        wider_sd_waa_matrix *
        wider_sd_maturity_matrix
    )
  ),
  tibble::tibble(
    scenario = "second reference age 15",
    year = second_ref_demo$years,
    spawning_biomass_proxy = colSums(
      second_ref_numbers_matrix *
        second_ref_waa_matrix *
        second_ref_maturity_matrix
    )
  ),
  tibble::tibble(
    scenario = "ALK report off",
    year = alk_report_off_demo$years,
    spawning_biomass_proxy = colSums(
      alk_report_off_numbers_matrix *
        alk_report_off_waa_matrix *
        alk_report_off_maturity_matrix
    )
  )
)
```

```{r spawning-biomass-proxy-comparison-plot, fig.alt = "Spawning biomass proxy across growth scenarios.", fig.width = 9, fig.height = 6}
ggplot2::ggplot(
  spawning_biomass_proxy_comparison,
  ggplot2::aes(
    x = year,
    y = spawning_biomass_proxy,
    color = scenario,
    linetype = scenario
  )
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point() +
  ggplot2::labs(
    title = "Spawning biomass proxy across growth scenarios",
    x = "Year",
    y = "Numbers-at-age x WAA x maturity",
    color = NULL,
    linetype = NULL
  ) +
  stockplotr::theme_noaa()
```
