---
title: "Cpp Polymorphism"
author: "Andrea Havron"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the [C++ Inheritance](https://noaa-fims.github.io/FIMS_training/articles/CppInheritance.html) example, we saw how derived classes can inherit from base classes. This example resulted in code that required `if()` statements to calculate a function for the different classes. Polymorphism can be used instead to specify different behaviors for the different classes to avoid conditional statements in the code, which are hard to maintain and extend. 

## C++ Polymorphism

 - a function or operator works different when used in different context
 - can implement when classes inherit from each other
 - the result is a single action or function that produces different results based on the derived class
 - **virtual**: C++ function that gets overridden at runtime
 
Set up base class as before. This time, the shared object is the function, `area()`. Note that the area function is proceeded by **virtual** and set to equal 0. This is done to specify this is a function being overwritten by functions from the child class.
```cpp
class Shape {
  public:
    //constructor
    Shape() {}
    
    virtual ~Shape() {}
    
    // Virtual function to calculate area
    virtual double area() = 0;
    
  };
```

The child classes are then specified. Again, we use **virtual** to indicate this function is overwriting the function in the base class.
```cpp
class Circle : public Shape {
  double radius;
  
  public:
    Circle(double r) : Shape(){
      radius = r;
    }
  
  // Override area() for Circle
  virtual double area() {
    return M_PI * radius * radius;
  }
};

class Rectangle : public Shape {
  private:
    double length, height;
  
  public:
    Rectangle(double l, double h) : Shape(){
      length = l;
      height = h;
    }
  
  // Override area() for Rectangle
  virtual double area()  {
    return length * height;
  }
};
```
We are going to expose these functions to R using Rcpp Modules. See [Intro to Rcpp](https://noaa-fims.github.io/FIMS_training/articles/IntroRcpp.html#rcpp-modules) for a refresher on Rcpp Modules.

```cpp
RCPP_MODULE(shape) {
  Rcpp::class_<Shape>("Shape")
  .method("area", &Shape::area);
  
  Rcpp::class_<Circle>("Circle")
    .derives<Shape>("Shape")
    .constructor<double>();
  
  Rcpp::class_<Rectangle>("Rectangle")
    .derives<Shape>("Shape")
    .constructor<double, double>();
}
```

Now we can load the module using `Rcpp::loadModule()` where the second argument is `TRUE` so that all objects are loaded into the R environment. The 'new' functions creates a new instance of the class.

```{r}
library(Rcpp)
sourceCpp("../src/my_polymorphism.cpp")
loadModule("shape", TRUE)
c <- new(Circle, 2)
r <- new(Rectangle, 3, 4)

c$area()
r$area()
```

