---
title: "Intro to Rcpp"
author: "Christine Stawitz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro to Rcpp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r developer-comment, include=FALSE}
# The section below creates a collpase/expand all button
```
<div id="expand-collapse-buttons" style="margin-bottom: 20px;">
  <button onclick="expandAll()" class="btn btn-primary">Expand All</button>
  <button onclick="collapseAll()" class="btn btn-secondary">Collapse All</button>
</div>

<script>
function expandAll() {
  const details = document.querySelectorAll('details');
  details.forEach((detail) => {
    detail.setAttribute('open', 'true');
  });
}

function collapseAll() {
  const details = document.querySelectorAll('details');
  details.forEach((detail) => {
    detail.removeAttribute('open');
  });
}
</script>

# Sources
- "Advanced `R`"  by Hadley Wickham (http://adv-r.had.co.nz/Rcpp.html)
- "`Rcpp`: Seamless `R` and `C++` integration" by Dirk Eddelbuttel and Romain Francois, Journal of Statistical Software (https://www.jstatsoft.org/article/view/v040i08)
- "`Rcpp` for everyone" by Masaki E. Tsuda (https://teuder.github.io/rcpp4everyone_en/)
- "`Rcpp` Modules" vignette by Eddelbuttel and Francois (https://cran.rstudio.com/web/packages/Rcpp/vignettes/Rcpp-modules.pdf)


# What is `RCpp` and why should I care?

## Definition
"The `RCpp` package provides a consistent API for seamlessly accessing, extending or modifying R objects at the C++ level" - Eddelbuttel and Francois.

## API
a set of functions and procedures allowing the creation of applications that access the features or data of an operating system, application, or other service.

* One of the most widely used `R` extensions (over 1000 packages)
* With very minimal knowledge of `C++` you can speed up function runs A LOT
* The most efficient `R` functions are written in `C++` and called from `R`


# Writing `C++` functions in `R`

* You can write `C++` functions inline in your R code using `cppFunction`
* You can also compile single lines of code directly using `evalCpp`

In R script or console:
<details>
<summary><strong>Example: Inline C++ from R</strong></summary>

```{r}
library(Rcpp)
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')
# add works like a regular R function
add
add(1, 2, 3)

evalCpp( "std::numeric_limits<double>::max()" )
```
</details>

# Calling `C++` from `R`

mean.cpp

<details>
<summary><strong>Example: C++ Mean Function</strong></summary>

```{Rcpp meanC}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}
```
</details>


* `#include <Rcpp.h>` allows you to call `Rcpp` syntax in the file. You can omit `using namespace Rcpp` if you use `::` i.e. `Rcpp::NumericVector`
* `// [[Rcpp::export]]` - put this above any function you want to make available in `R`
* `NumericVector, .size()` useful type, function implemented in `Rcpp`

<details>
<summary><strong>Example: Sourcing C++ Code in R</strong></summary>

```{r}
# code can be saved in .cpp file and sourced:
# sourceCpp("mean.cpp")

# code can be sourced directly
src <-
"#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}
"
sourceCpp(code = src)
```
</details>

<details>
<summary><strong>Example: Benchmarking R vs C++</strong></summary>

```{r microbench}
library(microbenchmark)
x <- runif(1e5)
microbenchmark(
  mean(x),
  meanC(x)
)

```
</details>

* `sourceCpp` compiles the file
* More `Rcpp` syntax [here](https://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf)


# Typing and methods

## Rcpp Types

- Scalar classes
    * `int`
    * `double`
    * `bool`
    * `String`
- Vector classes
    * `IntegerVector`
    * `NumericVector`
    * `LogicalVector`
    * `CharacterVector`


## Rcpp methods

- Static methods (called with `::` on class)
- Member functions or methods (called with `.` on object)

<details>
<summary><strong>Example: Rcpp Vectors and Methods</strong></summary>

```{r}
src <-
  "#include <Rcpp.h>
  using namespace Rcpp;

  // [[Rcpp::export]]
  NumericVector fun(){
    //create new vector of size 3
    NumericVector v = NumericVector::create(1,2,3);
    Rcout << NumericVector::get_na() << std::endl;

    //operate on object v
    Rcout << v.size() << std::endl;
    //append a new value to the vector
    v.push_back(4);

    return v;
  }
"

sourceCpp(code = src)
fun()
```
</details>

# `Rcpp` modules

<details>
<summary><strong>Example: Rcpp Module Skeleton</strong></summary>

```{Rcpp eval=FALSE}
RCPP_MODULE(unif_module) {
class_<Uniform>("Uniform")
.constructor<double,double>()
.field("min", &Uniform::min, "minimum value")
.field("max", &Uniform::max, "maximum value")
.method("draw", &Uniform::draw)
;
}
```
</details>

* You can expose these using the `Rcpp` macro `RCPP_MODULE`
* the `class_<Uniform>` constructor argument is what we will call the class from `R`
* `.field`, `.constructor`, `.method`, `.property`
* `.field` can be used with two or three arguments
* `.field_readonly` prevents it from being modified from `R`