---
title: "Design Document: FIMS Rcpp Interface Refactor to Rcpp::XPtr"
author: "FIMS Development Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Motivation

The current FIMS architecture relies on S4 Rcpp::Module classes to expose C++ interface
objects (such as `AgeCompDataInterface` or `PopulationInterface`) directly to R. This
design introduces two structural problems that compound each other.

### 1.1 The Copy Problem

When `methods::new(Population)` is called in R, Rcpp owns the C++ object's memory through
its own internal mechanism. A `std::shared_ptr` cannot safely be created pointing at a
Rcpp-managed object, because Rcpp controls its lifetime. However, `CreateTMBModel()` needs
to iterate over all interface objects to call `add_to_fims_tmb()` via a registry it owns.
The solution in the current design is to make a heap copy of every interface object in the
constructor:

```cpp
// PopulationInterface constructor — current approach
PopulationInterface() : PopulationInterfaceBase() {
    // ...
    std::shared_ptr<PopulationInterface> population =
        std::make_shared<PopulationInterface>(*this);  // heap copy
    FIMSRcppInterfaceBase::fims_interface_objects.push_back(population);
    PopulationInterfaceBase::live_objects[this->id] = population;
}
```

Because these global static collections (variables that are shared across the program and persist for the lifetime of the program) maintain strong `std::shared_ptr` references to
every object created, their reference counts never drop to zero. Even when an R variable is
overwritten (e.g., `my_fleet <- NULL`), the underlying C++ memory is never deallocated
until an explicit manual call to `clear()`. This prevents clean iterative modeling workflows
(bootstrapping, sensitivity runs, profile likelihood) within a single R session.

### 1.2 The SharedPrimitive Overhead Problem

When users initialize a module (e.g. pop <- methods::new(Population)), a copy gets stored in the registry. When the user later sets
`pop$n_years <- 30` on the R-side S4 object, this modifies the original Rcpp-managed object
but not the registry copy. `CreateTMBModel()` would see stale initial values.

The `SharedInt`, `SharedBoolean`, `SharedReal`, and `SharedString` wrapper classes were
introduced to solve this synchronization problem: by wrapping scalar fields in `shared_ptr`,
both the original and the registry copy share the same underlying value. Modifying the
original is automatically seen by the copy.

However, this comes at a significant memory cost. Each `SharedPrimitive` field requires one
heap allocation for a single scalar. On a 64-bit system, `make_shared<int>(0)` allocates
approximately 28 bytes (16-byte control block + 4-byte int + alignment padding) for 4 bytes
of actual data — a 7:1 overhead ratio. `PopulationInterface` alone has 8 `SharedInt`, 2
`SharedBoolean`, and 1 `SharedString` configuration fields: roughly 11 heap allocations
before any model data is set. Additionally, every `Variable` element in every
`VariableVector` carries a `SharedString estimation_type_m`, adding one heap allocation per
parameter element across the entire model.

The SharedPrimitive types and the constructor copies are symptoms of the same root cause:
the S4 module system forces a two-object design (Rcpp-owned original + heap registry copy)
that requires shared ownership of individual fields to stay synchronized.

### 1.3 The Rcpp::XPtr Solution

By transitioning the R-to-C++ boundary to `Rcpp::XPtr<std::shared_ptr<T>>`, both problems
are eliminated together.

**One object, no copies.** The creator function allocates the C++ object on the heap
directly and hands ownership to a `shared_ptr`. The XPtr wraps that `shared_ptr`. There is
no Rcpp-managed original and no registry copy — there is one object. User modifications via
standalone setter functions go directly to that object.

**SharedPrimitives become unnecessary.** Without a copy to keep synchronized with the
original, scalar fields can revert to plain C++ types (`int`, `bool`, `double`,
`std::string`). The `Variable::estimation_type_m` field can become an enum.

**Automated lifecycle management.** When an R variable holding an XPtr is overwritten or
goes out of scope, R's garbage collector deletes the external pointer wrapper, decrementing
the `shared_ptr` reference count. Memory is freed without requiring an explicit `clear()`
call.

---

## 2. Refactoring Steps

### Step 1: Replace `SharedPrimitive` scalar fields with plain C++ types

In every interface class, replace `SharedInt`, `SharedBoolean`, `SharedReal`, and
`SharedString` fields that hold scalar configuration values with their plain C++ equivalents.
This is only safe under XPtr because there is no registry copy that needs to remain
synchronized with user modifications.

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_population.hpp

// BEFORE — SharedPrimitive fields (one heap allocation each)
SharedInt n_ages = 0;
SharedInt n_fleets;
SharedInt n_years;
SharedInt n_lengths;
SharedInt maturity_id;
SharedInt growth_id;
SharedInt recruitment_id;
SharedInt recruitment_err_id;
SharedString name = fims::to_string("NA");

// AFTER — plain C++ types (no heap allocation per scalar)
int n_ages = 0;
int n_fleets = 0;
int n_years = 0;
int n_lengths = 0;
int maturity_id = -1;
int growth_id = -1;
int recruitment_id = -1;
int recruitment_err_id = -1;
std::string name = "NA";
```

`VariableVector` retains its `shared_ptr<vector<Variable>>` internal storage because R-side
access to individual elements relies on shared ownership within the vector. Only standalone
scalar configuration fields change.

`SharedBoolean` fields in base classes (`initialize_catch_at_age`,
`initialize_surplus_production`) follow the same pattern — replace with plain `bool`.

### Step 2: Replace `Variable::estimation_type_m` with an enum

Every `Variable` element in every `VariableVector` currently carries a
`SharedString estimation_type_m`, which heap-allocates one `std::string` per element.
Replace it with an enum stored as a single byte. The field is also renamed to
`estimation_status_m`: what it records is a parameter's status in the estimation
machinery, not a category of estimation.

The enum does not live beside `Variable`. It governs parameter registration throughout
the model core, which must compile without Rcpp, so it belongs in a shared header that
carries no Rcpp dependency:

```cpp
// inst/include/common/enumerations.hpp

namespace fims_enum {

enum class EstimationStatus : uint8_t {
  kAssumedKnown    = 0,
  kFixedEffects    = 1,
  kRandomEffects   = 2,
  kDerivedQuantity = 3
};

// Reports failure with std::invalid_argument rather than Rcpp::stop, so the
// header stays usable from the model core. Rcpp catches it at the R boundary
// and surfaces it as an ordinary R error.
inline EstimationStatus EstimationStatusFromString(const std::string &status);
inline std::string EstimationStatusToString(EstimationStatus status);

}  // namespace fims_enum
```

`kDerivedQuantity` is a fourth status the string field never distinguished: a quantity
computed from other parameters, which is reported but never handed to the optimizer.

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_interface_base.hpp

// BEFORE — one heap allocation per Variable element
class Variable {
 public:
  SharedString estimation_type_m = SharedString("constant");
  // ...
};

// AFTER — zero heap allocations, 1 byte per Variable element
class Variable {
 public:
  fims_enum::EstimationStatus estimation_status_m =
      fims_enum::EstimationStatus::kAssumedKnown;
  // ...
};
```

Update all callers that previously compared strings (e.g.,
`estimation_type_m.get() == "fixed_effects"`) to use the enum values:

```cpp
// BEFORE
if (this->log_M[i].estimation_type_m.get() == "fixed_effects") { ... }

// AFTER
if (this->log_M[i].estimation_status_m ==
    fims_enum::EstimationStatus::kFixedEffects) { ... }
```

Update `VariableVector::set_estimation_status()` to accept an
`Rcpp::CharacterVector` and convert string input to enum internally, preserving the
existing R-side API. `Variable::get_estimation_status()` converts back, so R continues
to see and set the status by name and never handles the integer.

### Step 3: Remove constructor copies and implement standalone lifecycle functions

Strip `live_objects` map insertions and `fims_interface_objects` push_back calls from all
interface class constructors. The constructor should perform only pure initialization:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_data.hpp

// BEFORE — constructor registers in two global registries
AgeCompDataInterface(int ymax = 0, int amax = 0) : DataInterfaceBase() {
    this->amax = amax;
    this->ymax = ymax;
    this->age_comp_data.resize(amax * ymax);
    this->uncertainty.resize(amax * ymax);
    DataInterfaceBase::live_objects[this->id] =
        std::make_shared<AgeCompDataInterface>(*this);
    FIMSRcppInterfaceBase::fims_interface_objects.push_back(
        DataInterfaceBase::live_objects[this->id]);
}

// AFTER — pure initialization only
AgeCompDataInterface(int ymax = 0, int amax = 0) : DataInterfaceBase() {
    this->amax = amax;
    this->ymax = ymax;
    this->age_comp_data.resize(amax * ymax);
    this->uncertainty.resize(amax * ymax);
}
```

Delete the copy constructor of every migrated class as well, and delete copy assignment
alongside it. Removing the copy constructor is not sufficient: C++ then synthesizes an
implicit one, so the class stays copyable and a copy would produce a second object holding
a duplicate `id`. `= delete` makes any copy a compile error instead:

```cpp
AgeCompDataInterface(const AgeCompDataInterface &) = delete;
AgeCompDataInterface &operator=(const AgeCompDataInterface &) = delete;
```

Implement standalone creator functions in the corresponding `.cpp` source file. Wrap the
`shared_ptr` (not a raw pointer) inside the XPtr so that the reference count is safely
decremented by R's garbage collector when the XPtr is collected.

Write one creator per *family* rather than per concrete class. The XPtr is typed to the
family's base class, and a type name selects which derived class to allocate. Convert that
name to an enum first — `switch` requires an integral or enum operand, and the enum gives
the compiler a chance to flag a missing case when a new type is added later.

A family's type enum lives beside the classes it names, so adding a functional form
touches one file. Only enums with users in more than one part of FIMS — `EstimationStatus`,
`DistributionType` — go in `common/enumerations.hpp`. Converters throw
`std::invalid_argument` rather than calling `Rcpp::stop`, which keeps them usable from the
model core; Rcpp catches the exception at the boundary and surfaces it as an R error.

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_data.hpp

enum class DataType : uint8_t {
  age_comp = 0, length_comp = 1, index = 2, catch_data = 3
};

inline DataType DataTypeFromString(const std::string &type) {
  if (type == "age_comp") return DataType::age_comp;
  // ... remaining types ...
  throw std::invalid_argument(
      "Invalid data type: '" + type +
      "'. Valid options are: age_comp, length_comp, index, catch.");
}
```

```cpp
// src/rcpp_data.cpp

using SharedData = std::shared_ptr<DataInterfaceBase>;

Rcpp::XPtr<SharedData> create_data_interface_(std::string data_type,
                                              int n_years, int n_bins = 0) {
    DataType data_type_ = DataTypeFromString(data_type);

    // Reject a bin count on a one-dimensional type rather than ignoring it,
    // so a mistake in the call surfaces here instead of as a wrong-shaped
    // model later.
    if (n_bins != 0 && (data_type_ == DataType::index ||
                        data_type_ == DataType::catch_data)) {
      Rcpp::stop("Data type '" + data_type +
                 "' is one-dimensional; `n_bins` must be 0, got " +
                 std::to_string(n_bins) + ".");
    }

    SharedData data_interface;
    switch (data_type_) {
      case DataType::age_comp:
        data_interface =
            std::make_shared<AgeCompDataInterface>(n_years, n_bins);  break;
      case DataType::length_comp:
        data_interface =
            std::make_shared<LengthCompDataInterface>(n_years, n_bins); break;
      case DataType::index:
        data_interface = std::make_shared<IndexDataInterface>(n_years); break;
      case DataType::catch_data:
        data_interface = std::make_shared<CatchDataInterface>(n_years); break;
    }
    // Heap-allocate the shared_ptr itself and hand it to the XPtr.
    // 'true' binds the XPtr to R's garbage collector: when the R variable
    // is collected, 'delete shared_ptr*' is called, decrementing the ref count.
    return Rcpp::XPtr<SharedData>(new SharedData(data_interface), true);
}
```

Only the creator branches on type, because it is the one place where the concrete class
does not yet exist. Everything after it reaches the right object through a virtual
function on the base, or does not need to know the type at all.

That principle applies to the setters too, and it collapses them much further than one
per family. A setter's only type-specific act is picking *which member* to write, and a
member can be identified by name. So instead of a function per field, the base class
answers three questions by name, and every derived class overrides them with a short
list of the names it recognizes:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_interface_base.hpp

class FIMSRcppInterfaceBase {
 public:
  virtual uint32_t get_id() = 0;
  virtual VariableVector *get_parameter(const std::string &name) {
    return nullptr;
  }
  virtual fims::Vector<double> *get_vector(const std::string &name) {
    return nullptr;
  }
  // ...
};
```

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_data.hpp — AgeCompDataInterface

virtual fims::Vector<double> *get_vector(const std::string &name) {
  if (name == "values")      return &this->age_comp_data;
  if (name == "uncertainty") return &this->uncertainty;
  return nullptr;
}
```

`IndexDataInterface` answers `"values"` with `index_data`, and so on. A name a class does
not recognize returns `nullptr`, which the caller turns into an R error naming the module
and the field, rather than a silent no-op.

Four functions then cover every module in FIMS, registered once in
`src/rcpp_interface_base.cpp` and taking the *base* XPtr, so they do not care which family
they were handed:

```cpp
// src/rcpp_interface_base.cpp

void set_parameter_(Rcpp::XPtr<SharedBase> xp, std::string name,
                    Rcpp::NumericVector values,
                    Rcpp::CharacterVector estimation_status);
void set_vector_(Rcpp::XPtr<SharedBase> xp, std::string name,
                 Rcpp::NumericVector values);
uint32_t get_module_id_(Rcpp::XPtr<SharedBase> xp);
uint32_t get_parameter_id_(Rcpp::XPtr<SharedBase> xp, std::string name);
```

`set_parameter_` writes estimable quantities (a `VariableVector`, carrying an estimation
status per element); `set_vector_` writes fixed numeric input such as observations and
uncertainties. `get_parameter_id_` returns a parameter's `VariableVector` ID, which is
what links a distribution to the quantity it applies to — see Section 4c.

Fields that are neither, because they are plain scalars rather than vectors, still get a
small family-specific setter:

```cpp
// src/rcpp_population.cpp

void set_population_constants_(Rcpp::XPtr<SharedPopulation> xp,
                               int n_years, int n_ages, int n_lengths) {
    (*xp)->n_years   = n_years;
    (*xp)->n_ages    = n_ages;
    (*xp)->n_lengths = n_lengths;
}
```

Register these functions inside the existing module registration blocks. Do **not** use
`// [[Rcpp::exports]]`, as this causes essential TMB callables to be overwritten by
`Rcpp::compileAttributes()`. Add a trailing underscore to every registered name to prevent
name collisions with the R wrapper functions defined in Section 4b. Remove the
`Rcpp::class_<T>` registration for any interface class that has been fully migrated to XPtr:

```cpp
// src/rcpp_data.cpp

void register_data(Rcpp::Module& m) {
    // Rcpp::class_<> registrations removed — users no longer call methods::new().
    Rcpp::function("create_data_interface_", &create_data_interface_);
    Rcpp::function("data_to_fims_xptr_",     &data_to_fims_xptr_);
}
```

Two registered functions cover all four data types, because setting values, setting
uncertainty, and reading the ID are all handled by the shared functions above. A
per-class design would need roughly twenty. Across the ten module source files the
migration leaves 52 registered functions where the `Rcpp::class_<T>` design had about 130.

#### Reporting after the registries are gone

Deleting `live_objects` removes something the model relied on. When
`CatchAtAgeInterface` writes its JSON output it has to describe modules it does not own:
a fleet names its selectivity, a population names its growth and maturity, but the
interface holds only their integer IDs. It previously turned an ID back into an object by
looking it up in the family's `live_objects` map. With the maps gone there is nothing to
look up.

The modules are already reachable from R, so R hands them in. `to_json()` takes the same
list of base XPtrs that the registry builds for `CreateTMBModel()`:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_models.hpp
virtual std::string to_json(const std::vector<SharedBase> &report_modules);
```

The list is untyped — every entry is a `FIMSRcppInterfaceBase` pointer — so `to_json()`
sorts it into one map per family on the way in. A single helper does the sorting by
attempting the downcast and reporting whether it succeeded:

```cpp
template <typename T>
static bool try_store_module(const SharedBase &module,
                             std::map<uint32_t, std::shared_ptr<T>> &modules) {
  if (std::shared_ptr<T> typed_module = std::dynamic_pointer_cast<T>(module)) {
    modules[typed_module->get_id()] = typed_module;
    return true;
  }
  return false;
}
```

Each module is offered to each map until one accepts it; anything no map claims is
ignored rather than treated as an error, because the registry legitimately contains
modules that are not reportable. The resulting maps are keyed by ID, so the rest of the
reporting code performs the same ID lookup it always did — the maps are built per call
from what R supplied, instead of persisting for the life of the session.

This is why reporting-only modules do not need a `shared_ptr` link from the model.
Structural links exist where one module *computes* with another; a module that is merely
described in the output is passed in at reporting time. In particular, distributions are
not linked to `CatchAtAgeInterface`.

### Step 4: Implement dispatch through a shared base class for `CreateTMBModel()`

`CreateTMBModel()` currently iterates `FIMSRcppInterfaceBase::fims_interface_objects` to
call `add_to_fims_tmb()` on every registered object. Under XPtr, the C++ global registry
is removed. This step replaces it with two pieces: a C++ mechanism for calling
`add_to_fims_tmb()` through a common base class pointer, and an R-side registry that
automatically tracks objects as they are created so the user does not need to build the list
by hand.

#### 4a. Why two XPtr types are needed

Each concrete creator function (e.g., `create_age_comp()`, `create_population()`) returns
an XPtr typed to its specific class — `XPtr<SharedAgeComp>`, `XPtr<SharedPopulation>`, etc.
These types are all different, so they cannot be stored together in a single list and passed
to `CreateTMBModel()`.

The solution is a second, less specific XPtr typed to the common base class
`FIMSRcppInterfaceBase`, which all interface classes inherit from. Because
`add_to_fims_tmb()` is declared as a `virtual` method on that base class, calling it through
a base class pointer is sufficient — C++ automatically invokes the correct derived-class
implementation at runtime. This is what "dispatch through a shared base class" means:
`CreateTMBModel()` does not need to know whether it is looking at an `AgeCompDataInterface`
or a `PopulationInterface`; it just calls `add_to_fims_tmb()` and C++ figures out the rest.

The two XPtr types point to the same underlying C++ object in memory. The concrete XPtr is
what the user works with (setting values, reading results). The base XPtr is only needed
by `CreateTMBModel()` to make that call. Both XPtrs share ownership of the object via
`shared_ptr`, so the object stays alive as long as either pointer exists.

**Define the base alias** in `rcpp_interface_base.hpp`:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_interface_base.hpp
using SharedBase = std::shared_ptr<FIMSRcppInterfaceBase>;
```

**Provide a conversion function for each concrete type.** This function takes the
concrete XPtr the user already has and returns a second XPtr pointing to the same object
but typed as the base class. C++ allows this conversion implicitly because
`AgeCompDataInterface` inherits from `FIMSRcppInterfaceBase` — no special cast syntax
is needed:

```cpp
// src/rcpp_data.cpp

// create_data_interface_ — allocates a NEW data interface object
// data_to_fims_xptr_     — does NOT create a new object; returns a second pointer
//                          to the SAME object typed as the interface base class
Rcpp::XPtr<SharedBase> data_to_fims_xptr_(Rcpp::XPtr<SharedData> xp) {
    SharedBase base = *xp;  // implicit conversion: shared_ptr<Derived> -> shared_ptr<Base>
    return Rcpp::XPtr<SharedBase>(new SharedBase(base), true);
}
```

One conversion function per family is enough, for the same reason as the creator: the
argument is already typed to the family's base class. A single conversion function across
*all* families is not possible, because `Rcpp::function` registers one concrete signature
and `XPtr<SharedData>`, `XPtr<SharedFleet>`, and `XPtr<SharedPopulation>` are distinct C++
types.

Register it inside the existing module registration block, again with a trailing underscore
so the clean names remain available for the R wrappers in Section 4b:

```cpp
void register_data(Rcpp::Module& m) {
    Rcpp::function("create_data_interface_", &create_data_interface_);
    Rcpp::function("data_to_fims_xptr_",     &data_to_fims_xptr_);
}
```

Register `CreateTMBModel` and `clear` the same way in their respective registration blocks:

```cpp
Rcpp::function("CreateTMBModel_", &CreateTMBModel);
Rcpp::function("clear_",          &clear_impl);
```

**Update `CreateTMBModel()` in C++ to accept a list of base XPtrs** and call
`add_to_fims_tmb()` on each:

```cpp
// inst/include/interface/rcpp/rcpp_interface.hpp

bool CreateTMBModel(Rcpp::List xptr_list) {
    init_logging();

    std::shared_ptr<fims_info::Information<TMB_FIMS_REAL_TYPE>> info0 =
        fims_info::Information<TMB_FIMS_REAL_TYPE>::GetInstance();
    info0->Clear();

    std::shared_ptr<fims_info::Information<TMBAD_FIMS_TYPE>> info =
        fims_info::Information<TMBAD_FIMS_TYPE>::GetInstance();
    info->Clear();

    for (int i = 0; i < xptr_list.size(); i++) {
        Rcpp::XPtr<SharedBase> xp =
            Rcpp::as<Rcpp::XPtr<SharedBase>>(xptr_list[i]);
        (*xp)->add_to_fims_tmb();  // C++ calls the correct derived-class method
    }

    info0->CreateModel();
    info0->CheckModel();
    info->CreateModel();

    std::shared_ptr<fims_model::Model<TMB_FIMS_REAL_TYPE>> m0 =
        fims_model::Model<TMB_FIMS_REAL_TYPE>::GetInstance();

    return true;
}
```

#### 4b. R-side automatic registry (intermediate step before R6)

Rather than requiring the user to manually build the list passed to `CreateTMBModel()` —
which risks silently omitting a component — an R-side package-level registry can be
maintained automatically. The registry is a simple list stored inside a private package
environment (not visible to the user). Thin R wrapper functions around each C++ creator
call both `create_<type>()` and `<type>_to_fims_xptr()` and add the result to the registry.
The user-facing `CreateTMBModel()` and `clear()` functions are also R wrappers that operate
on this registry.

This approach is more reliable than scanning the R environment for XPtr objects, because
registration happens at the moment of creation — it is not affected by variable names,
scoping, or objects stored inside lists.

```r
# R/fims_interface.R

# Private package-level registry — not exported, not user-visible
.fims_registry <- new.env(parent = emptyenv())
.fims_registry$objects <- list()

# R wrapper around C++ create_data_interface_:
# creates the XPtr, converts to base XPtr, registers it, returns the XPtr.
# The C++ side has one creator per family; the R side keeps a friendly name per
# type, so the type string never appears in user code. The trailing underscore
# on the C++ names avoids collision with these R wrappers, which own the clean
# names.
create_age_comp <- function(n_years, n_ages) {
  xp      <- create_data_interface_("age_comp", n_years, n_ages)
  base_xp <- data_to_fims_xptr_(xp)              # base XPtr — for CreateTMBModel()
  .fims_registry$objects <- c(.fims_registry$objects,
                               list(base_xp))     # auto-register
  xp                                               # return XPtr to user
}

create_index <- function(n_years) {
  xp      <- create_data_interface_("index", n_years)
  base_xp <- data_to_fims_xptr_(xp)
  .fims_registry$objects <- c(.fims_registry$objects, list(base_xp))
  xp
}

# Repeat for every module type:
# create_length_comp(), create_catch(), create_population(), create_fleet(), ...

# R wrapper around C++ CreateTMBModel_ — uses the registry automatically
CreateTMBModel <- function() {
  CreateTMBModel_(.fims_registry$objects)
}

# clear() resets both the C++ TMB singletons and the R registry
clear <- function() {
  clear_()
  .fims_registry$objects <- list()
}
```

With this in place, the user workflow is identical to the current S4 workflow — create
objects, set values, call `CreateTMBModel()` — with no manual list management required.
The `clear()` function simplifies on the C++ side to resetting the TMB `Information`
singletons and `id_g` counters, with the R registry reset handling the bookkeeping that
`live_objects` and `fims_interface_objects` previously provided.

**Migrating `initialize_population()`.** Linking fleets to a population gets simpler, not
harder. The S4 method took one fleet at a time, so `initialize_population()` loops over
the fleet entries of `linked_ids`:

```{r eval=FALSE}
# R/initialize_modules.R — current
for (i in seq_along(linked_ids)) {
  module$AddFleet(linked_ids[[i]])
}
```

The XPtr setter replaces the whole set in one call, and takes the fleet objects rather
than their IDs, so the loop collapses:

```{r eval=FALSE}
set_population_fleets_(population, fleets)
```

`initialize_population()` already receives the full set of fleets, so nothing new has to
be threaded through — but it currently receives their *IDs*, in `linked_ids`, and the
setter needs the XPtrs. Whichever function assembles `linked_ids` should carry the
objects alongside the IDs, or return them, so the population can be given both its
`growth_id`/`maturity_id`/`recruitment_id` numbers and its fleet objects.

Replacing rather than appending is deliberate. It is the only form that can *reduce* the
fleet set, which is what lets a population built by copying another one be given
different fleets. It also keeps `fleet_ids` and `fleets_m` in step by construction: both
are cleared and rebuilt in the same loop, so they cannot drift apart the way two
separately-maintained lists could. `FisheryModelInterfaceBase::SetPopulations()` and its
`set_model_populations_()` wrapper follow the same pattern for the same reasons.

**Migrating `set_param_vector()`.** Parameter assignment gets shorter in the same way, and
in the function that already exists for it. `set_param_vector()` filters the parameters
tibble down to one label and pulls its `value` and `estimation_status` columns; only its
last few lines touch the module, and those are a per-element loop through the S4 field:

```{r eval=FALSE}
# R/initialize_modules.R — current
module[[field]]$resize(length(field_value))
for (i in seq_along(field_value)) {
  module[[field]][i][["value"]] <- field_value[i]
  module[[field]][i][["estimation_status"]]$set(field_estimation_status[i])
}
```

The whole block becomes one call. `set_variable_vector()` resizes to match the values it is
given, so the explicit resize goes too:

```{r eval=FALSE}
set_parameter_(module_base, field, field_value, field_estimation_status)
```

Everything above that in the function — the filtering, the grouping by label, the missing
value check — stays in R and stays unchanged.

**Bulk assignment belongs in R, not C++.** It is tempting to hand the parameters tibble
across the boundary and let C++ set every field of a module at once. That would be the wrong
place for it. The tibble's `module_name`, `module_type`, and `fleet` columns decide *which*
module receives *which* rows, so a C++ bulk setter would have to reimplement the filtering
and grouping that `dplyr` already does, and would pull module dispatch into C++ as well.
Crossing the boundary is not the cost to optimize: a population has roughly five settable
parameters, and even a large model is a few dozen calls made once during setup. R also
produces far better diagnostics, because it knows which row of the tibble a bad label came
from, and C++ sees only the name.

**What C++ does still owe: enumerating parameter names.** `get_parameter()` answers *"do you
have a parameter called this?"*, but nothing can ask a module *"what parameters do you
have?"* That gap has three consequences: R cannot validate a label against the module before
sending it, `describe_model()` (Section 4c) cannot list a module's parameters, and the module
manifest in Step E of `Parameter_Registry.Rmd` has nothing to build from.

Enumeration should not be added to the current hand-written `get_parameter()` overrides,
which are `if` chains and hold no list to return. It arrives for free with Step A of
`Parameter_Registry.Rmd`, where `declare_parameter()` in each constructor populates a
`parameters_m` vector that already holds every name.

Nor should the names become an enum in the sense that `SelectivityType` is one. Parameter
names are per module type and there are many of them across the families, so a shared enum
would be a large flat list in one file that every contributor adding a parameter has to
edit — the churn that keeping the family type tags local was meant to avoid. The R side also
drives assignment from the string `label` column of the parameters tibble, so an enum would
add a string-to-enum conversion in front of a lookup that already takes a string. The name
is data, and it should stay data.

**Migrating `distribution_formulas.R`.** This file needs more attention than the other R
migrations, because both of the things it does — reaching a module's data ID, and reaching a
parameter's ID — worked through S4 mechanics that no longer exist.

It reaches the observed data ID by building a method name as a string and looking it up:

```{r eval=FALSE}
# R/distribution_formulas.R — current
data_id_name <- glue::glue("GetObserved{snake_to_pascal(data_type)}DataID")
getter_function <- tryCatch(do.call("$", list(module, data_id_name)),
                            error = function(e) NULL)
if (is.null(getter_function)) {
  cli::cli_abort(c(x = "Could not find observed data name, {.code {data_id_name}} ..."))
}
new_module$set_observed_data(getter_function())
```

`get_fleet_observed_data_ids_()` returns all four IDs at once as a named integer vector,
using the same names as the `data_ids` entries in the JSON output, so the string-building,
the `tryCatch`, and the error branch all collapse to a lookup:

```{r eval=FALSE}
new_module <- set_distribution_observed_data_(
  new_module, get_fleet_observed_data_ids_(fleet)[[data_type]]
)
```

This assumes `data_type` already uses `"agecomp"`, `"lengthcomp"`, `"index"`, and
`"catch"`. If it does not, converting it once is a better place for the mapping than
`snake_to_pascal()` reconstructing a C++ method name — a name the R side should not have to
know.

The parameter links are the harder half. Both call sites read a `VariableVector` ID through
the S4 field:

```{r eval=FALSE}
new_module$set_distribution_links("data", module$field(link_name)$get_id())
new_module$set_distribution_links("random_effects", module$field(par)$get_id())
```

`$field()` and `VariableVector$get_id()` were both `Rcpp::class_<>` mechanics, so with those
registrations removed there is no path from R to a parameter's ID. That is why
`get_parameter_id_(module_base, name)` exists; it is the replacement for this exact call:

```{r eval=FALSE}
set_distribution_links_(new_module, "random_effects",
                        get_parameter_id_(module_base, par))
```

Two details worth carrying into the rewrite. `get_parameter_id_()` takes the **base** XPtr,
so the wrapper needs the module's base pointer here, not only its family-typed one. And
`par` and `link` come from the parameters tibble as strings, which is exactly what
`get_parameter_id_()` wants — the names line up without translation, unlike the data-ID
path above.

#### 4c. Model description from the registry

Because the R wrapper functions know the type and configuration of each object at the
moment of creation, a parallel metadata list can be stored alongside the base XPtrs at no
additional C++ cost. This metadata enables a `describe_model()` function that prints a
human-readable summary of the model the user has assembled — useful for verifying that all
components are present and correctly configured before calling `CreateTMBModel()`.

Extend the registry environment with a `metadata` list, and populate it inside each R
wrapper function:

```r
# R/fims_interface.R

.fims_registry <- new.env(parent = emptyenv())
.fims_registry$objects  <- list()
.fims_registry$metadata <- list()

create_age_comp <- function(n_years, n_ages) {
  xp      <- create_data_interface_("age_comp", n_years, n_ages)
  base_xp <- data_to_fims_xptr_(xp)
  .fims_registry$objects  <- c(.fims_registry$objects,  list(base_xp))
  .fims_registry$metadata <- c(.fims_registry$metadata,
                                list(list(type   = "AgeComp",
                                          # get_module_id_ takes the base XPtr
                                          id     = get_module_id_(base_xp),
                                          n_years = n_years,
                                          n_ages  = n_ages)))
  xp
}

create_population <- function() {
  xp      <- create_population_()
  base_xp <- population_to_fims_xptr_(xp)
  .fims_registry$objects  <- c(.fims_registry$objects,  list(base_xp))
  .fims_registry$metadata <- c(.fims_registry$metadata,
                                list(list(type = "Population",
                                          id   = get_id_(xp))))
  xp
}

# Repeat the same pattern for every module type.

# clear() resets both the C++ TMB singletons and the full R registry
clear <- function() {
  clear_()
  .fims_registry$objects  <- list()
  .fims_registry$metadata <- list()
}
```

`describe_model()` iterates the metadata list and prints a summary. Because this function
operates entirely on the R-side metadata, it requires no C++ changes:

```r
describe_model <- function() {
  if (length(.fims_registry$metadata) == 0) {
    message("No model components have been registered.")
    return(invisible(NULL))
  }
  cat("Model components registered:\n")
  cat(strrep("-", 40), "\n")
  for (m in .fims_registry$metadata) {
    # Build a string of any extra fields beyond type and id
    extras <- m[!names(m) %in% c("type", "id")]
    extra_str <- if (length(extras) > 0) {
      paste(names(extras), unlist(extras), sep = " = ", collapse = ", ")
    } else ""
    cat(sprintf("  %-20s id: %d  %s\n", m$type, m$id, extra_str))
  }
  cat(strrep("-", 40), "\n")
  cat(sprintf("Total: %d component(s)\n", length(.fims_registry$metadata)))
  invisible(.fims_registry$metadata)
}
```

Example output after setting up a simple model:

```
Model components registered:
----------------------------------------
  AgeComp              id: 1  n_years = 10, n_ages = 5
  Index                id: 2  n_years = 10
  BevertonHoltRecruit  id: 3
  LogisticSelectivity  id: 4
  Fleet                id: 5
  Population           id: 6
----------------------------------------
Total: 6 component(s)
```

`describe_model()` is intended to be called interactively before `CreateTMBModel()` as a
sanity check. Additional metadata fields — such as linked component IDs (e.g.,
`recruitment_id`, `selectivity_id`) or parameter dimensions — can be added to each wrapper
function without modifying any C++ code.

#### 4d. Forward compatibility with per-model contexts

FIMS currently holds exactly one model per R process, because `fims_info::Information<Type>`
and `fims_model::Model<Type>` are singletons reached through `GetInstance()`. Removing that
restriction is a separate piece of work — see `Singleton_Removal.Rmd` — but four decisions
made while writing Step 4 determine whether that work is a small change or a second pass
over the whole interface layer. All four are close to free now and expensive to retrofit.

**Pass a context to `add_to_fims_tmb()`.** Of the 51 `GetInstance()` call sites in the
codebase, 46 are in the interface layer that Step 4 rewrites. If `add_to_fims_tmb()` takes
its `Information` objects as a parameter rather than reaching for the singleton, those 46
sites are converted once, as part of work already being done, instead of twice.

No new type is required. `fims_model::Model<Type>` already owns `fims_information`, so the
context is a `Model<Type>` that is simply no longer a singleton. Because
`add_to_fims_tmb()` dispatches to both type instantiations, a small struct holding the pair
is enough:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_interface_base.hpp

struct FIMSModelContext {
  std::shared_ptr<fims_info::Information<TMB_FIMS_REAL_TYPE>> info0;
  std::shared_ptr<fims_info::Information<TMBAD_FIMS_TYPE>>    info;
};

class FIMSRcppInterfaceBase {
 public:
  virtual bool add_to_fims_tmb(FIMSModelContext &ctx) { ... }
};
```

In the interim `CreateTMBModel()` fills the struct from `GetInstance()`, so behavior is
unchanged. What changes is that every interface object now reads its `Information` from an
argument instead of a global, and switching to a per-model registry later becomes a change
to one function rather than to forty-six.

**Return a model handle from `CreateTMBModel()`.** It currently returns `bool` and always
returns `true`, and there is a single R call site. Returning an identifier — or an XPtr to
the context — costs nothing now and means the R wrappers carry a handle from the outset.
That handle is both the key a future registry looks up and the object R attaches to the
TMB objective function to tie their lifetimes together.

**Stop resetting `id_g` in `clear()`; let IDs be monotonic.** Rewinding the counters while
live objects still hold IDs issued before the reset is already a latent aliasing hazard, and
it is a hard blocker for per-context maps: `Information::variable_map` is keyed by ID, so
two contexts can only coexist if IDs are globally unique. Making "never reset" the uniform
rule also resolves the existing inconsistency where `Variable::id_g` and
`VariableVector::id_g` are defined as `0` but reset to `1`, which is never
reset at all.

**Let the XPtr own the context and anything global hold a weak reference.** A TMB objective
function holds no C++ reference back to the model it evaluates. If R releases the handle
while `obj` is still alive, the next `obj$fn()` call has nothing to dereference. The
registry should therefore hold `std::weak_ptr`, the XPtr should hold the strong reference,
and the R wrapper that calls `TMB::MakeADFun()` should attach the model handle to `obj` as
an attribute so R's garbage collector keeps the two alive together. Establishing this
direction in the Section 4b registry costs nothing; reversing ownership semantics after the
API ships does not.

---

## 3. Usage Example

### 3.1 R Workflow with Automatic Registry

With the R-side registry in place (Step 4b), the user workflow closely mirrors the current
S4 workflow. `CreateTMBModel()` is called with no arguments — the registry built up during
object creation is used automatically. Note that `$method()` syntax does not work directly
on raw XPtr objects; all field access goes through standalone setter functions. See
Section 3.2 for R6 wrappers that restore familiar `$` ergonomics.

```{r eval=FALSE}
library(FIMS)

# ==================== SETUP ====================

# 1. Create model components.
#    Each R wrapper calls the C++ creator AND registers the object automatically.
#    The concrete XPtr is returned for use in setter calls below.
age_comp <- create_age_comp(n_years = 10, n_ages = 5)
index    <- create_index(n_years = 10)
pop      <- create_population()
growth   <- create_empirical_growth()

# 2. Set values via the shared setter functions.
#    set_vector_() and set_parameter_() cover every module in FIMS: the field is
#    named rather than typed, so no module type appears in the call. They take
#    the base XPtr, which the R wrappers hold alongside the concrete one.
set_vector(age_comp, "values", my_age_comp_data)
set_vector(age_comp, "uncertainty", my_age_comp_cv)
set_vector(index, "values", my_index_data)
set_population_constants(pop, n_years = 10L, n_ages = 5L, n_lengths = 0L)
set_population_process_ids(pop, growth_id = get_module_id(growth))

# 3. Build the TMB model — registry is passed automatically, no list needed
CreateTMBModel()

# fit model...


# ==================== ITERATIVE WORKFLOW ====================

# 4. For a second scenario (e.g., a sensitivity run with different data),
#    update values directly on the existing concrete XPtrs.
#    The registry already holds base XPtrs pointing to the same underlying C++ objects,
#    so any change made here is automatically visible when CreateTMBModel() runs again —
#    no re-registration or clear() needed between scenarios.
set_vector(age_comp, "values", new_age_comp_data)   # update age composition data
set_vector(index, "values", new_index_data)         # update index data
set_parameter(pop, "log_M",                         # update starting values
              values = new_log_M,
              estimation_status = "assumed_known")
set_parameter(pop, "log_init_naa",
              values = new_log_init_naa,
              estimation_status = "fixed_effects")

# 5. Rebuild — CreateTMBModel() internally calls fims::Information::Clear() on both
#    TMB singletons (double and TMBAD types) before re-populating from the registry,
#    so the previous model run's state is fully reset. The R registry itself is NOT
#    cleared, which is what allows the same objects to be reused across scenarios.
CreateTMBModel()

# fit model...


# ==================== CLEANUP ====================

# 6. clear() resets the TMB singletons AND the R registry.
#    Setting R variables to NULL releases the concrete XPtrs;
#    shared_ptr ref counts drop to 0 on next GC and memory is freed.
clear()
age_comp_xp <- NULL
index_xp    <- NULL
pop_xp      <- NULL
gc()
```

### 3.2 R6 Wrapper Workflow - Future Development, will not be implemented in this issue

R6 classes restore `$` ergonomics by wrapping both the concrete XPtr (for typed field
access) and the base XPtr (for `CreateTMBModel()` dispatch) inside a single user-facing
object:

```{r eval=FALSE}
AgeComp <- R6::R6Class("AgeComp",
  private = list(
    .xp      = NULL,
    .base_xp = NULL
  ),
  public = list(
    initialize = function(n_years, n_ages) {
      private$.xp      <- create_data_interface_("age_comp", n_years, n_ages)
      private$.base_xp <- data_to_fims_xptr_(private$.xp)
    },
    set_values = function(values) {
      set_vector_(private$.base_xp, "values", values)
    },
    fims_xptr = function() private$.base_xp
  )
)

Population <- R6::R6Class("Population",
  private = list(
    .xp      = NULL,
    .base_xp = NULL
  ),
  public = list(
    initialize = function() {
      private$.xp      <- create_population_()
      private$.base_xp <- population_to_fims_xptr_(private$.xp)
    },
    set_n_years = function(n) set_population_n_years_(private$.xp, n),
    set_n_ages  = function(n) set_population_n_ages_(private$.xp, n),
    fims_xptr   = function() private$.base_xp
  )
)

# User workflow
age_comp <- AgeComp$new(n_years = 10, n_ages = 5)
age_comp$set_values(my_age_comp_data)

pop <- Population$new()
pop$set_n_years(10L)
pop$set_n_ages(5L)

# Under R6, CreateTMBModel() can receive explicit base XPtrs from each object,
# or the R-side registry (Section 4b/4c) can be used here too — both patterns work.
CreateTMBModel(list(age_comp$fims_xptr(), pop$fims_xptr()))
```

---

## 4. Memory Impact

The following table summarizes heap allocations eliminated per `PopulationInterface`
instance by the full refactor. The XPtr approach only realizes these savings when all four
steps are applied together; a partial migration that retains SharedPrimitives or constructor
copies will not produce a meaningful reduction.

| Source | Before | After |
|---|---|---|
| `SharedInt` configuration fields (8) | 8 heap allocs | 0 |
| `SharedBoolean` fields (2 in base) | 2 heap allocs | 0 |
| `SharedString` field (1) | 1 heap alloc | 0 |
| Constructor registry copy | 1 `make_shared<PopulationInterface>` | 0 |
| `Variable::estimation_type_m` per element | 1 heap alloc per `Variable` | 0 (`EstimationStatus` enum) |
| `live_objects` map per class | 1 `std::map` + N entries | removed |
| `fims_interface_objects` global vector | 1 entry per object | removed |
| `RealVector` shared storage per instance | 1 `make_shared<std::vector>` | removed (class deleted) |

The `clear()` function simplifies substantially: instead of manually resetting every
class's `live_objects` map and `id_g` counter, it only needs to reset the TMB `Information`
singletons and the `id_g` counters — no per-class map iteration required.


## 5. Notes

**Copy operations are deleted, not removed.** During Step 3, the copy constructor and
copy assignment operator of every interface class were set to `= delete`. Copies existed
to populate the `live_objects` and `fims_interface_objects` registries. With those
registries gone, a copy would only create a second object carrying a duplicate `id`,
which would then diverge from the object the `Rcpp::XPtr` points at. Merely removing the
copy constructor is not enough — C++ synthesizes an implicit one and the class stays
copyable. `= delete` makes a copy a compile error instead of a silent bug.

**`RealVector` was removed.** It wrapped a `std::vector<double>` in a `shared_ptr` so that
the registry copies of an interface would share one buffer with the original. With the
copies gone the sharing had no purpose, and unlike `VariableVector` its `id_m` was never
used as a `variable_map` key. Data vectors are now plain `fims::Vector<double>` with value
semantics. `Rcpp::NumericVector` was considered as the replacement and rejected: holding a
`SEXP` as a long-lived member pins R memory through `R_PreserveObject` and confines access
to the main R thread.

**Still to do.**

- `deep_copy()` for interface modules. Now that copying is a compile error, a user who
  wants a second scenario built from an existing module needs an explicit function. The
  design is settled but unimplemented: copy the input field values, share rather than
  duplicate the structural links, assign fresh IDs, and copy input fields only — never
  derived quantities, which belong to a model run rather than to the module.
- The R side is unmigrated. `initialize_modules.R`, `distribution_formulas.R`, and
  `fimsfit.R` still build modules with `methods::new()`. The migration notes in Section 4
  cover `initialize_population()`, `set_param_vector()`, and `distribution_formulas.R`.
- The vignettes still document `live_objects` and `methods::new()` as the current API and
  need rewriting once the R side lands.
