---
title: "Design Document: FIMS Rcpp Interface Refactor to Rcpp::XPtr"
author: "FIMS Development Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Motivation

The current FIMS architecture relies on S4 Rcpp::Module classes to expose C++ interface
objects (such as `AgeCompDataInterface` or `PopulationInterface`) directly to R. This
design introduces two structural problems that compound each other.

### 1.1 The Copy Problem

When `methods::new(Population)` is called in R, Rcpp owns the C++ object's memory through
its own internal mechanism. A `std::shared_ptr` cannot safely be created pointing at a
Rcpp-managed object, because Rcpp controls its lifetime. However, `CreateTMBModel()` needs
to iterate over all interface objects to call `add_to_fims_tmb()` via a registry it owns.
The solution in the current design is to make a heap copy of every interface object in the
constructor:

```cpp
// PopulationInterface constructor — current approach
PopulationInterface() : PopulationInterfaceBase() {
    // ...
    std::shared_ptr<PopulationInterface> population =
        std::make_shared<PopulationInterface>(*this);  // heap copy
    FIMSRcppInterfaceBase::fims_interface_objects.push_back(population);
    PopulationInterfaceBase::live_objects[this->id] = population;
}
```

Because these global static collections (variables that are shared across the program and persist for the lifetime of the program) maintain strong `std::shared_ptr` references to
every object created, their reference counts never drop to zero. Even when an R variable is
overwritten (e.g., `my_fleet <- NULL`), the underlying C++ memory is never deallocated
until an explicit manual call to `clear()`. This prevents clean iterative modeling workflows
(bootstrapping, sensitivity runs, profile likelihood) within a single R session.

### 1.2 The SharedPrimitive Overhead Problem

When users initialize a module (e.g. pop <- methods::new(Population)), a copy gets stored in the registry. When the user later sets
`pop$n_years <- 30` on the R-side S4 object, this modifies the original Rcpp-managed object
but not the registry copy. `CreateTMBModel()` would see stale initial values.

The `SharedInt`, `SharedBoolean`, `SharedReal`, and `SharedString` wrapper classes were
introduced to solve this synchronization problem: by wrapping scalar fields in `shared_ptr`,
both the original and the registry copy share the same underlying value. Modifying the
original is automatically seen by the copy.

However, this comes at a significant memory cost. Each `SharedPrimitive` field requires one
heap allocation for a single scalar. On a 64-bit system, `make_shared<int>(0)` allocates
approximately 28 bytes (16-byte control block + 4-byte int + alignment padding) for 4 bytes
of actual data — a 7:1 overhead ratio. `PopulationInterface` alone has 8 `SharedInt`, 2
`SharedBoolean`, and 1 `SharedString` configuration fields: roughly 11 heap allocations
before any model data is set. Additionally, every `Variable` element in every
`VariableVector` carries a `SharedString estimation_type_m`, adding one heap allocation per
parameter element across the entire model.

The SharedPrimitive types and the constructor copies are symptoms of the same root cause:
the S4 module system forces a two-object design (Rcpp-owned original + heap registry copy)
that requires shared ownership of individual fields to stay synchronized.

### 1.3 The Rcpp::XPtr Solution

By transitioning the R-to-C++ boundary to `Rcpp::XPtr<std::shared_ptr<T>>`, both problems
are eliminated together.

**One object, no copies.** The creator function allocates the C++ object on the heap
directly and hands ownership to a `shared_ptr`. The XPtr wraps that `shared_ptr`. There is
no Rcpp-managed original and no registry copy — there is one object. User modifications via
standalone setter functions go directly to that object.

**SharedPrimitives become unnecessary.** Without a copy to keep synchronized with the
original, scalar fields can revert to plain C++ types (`int`, `bool`, `double`,
`std::string`). The `Variable::estimation_type_m` field can become an enum.

**Automated lifecycle management.** When an R variable holding an XPtr is overwritten or
goes out of scope, R's garbage collector deletes the external pointer wrapper, decrementing
the `shared_ptr` reference count. Memory is freed without requiring an explicit `clear()`
call.

---

## 2. Refactoring Steps

### Step 1: Replace `SharedPrimitive` scalar fields with plain C++ types

In every interface class, replace `SharedInt`, `SharedBoolean`, `SharedReal`, and
`SharedString` fields that hold scalar configuration values with their plain C++ equivalents.
This is only safe under XPtr because there is no registry copy that needs to remain
synchronized with user modifications.

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_population.hpp

// BEFORE — SharedPrimitive fields (one heap allocation each)
SharedInt n_ages = 0;
SharedInt n_fleets;
SharedInt n_years;
SharedInt n_lengths;
SharedInt maturity_id;
SharedInt growth_id;
SharedInt recruitment_id;
SharedInt recruitment_err_id;
SharedString name = fims::to_string("NA");

// AFTER — plain C++ types (no heap allocation per scalar)
int n_ages = 0;
int n_fleets = 0;
int n_years = 0;
int n_lengths = 0;
int maturity_id = -1;
int growth_id = -1;
int recruitment_id = -1;
int recruitment_err_id = -1;
std::string name = "NA";
```

`VariableVector` retains its `shared_ptr<vector<Variable>>` internal storage because R-side
access to individual elements relies on shared ownership within the vector. Only standalone
scalar configuration fields change.

`SharedBoolean` fields in base classes (`initialize_catch_at_age`,
`initialize_surplus_production`) follow the same pattern — replace with plain `bool`.

### Step 2: Replace `Variable::estimation_type_m` with an enum

Every `Variable` element in every `VariableVector` currently carries a
`SharedString estimation_type_m`, which heap-allocates one `std::string` per element. The
field always holds exactly one of three values. Replace it with an enum stored as a single
byte:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_interface_base.hpp

// BEFORE — one heap allocation per Variable element
class Variable {
 public:
  SharedString estimation_type_m = SharedString("constant");
  // ...
};

// AFTER — zero heap allocations, 1 byte per Variable element
enum class EstimationType : uint8_t {
  constant      = 0,
  fixed_effects  = 1,
  random_effects = 2
};

class Variable {
 public:
  EstimationType estimation_type_m = EstimationType::constant;
  // ...
};
```

Update all callers that previously compared strings (e.g.,
`estimation_type_m.get() == "fixed_effects"`) to use the enum values:

```cpp
// BEFORE
if (this->log_M[i].estimation_type_m.get() == "fixed_effects") { ... }

// AFTER
if (this->log_M[i].estimation_type_m == EstimationType::fixed_effects) { ... }
```

Update `VariableVector::set_estimation_types()` to accept an
`Rcpp::CharacterVector` and convert string input to enum internally, preserving the
existing R-side API.

### Step 3: Remove constructor copies and implement standalone lifecycle functions

Strip `live_objects` map insertions and `fims_interface_objects` push_back calls from all
interface class constructors. The constructor should perform only pure initialization:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_data.hpp

// BEFORE — constructor registers in two global registries
AgeCompDataInterface(int  = 0, int n_ages = 0) : DataInterfaceBase() {
    this->n_ages = n_ages;
    this-> = ;
    this->age_comp_data.resize(n_ages * );
    this->uncertainty.resize(n_ages * );
    DataInterfaceBase::live_objects[this->id] =
        std::make_shared<AgeCompDataInterface>(*this);
    FIMSRcppInterfaceBase::fims_interface_objects.push_back(
        DataInterfaceBase::live_objects[this->id]);
}

// AFTER — pure initialization only
AgeCompDataInterface(int  = 0, int n_ages = 0) : DataInterfaceBase() {
    this->n_ages = n_ages;
    this-> = ;
    this->age_comp_data.resize(n_ages * );
    this->uncertainty.resize(n_ages * );
}
```

Implement standalone creator functions in the corresponding `.cpp` source file. Wrap the
`shared_ptr` (not a raw pointer) inside the XPtr so that the reference count is safely
decremented by R's garbage collector when the XPtr is collected:

```cpp
// src/rcpp_data.cpp

using SharedAgeComp = std::shared_ptr<AgeCompDataInterface>;

Rcpp::XPtr<SharedAgeComp> create_age_comp(int , int n_ages) {
    auto obj = std::make_shared<AgeCompDataInterface>(, n_ages);
    // Heap-allocate the shared_ptr itself and hand it to the XPtr.
    // 'true' binds the XPtr to R's garbage collector: when the R variable
    // is collected, 'delete shared_ptr*' is called, decrementing the ref count.
    return Rcpp::XPtr<SharedAgeComp>(new SharedAgeComp(obj), true);
}
```

Implement standalone setter functions for each field that the user needs to set from R.
Because there is one object and no copy to synchronize, plain C++ types suffice:

```cpp
void set_age_comp_values(Rcpp::XPtr<SharedAgeComp> xp,
                         Rcpp::NumericVector values) {
    auto& data = (*xp)->age_comp_data;
    data.resize(values.size());
    for (int i = 0; i < values.size(); ++i) {
        data[i] = values[i];
    }
}
```

```cpp
void set_population_constants(Rcpp::XPtr<SharedPopulation> xp,
                              int n_ages,
                              int n_years,
                              int n_lengths) {
    (*xp)->n_ages   = n_ages;
    (*xp)->n_years  = n_years;
    (*xp)->n_lengths = n_lengths;
}
```
```cpp
void set_population_parameters(Rcpp::XPtr<SharedPopulation> xp,
                               Rcpp::NumericVector log_M_values,
                               Rcpp::NumericVector log_init_naa_values) {
    // xp returns the Xptr to the shared pointer
    // *xp returns the shared pointer to the PopulationInterface
    // **xp returns the PopulationInterface object
    PopulationInterface& pop = **xp;

    pop.log_M.resize(log_M_values.size());
    for (int i = 0; i < log_M_values.size(); i++) {

        Variable& v         = pop.log_M.storage_m->at(i);
        v.initial_value_m   = log_M_values[i];
        v.estimation_type_m = EstimationType::constant;
    }

    pop.log_init_naa.resize(log_init_naa_values.size());
    for (int i = 0; i < log_init_naa_values.size(); i++) {
        Variable& v         = pop.log_init_naa.storage_m->at(i);
        v.initial_value_m   = log_init_naa_values[i];
        v.estimation_type_m = EstimationType::fixed_effects;
    }
}
```

Register these functions inside the existing module registration blocks. Do **not** use
`// [[Rcpp::exports]]`, as this causes essential TMB callables to be overwritten by
`Rcpp::compileAttributes()`. Add a trailing underscore to every registered name to prevent
name collisions with the R wrapper functions defined in Section 4b. Remove the
`Rcpp::class_<T>` registration for any interface class that has been fully migrated to XPtr:

```cpp
// src/rcpp_data.cpp

void register_data(Rcpp::Module& m) {
    // Rcpp::class_<AgeCompDataInterface> removed — users no longer call methods::new()
    Rcpp::function("create_age_comp_",          &create_age_comp);
    Rcpp::function("set_age_comp_values_",      &set_age_comp_values);
    Rcpp::function("set_population_constants_", &set_population_constants);
    Rcpp::function("set_population_parameters_",&set_population_parameters);
    // ... repeat for LengthComp, Index, Landings ...
}
```

### Step 4: Implement dispatch through a shared base class for `CreateTMBModel()`

`CreateTMBModel()` currently iterates `FIMSRcppInterfaceBase::fims_interface_objects` to
call `add_to_fims_tmb()` on every registered object. Under XPtr, the C++ global registry
is removed. This step replaces it with two pieces: a C++ mechanism for calling
`add_to_fims_tmb()` through a common base class pointer, and an R-side registry that
automatically tracks objects as they are created so the user does not need to build the list
by hand.

#### 4a. Why two XPtr types are needed

Each concrete creator function (e.g., `create_age_comp()`, `create_population()`) returns
an XPtr typed to its specific class — `XPtr<SharedAgeComp>`, `XPtr<SharedPopulation>`, etc.
These types are all different, so they cannot be stored together in a single list and passed
to `CreateTMBModel()`.

The solution is a second, less specific XPtr typed to the common base class
`FIMSRcppInterfaceBase`, which all interface classes inherit from. Because
`add_to_fims_tmb()` is declared as a `virtual` method on that base class, calling it through
a base class pointer is sufficient — C++ automatically invokes the correct derived-class
implementation at runtime. This is what "dispatch through a shared base class" means:
`CreateTMBModel()` does not need to know whether it is looking at an `AgeCompDataInterface`
or a `PopulationInterface`; it just calls `add_to_fims_tmb()` and C++ figures out the rest.

The two XPtr types point to the same underlying C++ object in memory. The concrete XPtr is
what the user works with (setting values, reading results). The base XPtr is only needed
by `CreateTMBModel()` to make that call. Both XPtrs share ownership of the object via
`shared_ptr`, so the object stays alive as long as either pointer exists.

**Define the base alias** in `rcpp_interface_base.hpp`:

```cpp
// inst/include/interface/rcpp/rcpp_objects/rcpp_interface_base.hpp
using SharedBase = std::shared_ptr<FIMSRcppInterfaceBase>;
```

**Provide a conversion function for each concrete type.** This function takes the
concrete XPtr the user already has and returns a second XPtr pointing to the same object
but typed as the base class. C++ allows this conversion implicitly because
`AgeCompDataInterface` inherits from `FIMSRcppInterfaceBase` — no special cast syntax
is needed:

```cpp
// src/rcpp_data.cpp

// create_age_comp_      — allocates a NEW AgeCompDataInterface object
// age_comp_to_fims_xptr_ — does NOT create a new object; returns a second pointer
//                          to the SAME object typed as the base class
Rcpp::XPtr<SharedBase> age_comp_to_fims_xptr(Rcpp::XPtr<SharedAgeComp> xp) {
    SharedBase base = *xp;  // implicit conversion: shared_ptr<Derived> -> shared_ptr<Base>
    return Rcpp::XPtr<SharedBase>(new SharedBase(base), true);
}
```

Register both functions inside the existing module registration block, again with trailing
underscores so the clean names remain available for the R wrappers in Section 4b:

```cpp
void register_data(Rcpp::Module& m) {
    Rcpp::function("create_age_comp_",         &create_age_comp);
    Rcpp::function("set_age_comp_values_",     &set_age_comp_values);
    Rcpp::function("age_comp_to_fims_xptr_",   &age_comp_to_fims_xptr);
    // ... repeat for LengthComp, Index, Landings ...
}
```

Register `CreateTMBModel` and `clear` the same way in their respective registration blocks:

```cpp
Rcpp::function("CreateTMBModel_", &CreateTMBModel);
Rcpp::function("clear_",          &clear_impl);
```
```

**Update `CreateTMBModel()` in C++ to accept a list of base XPtrs** and call
`add_to_fims_tmb()` on each:

```cpp
// inst/include/interface/rcpp/rcpp_interface.hpp

bool CreateTMBModel(Rcpp::List xptr_list) {
    init_logging();

    std::shared_ptr<fims_info::Information<TMB_FIMS_REAL_TYPE>> info0 =
        fims_info::Information<TMB_FIMS_REAL_TYPE>::GetInstance();
    info0->Clear();

    std::shared_ptr<fims_info::Information<TMBAD_FIMS_TYPE>> info =
        fims_info::Information<TMBAD_FIMS_TYPE>::GetInstance();
    info->Clear();

    for (int i = 0; i < xptr_list.size(); i++) {
        Rcpp::XPtr<SharedBase> xp =
            Rcpp::as<Rcpp::XPtr<SharedBase>>(xptr_list[i]);
        (*xp)->add_to_fims_tmb();  // C++ calls the correct derived-class method
    }

    info0->CreateModel();
    info0->CheckModel();
    info->CreateModel();

    std::shared_ptr<fims_model::Model<TMB_FIMS_REAL_TYPE>> m0 =
        fims_model::Model<TMB_FIMS_REAL_TYPE>::GetInstance();

    return true;
}
```

#### 4b. R-side automatic registry (intermediate step before R6)

Rather than requiring the user to manually build the list passed to `CreateTMBModel()` —
which risks silently omitting a component — an R-side package-level registry can be
maintained automatically. The registry is a simple list stored inside a private package
environment (not visible to the user). Thin R wrapper functions around each C++ creator
call both `create_<type>()` and `<type>_to_fims_xptr()` and add the result to the registry.
The user-facing `CreateTMBModel()` and `clear()` functions are also R wrappers that operate
on this registry.

This approach is more reliable than scanning the R environment for XPtr objects, because
registration happens at the moment of creation — it is not affected by variable names,
scoping, or objects stored inside lists.

```r
# R/fims_interface.R

# Private package-level registry — not exported, not user-visible
.fims_registry <- new.env(parent = emptyenv())
.fims_registry$objects <- list()

# R wrapper around C++ create_age_comp_:
# creates the concrete XPtr, converts to base XPtr, registers it, returns concrete XPtr.
# The trailing underscore on the C++ names (create_age_comp_, age_comp_to_fims_xptr_)
# avoids collision with these R wrapper functions, which own the clean names.
create_age_comp <- function(n_years, n_ages) {
  xp      <- create_age_comp_(n_years, n_ages)   # concrete XPtr — user works with this
  base_xp <- age_comp_to_fims_xptr_(xp)          # base XPtr — for CreateTMBModel()
  .fims_registry$objects <- c(.fims_registry$objects,
                               list(base_xp))     # auto-register
  xp                                               # return concrete XPtr to user
}

# Repeat for every module type:
# create_population(), create_fleet(), create_index(), etc.

# R wrapper around C++ CreateTMBModel_ — uses the registry automatically
CreateTMBModel <- function() {
  CreateTMBModel_(.fims_registry$objects)
}

# clear() resets both the C++ TMB singletons and the R registry
clear <- function() {
  clear_()
  .fims_registry$objects <- list()
}
```

With this in place, the user workflow is identical to the current S4 workflow — create
objects, set values, call `CreateTMBModel()` — with no manual list management required.
The `clear()` function simplifies on the C++ side to resetting the TMB `Information`
singletons and `id_g` counters, with the R registry reset handling the bookkeeping that
`live_objects` and `fims_interface_objects` previously provided.

#### 4c. Model description from the registry

Because the R wrapper functions know the type and configuration of each object at the
moment of creation, a parallel metadata list can be stored alongside the base XPtrs at no
additional C++ cost. This metadata enables a `describe_model()` function that prints a
human-readable summary of the model the user has assembled — useful for verifying that all
components are present and correctly configured before calling `CreateTMBModel()`.

Extend the registry environment with a `metadata` list, and populate it inside each R
wrapper function:

```r
# R/fims_interface.R

.fims_registry <- new.env(parent = emptyenv())
.fims_registry$objects  <- list()
.fims_registry$metadata <- list()

create_age_comp <- function(n_years, n_ages) {
  xp      <- create_age_comp_(n_years, n_ages)
  base_xp <- age_comp_to_fims_xptr_(xp)
  .fims_registry$objects  <- c(.fims_registry$objects,  list(base_xp))
  .fims_registry$metadata <- c(.fims_registry$metadata,
                                list(list(type   = "AgeComp",
                                          id     = get_id_(xp),
                                          n_years = n_years,
                                          n_ages  = n_ages)))
  xp
}

create_population <- function() {
  xp      <- create_population_()
  base_xp <- population_to_fims_xptr_(xp)
  .fims_registry$objects  <- c(.fims_registry$objects,  list(base_xp))
  .fims_registry$metadata <- c(.fims_registry$metadata,
                                list(list(type = "Population",
                                          id   = get_id_(xp))))
  xp
}

# Repeat the same pattern for every module type.

# clear() resets both the C++ TMB singletons and the full R registry
clear <- function() {
  clear_()
  .fims_registry$objects  <- list()
  .fims_registry$metadata <- list()
}
```

`describe_model()` iterates the metadata list and prints a summary. Because this function
operates entirely on the R-side metadata, it requires no C++ changes:

```r
describe_model <- function() {
  if (length(.fims_registry$metadata) == 0) {
    message("No model components have been registered.")
    return(invisible(NULL))
  }
  cat("Model components registered:\n")
  cat(strrep("-", 40), "\n")
  for (m in .fims_registry$metadata) {
    # Build a string of any extra fields beyond type and id
    extras <- m[!names(m) %in% c("type", "id")]
    extra_str <- if (length(extras) > 0) {
      paste(names(extras), unlist(extras), sep = " = ", collapse = ", ")
    } else ""
    cat(sprintf("  %-20s id: %d  %s\n", m$type, m$id, extra_str))
  }
  cat(strrep("-", 40), "\n")
  cat(sprintf("Total: %d component(s)\n", length(.fims_registry$metadata)))
  invisible(.fims_registry$metadata)
}
```

Example output after setting up a simple model:

```
Model components registered:
----------------------------------------
  AgeComp              id: 1  n_years = 10, n_ages = 5
  Index                id: 2  n_years = 10
  BevertonHoltRecruit  id: 3
  LogisticSelectivity  id: 4
  Fleet                id: 5
  Population           id: 6
----------------------------------------
Total: 6 component(s)
```

`describe_model()` is intended to be called interactively before `CreateTMBModel()` as a
sanity check. Additional metadata fields — such as linked component IDs (e.g.,
`recruitment_id`, `selectivity_id`) or parameter dimensions — can be added to each wrapper
function without modifying any C++ code.

---

## 3. Usage Example

### 3.1 R Workflow with Automatic Registry

With the R-side registry in place (Step 4b), the user workflow closely mirrors the current
S4 workflow. `CreateTMBModel()` is called with no arguments — the registry built up during
object creation is used automatically. Note that `$method()` syntax does not work directly
on raw XPtr objects; all field access goes through standalone setter functions. See
Section 3.2 for R6 wrappers that restore familiar `$` ergonomics.

```{r eval=FALSE}
library(FIMS)

# ==================== SETUP ====================

# 1. Create model components.
#    Each R wrapper calls the C++ creator AND registers the object automatically.
#    The concrete XPtr is returned for use in setter calls below.
age_comp <- create_age_comp(n_years = 10, n_ages = 5)
index    <- create_index(n_years = 10)
pop      <- create_population()
growth   <- create_empirical_growth()

# 2. Set values via standalone setter functions
set_age_comp_values(age_comp, my_age_comp_data)
set_index_values(index, my_index_data)
set_population_constants(pop, n_years = 10L, n_ages = 5L, n_length = 0L)
set_population_growth_id(pop, get_id(growth))

# 3. Build the TMB model — registry is passed automatically, no list needed
CreateTMBModel()

# fit model...


# ==================== ITERATIVE WORKFLOW ====================

# 4. For a second scenario (e.g., a sensitivity run with different data),
#    update values directly on the existing concrete XPtrs.
#    The registry already holds base XPtrs pointing to the same underlying C++ objects,
#    so any change made here is automatically visible when CreateTMBModel() runs again —
#    no re-registration or clear() needed between scenarios.
set_age_comp_values(age_comp, new_age_comp_data)   # update age composition data
set_index_values(index, new_index_data)             # update index data
set_population_parameters(pop,                      # update parameter starting values
                          log_M_values       = new_log_M,
                          log_init_naa_values = new_log_init_naa)

# 5. Rebuild — CreateTMBModel() internally calls fims::Information::Clear() on both
#    TMB singletons (double and TMBAD types) before re-populating from the registry,
#    so the previous model run's state is fully reset. The R registry itself is NOT
#    cleared, which is what allows the same objects to be reused across scenarios.
CreateTMBModel()

# fit model...


# ==================== CLEANUP ====================

# 6. clear() resets the TMB singletons AND the R registry.
#    Setting R variables to NULL releases the concrete XPtrs;
#    shared_ptr ref counts drop to 0 on next GC and memory is freed.
clear()
age_comp_xp <- NULL
index_xp    <- NULL
pop_xp      <- NULL
gc()
```

### 3.2 R6 Wrapper Workflow - Future Development, will not be implemented in this issue

R6 classes restore `$` ergonomics by wrapping both the concrete XPtr (for typed field
access) and the base XPtr (for `CreateTMBModel()` dispatch) inside a single user-facing
object:

```{r eval=FALSE}
AgeComp <- R6::R6Class("AgeComp",
  private = list(
    .xp      = NULL,
    .base_xp = NULL
  ),
  public = list(
    initialize = function(n_years, n_ages) {
      private$.xp      <- create_age_comp_(n_years, n_ages)
      private$.base_xp <- age_comp_to_fims_xptr_(private$.xp)
    },
    set_values = function(values) {
      set_age_comp_values_(private$.xp, values)
    },
    fims_xptr = function() private$.base_xp
  )
)

Population <- R6::R6Class("Population",
  private = list(
    .xp      = NULL,
    .base_xp = NULL
  ),
  public = list(
    initialize = function() {
      private$.xp      <- create_population_()
      private$.base_xp <- population_to_fims_xptr_(private$.xp)
    },
    set_n_years = function(n) set_population_n_years_(private$.xp, n),
    set_n_ages  = function(n) set_population_n_ages_(private$.xp, n),
    fims_xptr   = function() private$.base_xp
  )
)

# User workflow
age_comp <- AgeComp$new(n_years = 10, n_ages = 5)
age_comp$set_values(my_age_comp_data)

pop <- Population$new()
pop$set_n_years(10L)
pop$set_n_ages(5L)

# Under R6, CreateTMBModel() can receive explicit base XPtrs from each object,
# or the R-side registry (Section 4b/4c) can be used here too — both patterns work.
CreateTMBModel(list(age_comp$fims_xptr(), pop$fims_xptr()))
```

---

## 4. Memory Impact

The following table summarizes heap allocations eliminated per `PopulationInterface`
instance by the full refactor. The XPtr approach only realizes these savings when all four
steps are applied together; a partial migration that retains SharedPrimitives or constructor
copies will not produce a meaningful reduction.

| Source | Before | After |
|---|---|---|
| `SharedInt` configuration fields (8) | 8 heap allocs | 0 |
| `SharedBoolean` fields (2 in base) | 2 heap allocs | 0 |
| `SharedString` field (1) | 1 heap alloc | 0 |
| Constructor registry copy | 1 `make_shared<PopulationInterface>` | 0 |
| `Variable::estimation_type_m` per element | 1 heap alloc per `Variable` | 0 (enum) |
| `live_objects` map per class | 1 `std::map` + N entries | removed |
| `fims_interface_objects` global vector | 1 entry per object | removed |

The `clear()` function simplifies substantially: instead of manually resetting every
class's `live_objects` map and `id_g` counter, it only needs to reset the TMB `Information`
singletons and the `id_g` counters — no per-class map iteration required.
